---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = TRUE
)
package_version <- read.dcf("DESCRIPTION", fields = "Version")[[1L]]
```

# fmrireg

<!-- badges: start -->
[![R-CMD-check](https://github.com/bbuchsbaum/fmrireg/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbuchsbaum/fmrireg/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/bbuchsbaum/fmrireg/branch/main/graph/badge.svg)](https://app.codecov.io/gh/bbuchsbaum/fmrireg?branch=main)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/license/mit)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

[Documentation](https://bbuchsbaum.github.io/fmrireg/) ·
[Getting started](https://bbuchsbaum.github.io/fmrireg/articles/fmrireg.html) ·
[API reference](https://bbuchsbaum.github.io/fmrireg/reference/) ·
[Changelog](NEWS.md)

**fmrireg** is an R package for specifying and fitting voxelwise fMRI
regression models from event tables and preprocessed time series. It turns
formulas such as `onset ~ hrf(condition)` into HRF-convolved designs, estimates
condition effects across voxels, and carries named contrasts into group
analysis.

> **Status:** Version `r package_version` is experimental and requires R 4.1 or
> later. APIs may change while the package is under active development.

## Installation

Install the development version from GitHub:

```r
install.packages("remotes")
remotes::install_github("bbuchsbaum/fmrireg")
```

## Quick start

This synthetic example makes one task-responsive time series and fits it
without external imaging files:

```{r quickstart}
library(fmrireg)

TR <- 2
n_timepoints <- 120
acquisition_time <- seq(TR / 2, by = TR, length.out = n_timepoints)
set.seed(42)
events <- data.frame(
  onset = cumsum(sample(c(6, 8, 10), 24, replace = TRUE)),
  condition = factor(sample(rep(c("faces", "scenes"), each = 12))),
  duration = 0,
  run = 1L
)

task_regressor <- function(condition) {
  fmrihrf::regressor(
    events$onset[events$condition == condition]
  ) |>
    fmrihrf::evaluate(acquisition_time) |>
    as.numeric()
}

faces <- task_regressor("faces")
scenes <- task_regressor("scenes")
bold <- cbind(
  face_roi = 100 + 1.5 * faces + 0.1 * scenes + rnorm(n_timepoints, sd = 0.04)
)

dataset <- matrix_dataset(
  bold, TR = TR, run_length = n_timepoints,
  event_table = events
)
fit <- fmri_lm(
  onset ~ hrf(condition), block = ~ run, dataset = dataset
)

results <- subset(
  tidy(fit), term %in% levels(events$condition),
  select = c(voxel, term, estimate, std_error)
)
results$estimate <- round(results$estimate, 2)
results$std_error <- round(results$std_error, 2)
results
```

```{r quickstart-check, include=FALSE}
stopifnot(
  nrow(results) == 2L,
  all(is.finite(results$estimate)),
  all(is.finite(results$std_error)),
  identical(sort(unique(results$term)), c("faces", "scenes")),
  results$estimate[results$voxel == 1 & results$term == "faces"] >
    results$estimate[results$voxel == 1 & results$term == "scenes"]
)
```

The recovered pattern matches the simulation: the region responds more to
faces than to scenes. For real analyses, replace `bold` and `events` with
time-by-voxel data and the matching event table; the modeling interface is the
same.

## What it covers

- Build task and baseline models from formulas, including canonical, flexible,
  and custom hemodynamic response functions.
- Fit matrix-backed or imaging-backed datasets with ordinary, robust, and
  autoregressive regression paths.
- Define named t- and F-contrasts and export coefficient or statistic maps.
- Carry subject-level betas and standard errors—or t-statistics and degrees of
  freedom—into group analysis.
- Use controlled simulations, known-truth benchmark datasets, and explicit
  approximation diagnostics for accelerated methods.

## Fit and boundaries

fmrireg is the regression layer of an fMRI workflow. It expects preprocessed
time series and aligned event information; it does not perform motion
correction, anatomical registration, or spatial normalization. Matrix-backed
examples are useful for regions of interest and testing, while the dataset and
export guides cover image- and file-backed workflows.

The package is still experimental. Exact, AR-corrected, robust, and accelerated
engines have different statistical contracts; use the relevant guide rather
than assuming that every option is interchangeable.

## Documentation

- [Overview and workflow](https://bbuchsbaum.github.io/fmrireg/articles/fmrireg.html)
  — start with a complete fit, then unpack datasets, designs, and contrasts.
- [Dataset management](https://bbuchsbaum.github.io/fmrireg/articles/a_10_dataset.html),
  [simulation](https://bbuchsbaum.github.io/fmrireg/articles/a_08_simulation.html),
  [linear modeling](https://bbuchsbaum.github.io/fmrireg/articles/a_09_linear_model.html),
  and [contrasts](https://bbuchsbaum.github.io/fmrireg/articles/a_05_contrasts.html)
  — the core single-subject workflow.
- [Multi-subject fan-out](https://bbuchsbaum.github.io/fmrireg/articles/multisubject_fanout.html)
  and [group analysis](https://bbuchsbaum.github.io/fmrireg/articles/group_analysis.html)
  — move from subject fits to group inference.
- [Functional connectivity](https://bbuchsbaum.github.io/fmrireg/articles/functional_connectivity.html)
  — seed-regressor inference with autocorrelation and multiplicity control.
- [Sketched GLM](https://bbuchsbaum.github.io/fmrireg/articles/sketched-ar.html),
  [plugin development](https://bbuchsbaum.github.io/fmrireg/articles/plugin-development.html),
  and [benchmark datasets](https://bbuchsbaum.github.io/fmrireg/articles/benchmark_datasets.html)
  — advanced engines, extensions, and numerical oracles.
- [API reference](https://bbuchsbaum.github.io/fmrireg/reference/) — exported
  functions, arguments, and return values.

## Optional command-line wrapper

Install the wrapper into a directory on `PATH`:

```r
fmrireg::install_cli("~/.local/bin", overwrite = TRUE)
```

Then inspect available commands or bundled benchmark datasets:

```sh
fmrireg --help
fmrireg benchmark list
```

## Contributing

See [CONTRIBUTING.md](CONTRIBUTING.md) for the development workflow and test
commands. Please report reproducible bugs through the
[issue tracker](https://github.com/bbuchsbaum/fmrireg/issues).

## Citation

Run `citation("fmrireg")` for the current package citation.

## License

fmrireg is licensed under the [MIT License](LICENSE).
