---
title: "Hemodynamic Response Functions"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hemodynamic Response Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Hemodynamic response function

A hemodynamic response function (HRF) is used to model the temporal evoluation of an fMRI response to an experimental "event". An HRF is a function of *time* and usually peaks 4-6s afer the event onset.

There are a number of pre-defined HRFs in `fmrireg` that make it easy to work with them. For example, we can use the SPM "canonical" HRF. We use the function `hrf_blocked` which can generate flexible hemodynamic responses. 

```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
library(fmrireg)
```

```{r, message=FALSE, fig.width = 7, fig.height = 5}

library(magrittr)

## SPM canonical hrf
f1 <- hrf_blocked(hrf_spmg1, width=0)

## Gaussian hrf with default parameters (mean=6, sd=1.5)
f2 <- hrf_blocked(hrf_gaussian, width=0) 

plot(seq(0,20,by=.1), f1(seq(0,20,by=.1)), type="l", xlab="Time", ylab="BOLD response", main="Gaussian and SPM Canonical HRF", ylim=c(-.1, .3))
lines(seq(0,20,by=.1), f2(seq(0,20,by=.1)), type="l", col=2)

```

In the above, the "width" parameters tells the `hrf_block` function that the event is "instantaneous", i.e. that it is not extended in time. This is the standard way of modeling short-lasting (e.g. < 2 seconds) in fMRI.

## Changing Default Parameters

Standard HRF functions have default parameters that determine the shape of the response. For the Gaussian HRF, these parameters are easy to understand. The `mean` specifies the "peak" of the HRF and the `sd` parameter specifies its "width". 

```{r, message=FALSE,fig.width = 7, fig.height = 5}

## the 'gen_hrf' function takes an hrf function and returns a new function substituting the supplied parameters as arguments.
hrf_7_3 <- gen_hrf(hrf_gaussian, mean=7, sd=3)
hrf_5_2 <- gen_hrf(hrf_gaussian, mean=5, sd=2)
hrf_4_1 <- gen_hrf(hrf_gaussian, mean=4, sd=1)

time <- seq(0,20,by=.1)

vals1 <- hrf_7_3(time)
vals2 <- hrf_5_2(time)
vals3 <- hrf_4_1(time)


plot(seq(0,20,by=.1), vals1, type="l", xlab="Time", ylab="BOLD response", ylim=range(c(vals1, vals2, vals3)), main="Three Gaussian HRFs with different parameters")
lines(seq(0,20,by=.1), vals2, type="l", col=2)
lines(seq(0,20,by=.1), vals3, type="l", col=3)

```


## Setting the duration of a response

We can also set the duration of the input stimulus as follows:

```{r, message=FALSE,fig.width = 7, fig.height = 5}

hrf_w1 <- hrf_blocked(hrf_spmg1, width=1)
hrf_w2 <- hrf_blocked(hrf_spmg1, width=2)
hrf_w3 <- hrf_blocked(hrf_spmg1, width=3)

time <- seq(0,20,by=.1)
plot(time, hrf_w1(time), type="l", xlab="Time", ylab="BOLD response", ylim=range(0, 6), main="HRFs with Different Durations")
lines(time, hrf_w2(time), type="l", col=2)
lines(time, hrf_w3(time), type="l", col=3)

```

We can also require that the HRFs be normalized so that, regardless of the block duration, the peak amplitude of the HRF = 1.

```{r, message=FALSE,fig.width = 7, fig.height = 5}

hrf_w1 <- hrf_blocked(hrf_spmg1, width=1, normalize=TRUE)
hrf_w2 <- hrf_blocked(hrf_spmg1, width=2, normalize=TRUE)
hrf_w3 <- hrf_blocked(hrf_spmg1, width=3, normalize=TRUE)

plot(time, hrf_w1(time), type="l", xlab="Time", ylab="BOLD response", ylim=range(-.2, 1.2), main="HRFs with Different Durations, but normalized")
lines(time, hrf_w2(time), type="l", col=2)
lines(time, hrf_w3(time), type="l", col=3)

```

## Modeling HRF saturation with the `summation` argument

By default an increase in the duration results in the summation of a series of shifted responses, which is consistent with the notion that the hemodyamic response "adds up" or accumulates over time. But we can disable this additive behavior by setting `summate=FALSE`. This is useful if one assumes the HRF "saturates" (i..e does not increase in amplitude) with extended stimulation.


```{r, message=FALSE,fig.width = 7, fig.height = 5}

hrf_w1 <- hrf_blocked(hrf_spmg1, width=2, summate=FALSE)
hrf_w2 <- hrf_blocked(hrf_spmg1, width=4, summate=FALSE)
hrf_w3 <- hrf_blocked(hrf_spmg1, width=6, summate=FALSE)

time <- seq(0,24,by=.1)

plot(time, hrf_w1(time), type="l", xlab="Time", ylab="BOLD response", ylim=c(-.1,.25), main="Non-Summating HRFs with Different Durations ")
lines(time, hrf_w2(time), type="l", col=2)
lines(time, hrf_w3(time), type="l", col=3)

```

If we want the peak of the HRF to equal 1, we use the `normalize` argument as follows:

```{r, message=FALSE,fig.width = 7, fig.height = 5}

hrf_w1 <- hrf_blocked(hrf_spmg1, width=2, summate=FALSE, normalize=TRUE)
hrf_w2 <- hrf_blocked(hrf_spmg1, width=4, summate=FALSE, normalize=TRUE)
hrf_w3 <- hrf_blocked(hrf_spmg1, width=6, summate=FALSE, normalize=TRUE)

time <- seq(0,24,by=.1)

plot(time, hrf_w1(time), type="l", xlab="Time", ylab="BOLD response", ylim=c(-.1,1.1), main="Normalized and Non-Summating HRFs with Different Durations ")
lines(time, hrf_w2(time), type="l", col=2)
lines(time, hrf_w3(time), type="l", col=3)

```


## Multivariate HRF: B-splines and Derivatives

Sometimes we want to model an event with multiple "basis functions". In this case the HRF is a function from time -> *d*-dimensional vector, where *d* is the dimension of the basis set. 

```{r, message=FALSE,fig.width = 7, fig.height = 5}

## A third degree B-Spline with four bases.
hrf_bs_4 <- gen_hrf(hrf_bspline, N=4)
time <- seq(0,20, by=.1)
bmat1 <- hrf_bs_4(time)
matplot(time, bmat1, xlab="Time", ylab="BOLD Response", type='l', main="Bspline HRF basis set (N=4)")

## A first degree B-Spline with 12 bases.
hrf_bs_12 <- gen_hrf(hrf_bspline, degree=1, N=12)
bmat1 <- hrf_bs_12(time)
matplot(time, bmat1, xlab="Time", ylab="BOLD Response", type='l', main="Bspline HRF basis set (N=12)")

## the SPM + first derivative is a "built-in" HRF function.
bmat1 <- HRF_SPMG2(time)
matplot(time, bmat1, xlab="Time", ylab="BOLD Response", type='l', main="SPM + first derivative")

## the SPM + first derivative + second derivative is a "built-in" HRF function.
bmat1 <- HRF_SPMG3(time)
matplot(bmat1, xlab="Time", ylab="BOLD Response", type='l', main="SPM + 1st and 2nd derivative")

```


## Creating "lagged" and "blocked" HRFs

In some cases, it might be useful to create an HRF function that has a temporal offset (a "lag" or "delay"). We might also want to create an HRF that has a fixed width or duration that is larger than zero. Below are examples of suched lagged and blocked HRFs.


```{r, message=FALSE,fig.width = 7, fig.height = 5}

## constructing a series of lagged HRFs
hrf_lag_1 <- hrf_gaussian %>% gen_hrf_lagged(1) 
hrf_lag_3 <- hrf_gaussian %>% gen_hrf_lagged(3)
hrf_lag_5 <- hrf_gaussian %>% gen_hrf_lagged(5) 
time <- seq(0,22,by=.2)
plot(time,  hrf_lag_1(time), type='l', xlab="time", ylab="BOLD")
lines(time, hrf_lag_3(time), col=2)
lines(time, hrf_lag_5(time), col=3)

## constructing lagged and blocked HRFs
hrf_block_1 <- hrf_gaussian %>%  hrf_lagged(1) %>% hrf_blocked(width=1, normalize=TRUE)
hrf_block_3 <- hrf_gaussian %>% hrf_lagged(3) %>% hrf_blocked(width=3, normalize=TRUE)
hrf_block_5 <- hrf_gaussian %>% hrf_lagged(5) %>% hrf_blocked(width=5, normalize=TRUE) 
plot(time, hrf_block_1(time), type='l', xlab="time", ylab="BOLD")
lines(time, hrf_block_3(time), col=2)
lines(time, hrf_block_5(time), col=3)


```

## Constructing a custom basis set using a set of lagged HRFs

Rather than using a b-spline basis set, we could construct our own set of functions using a combination of lagged and/or blocked HRFs. Suppose we want a series of shifted regressor functions that span a 20s window. We can use a set of shifted `hrf`s to achieve this.


```{r, message=FALSE,fig.width = 7, fig.height = 5}

hrf_set <- do.call(gen_hrf_set,
                   lapply(seq(0,20,by=2), function(i) {  hrf_gaussian %>% hrf_lagged(i) }))

matplot(hrf_set(0:36), type='l', xlab="Time", ylab="BOLD Activity")

```

## Creating an "empirical" HRF from a measured hemodynamic respose

Suppose you have measured a hemodynamic response and what like to use that as a response model for further analyses? In this case you can use the function `gen_empirical_hrf` to create a hemodynamic response function. Below we extract our model by averaging over several simulated response profiles.


```{r, message=FALSE}

mat <- replicate(20, {
  h <- hrf_spmg1 %>% hrf_lagged(lag=runif(n=1, min=-2, max=2)) %>% hrf_blocked(width=runif(n=1, min=0, max=4))
  h(0:24)
})

profile <- rowMeans(mat)
emp_hrf <- gen_empirical_hrf(0:24, profile)
plot(seq(0, 24, by=.1), emp_hrf(seq(0, 24, by=.1)), type='l', xlab="time", ylab="response", main="Empircal HRF")

```

We can also create a set of empirical basis functions using, for example, principal components analysis of a set of HRFs with plausible parameters for lag and width. Here we will simulate 50 HRFs and take the first 3 principal components as our basis set.

```{r, message=FALSE}

mat <- replicate(50, {
  h <- hrf_spmg1 %>% hrf_lagged(lag=runif(n=1, min=-2, max=2)) %>% hrf_blocked(width=runif(n=1, min=0, max=2))
  h(0:24)
})
matplot(mat, type='l', ylab="response", xlab="Time")

```

Now we perform PCA pn the matrix of simulated responses and use the components as a basis set. The resulting function `emp_hrf_basis` can subsequently be used in regression modelling like any of the built-in hemodynamic response functions.

```{r}

pcs <- prcomp(mat)$x[,1:3]
pcs <- apply(pcs, 2, function(x) x - x[1])
pcs_hrfs <- apply(pcs, 2, function(x) gen_empirical_hrf(0:24, x/max(abs(x))))
emp_hrf_basis <- gen_hrf_set(pcs_hrfs$PC1,pcs_hrfs$PC2,pcs_hrfs$PC3 )
matplot(0:24, emp_hrf_basis(0:24), type='l', ylab="Response", xlab="Time", main="PCA basis functions")
```


