---
title: "Modeling Baseline and Nuisance Effects"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling Baseline and Nuisance Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(fmrireg)
library(dplyr)
library(ggplot2)
library(tidyr)
library(Matrix) # May be needed for sparse matrix ops internally
```

## The Purpose of a Baseline Model

In fMRI analysis, the BOLD signal contains not only task-related activity but also various sources of noise and drift. A **baseline model** aims to capture and account for this non-neuronal variance, ensuring that estimates of task effects are more accurate.

These baseline regressors typically model low-frequency scanner drift (using basis functions) and other known sources of noise like head motion parameters or physiological fluctuations.

`fmrireg` uses the `baseline_model()` function to construct this part of the design matrix.

## Modeling Drift with Basis Functions

A common approach to modeling slow scanner drift is to include a set of basis functions in the model for each run. `fmrireg` supports several basis sets specified via the `basis` argument:

*   `"poly"`: Polynomial functions (orthogonalized). Use `degree` or `poly_degree` to specify the order.
*   `"bs"`: B-spline basis functions. Use `degree` to specify the number of splines.
*   `"ns"`: Natural spline basis functions. Use `degree` to specify the number of splines.
*   `"constant"`: Includes only an intercept term for each run.

These basis functions are generated separately for each scanning run defined in the `sampling_frame`.

```{r basis_sets}
# Define a sampling frame for two runs of 100 scans each, TR=2s
TR <- 2
sframe <- sampling_frame(blocklens = c(100, 100), TR = TR)

# 1. Polynomial Basis (degree 5)
bmodel_poly <- baseline_model(basis = "poly", degree = 5, sframe = sframe)
print(bmodel_poly)

# 2. B-spline Basis (degree 5)
bmodel_bs <- baseline_model(basis = "bs", degree = 5, sframe = sframe)
# print(bmodel_bs)

# 3. Natural Spline Basis (degree 5)
bmodel_ns <- baseline_model(basis = "ns", degree = 5, sframe = sframe)
# print(bmodel_ns)

# 4. Constant Basis (Intercept only per run)
bmodel_const <- baseline_model(basis = "constant", sframe = sframe)
# print(bmodel_const)
```

## Visualizing Baseline Regressors

We can use `plot()` to visualize the generated regressors. The function returns a `ggplot` object.

**Term Naming:** By default, drift terms are named `baseline_<basis>_<degree>` (e.g., `baseline_poly_5`, `baseline_bs_3`). The constant term (if present) is named `constant`, and custom nuisance regressors are grouped under the term name `nuisance`.

**Plotting Specific Terms:** You can plot specific terms using the `term_name` argument. The plotting function supports partial matching, so you can often use shorter names like `"poly"`, `"bs"`, or `"nuisance"` if they uniquely identify a term.

```{r plot_basis}
# Plot the polynomial regressors (using partial match "poly")
plot(bmodel_poly, term_name = "poly")

# Plot the B-spline regressors (using partial match "bs")
plot(bmodel_bs, term_name = "bs")

# Plot the Natural spline regressors (using partial match "ns")
plot(bmodel_ns, term_name = "ns")

# Plotting the Constant (Intercept) term is generally not informative
# print(bmodel_const) # Shows it has a 'constant' term
```

Notice how the regressors are generated independently for each block (run) specified in the `sampling_frame`.

## Adding Arbitrary Nuisance Regressors

Beyond structured drift terms, we often want to include other regressors derived from data (e.g., motion parameters, physiological recordings, CSF signal). These can be added using the `nuisance_list` argument.

**Important:** The `nuisance_list` must be a *list* where each element corresponds to a scanning run. Each element should be a `data.frame` or `matrix` containing the nuisance regressors for that specific run. These will be grouped under the term name `"nuisance"`.

```{r nuisance_list}
# Simulate nuisance regressors (e.g., 6 motion parameters)
n_scans_run1 <- blocklens(sframe)[1]
n_scans_run2 <- blocklens(sframe)[2]

# Create nuisance data frames for each run
nuis_run1 <- as.data.frame(matrix(rnorm(n_scans_run1 * 6), n_scans_run1, 6))
names(nuis_run1) <- paste0("motion_", 1:6)

nuis_run2 <- as.data.frame(matrix(rnorm(n_scans_run2 * 6), n_scans_run2, 6))
names(nuis_run2) <- paste0("motion_", 1:6)

# Combine into a list
nuisance_regressors <- list(nuis_run1, nuis_run2)

# Create a baseline model including only these nuisance regressors
# (Set basis = NULL, degree = 0 to exclude drift terms)
bmodel_nuis_only <- baseline_model(basis = NULL, degree = 0, sframe = sframe, 
                                 nuisance_list = nuisance_regressors)
print(bmodel_nuis_only)

# Plot the nuisance regressors (term_name = "nuisance")
plot(bmodel_nuis_only, term_name = "nuisance") + 
  labs(title = "Nuisance Regressors Only (e.g., Motion)")
```

## Combining Basis Sets and Nuisance Regressors

You can include both a structured basis set (like polynomials) and custom nuisance regressors in the same baseline model.

```{r combined_baseline}
bmodel_combined <- baseline_model(basis = "poly", degree = 5, sframe = sframe, 
                                  nuisance_list = nuisance_regressors)
print(bmodel_combined)

# Check the terms included
term_names <- names(terms(bmodel_combined))
print(term_names) # e.g., constant, baseline_poly_5, nuisance
# baseline_terms(bmodel_combined) # Alias for terms()

# Plot the polynomial terms (using partial match "poly")
plot(bmodel_combined, term_name = "poly") + 
  labs(title = "Polynomial Drift Terms (from Combined Model)")

# Plot the nuisance terms (using exact match "nuisance")
plot(bmodel_combined, term_name = "nuisance") + 
  labs(title = "Nuisance Terms (from Combined Model)")
```

## Accessing the Design Matrix

The full design matrix for the baseline model (containing all basis and nuisance regressors, properly formatted per block) can be obtained using `design_matrix()`.

```{r design_matrix_baseline}
dmat_baseline <- design_matrix(bmodel_combined)

cat("Dimensions of baseline design matrix:", dim(dmat_baseline), "\n")
cat("Column names:", paste(colnames(dmat_baseline)[1:10], "..."), "\n")

# head(dmat_baseline[, 1:8]) # Show first few columns

# Visualize the full baseline matrix as a heatmap (optional)
# design_map.baseline_model(bmodel_combined, rotate_x_text = TRUE)
```

This baseline design matrix is combined with the task-related design matrix (from an `event_model`) within functions like `fmri_model` or `fmri_lm` to create the complete design matrix for GLM analysis.


