---
title: "Building Event-Related fMRI Models"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Event Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(fmrireg)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
```

## Introduction: Modeling Task Effects

An **event model** describes the expected BOLD signal changes related to experimental events (stimuli, conditions, responses, etc.). It forms the core of the task-related component of an fMRI General Linear Model (GLM).

`fmrireg` primarily uses the `event_model()` function to create these models. It takes experimental design information (event onsets, conditions, durations) and combines it with Hemodynamic Response Function (HRF) specifications to generate the task regressors.

This vignette focuses on the common formula-based interface for `event_model()`. There's also a list-based interface (`event_model.list`) and a more programmatic `create_event_model()` function for advanced use cases.

## A Simple Single-Factor Design (Single Run)

Consider a basic design with four stimulus types (*face*, *scene*, *tool*, *object*), each presented 4 times for 2s, separated by a variable ISI (4-7s). The total scan duration is 70 TRs (TR=2s).

```{r setup_simple_design}
TR <- 2
cond <- c("face", "scene", "tool", "object")
NSTIM <- length(cond) * 4

# Construct the design table
set.seed(123) # for reproducibility
simple_design <- data.frame(
  stim = factor(sample(rep(cond, 4))),
  ISI = sample(10:20, NSTIM, replace = TRUE),  # Increased ISI range for better spacing
  run = rep(1, NSTIM),
  trial = factor(1:NSTIM)
)

# Calculate onsets (cumulative sum of duration (2s) + ISI)
simple_design$onset <- cumsum(c(0, simple_design$ISI[-NSTIM] + 2))

# Define the sampling frame (temporal structure of the scan)
sframe_single_run <- sampling_frame(blocklens = 140, TR = TR)

head(simple_design)
```

Now, we create the `event_model`. The formula `onset ~ hrf(stim)` specifies that:

*   `onset` is the column in `simple_design` containing event start times.
*   `hrf(stim)` defines a term where the `stim` factor levels determine the conditions. Each level will be convolved with the default HRF (`HRF_SPMG1`).

The `block = ~ run` argument links events to scanning runs (here, just one run).

```{r create_simple_model}
emodel_simple <- event_model(onset ~ hrf(stim), 
                             data = simple_design, 
                             block = ~ run, 
                             sampling_frame = sframe_single_run)

# Print the model summary
print(emodel_simple)
```

## Visualizing the Event Model

We can plot the generated regressors using `plot()` (which uses `ggplot2`) or `plotly()` for an interactive version.

```{r plot_simple_model}
# Static plot (ggplot2)
plot(emodel_simple)

# Interactive plot (plotly)
# plotly(emodel_simple)
```

The plot shows the predicted BOLD timecourse for each condition (`stim` level) convolved with the default HRF.

## Design with Multiple Runs

Let's extend the design to two runs.

```{r setup_multi_block}
# Construct a design table with two runs
design_list <- lapply(1:2, function(run_idx) {
  df <- data.frame(
    stim = factor(sample(rep(cond, 4))),
    ISI = sample(10:20, NSTIM, replace = TRUE),  # Increased ISI range for better spacing
    run = rep(run_idx, NSTIM)
  )
  df$onset <- cumsum(c(0, df$ISI[-NSTIM] + 2))
  df
})
design_multi_run <- bind_rows(design_list)

# Sampling frame for two runs of 140s each
sframe_multi_run <- sampling_frame(blocklens = c(140, 140), TR = TR)

head(design_multi_run)
```

The `event_model` call remains the same, but now `block = ~ run` correctly maps events to their respective runs based on the `run` column and the `sampling_frame`.

```{r create_multi_block_model}
emodel_multi_run <- event_model(onset ~ hrf(stim), 
                                data = design_multi_run, 
                                block = ~ run, 
                                sampling_frame = sframe_multi_run)

print(emodel_multi_run)
plot(emodel_multi_run) + 
  facet_wrap(~.block, ncol = 1) # Show blocks separately
```

## Two-Factor Design

Now consider a design crossing stimulus type (`stim`) with task instruction (`task`: attend vs. ignore).

```{r setup_two_factor}
cond1 <- c("face", "scene", "tool", "object")
cond2 <- c("attend", "ignore")
comb <- expand.grid(stim = cond1, task = cond2)
NSTIM_TF <- nrow(comb) * 4 # 8 conditions * 4 reps per run

# Design for two runs
design_two_factor_list <- lapply(1:2, function(run_idx) {
  ind <- sample(rep(1:nrow(comb), length.out = NSTIM_TF))
  df <- data.frame(
    stim = factor(comb$stim[ind]),
    task = factor(comb$task[ind]),
    ISI = sample(6:15, NSTIM_TF, replace = TRUE),  # Increased ISI range for better spacing
    run = rep(run_idx, NSTIM_TF)
  )
  df$onset <- cumsum(c(0, df$ISI[-NSTIM_TF] + 2))
  df
})
design_two_factor <- bind_rows(design_two_factor_list)

# Sampling frame for two runs, potentially longer
sframe_two_factor <- sampling_frame(blocklens = c(200, 200), TR = TR)

head(design_two_factor)
```

The formula `onset ~ hrf(stim, task)` automatically creates regressors for the interaction of `stim` and `task`.

```{r create_two_factor_model}
emodel_two_factor <- event_model(onset ~ hrf(stim, task), 
                                 data = design_two_factor, 
                                 block = ~ run, 
                                 sampling_frame = sframe_two_factor)
print(emodel_two_factor)

# Plotting all interaction terms can be busy; consider plotly
# plot(emodel_two_factor)
# plotly(emodel_two_factor)
```

## Amplitude Modulation (Parametric Regressors)

We can model how a continuous variable (like reaction time, RT) modulates the amplitude of the BOLD response. This is done by including the modulating variable within the `hrf()` call.

```{r setup_pmod}
# Use the simple single-run design and add a simulated RT column
simple_design$RT <- rnorm(nrow(simple_design), mean = 700, sd = 100)

# It's often recommended to center parametric modulators
simple_design$RT_centered <- scale(simple_design$RT, center = TRUE, scale = FALSE)[,1]
head(simple_design)
```

Here, `hrf(RT_centered)` creates a single parametric regressor reflecting RT variations, orthogonal to the average stimulus effect modeled by `hrf(stim)`.

```{r create_pmod_model}
emodel_pmod <- event_model(onset ~ hrf(stim) + hrf(RT_centered), 
                           data = simple_design, 
                           block = ~ run, 
                           sampling_frame = sframe_single_run)
print(emodel_pmod)

# Plot the RT parametric modulator term
plot(emodel_pmod, term_name = "RT_centered") 
# plotly(emodel_pmod, term_name = "RT_centered")
```

## Interaction Between Factors and Amplitude Modulation

We can also model how a parametric modulator interacts with a factor. For example, does RT modulate the response differently for faces vs. scenes? The formula `hrf(stim, RT_centered)` creates separate RT modulators for each level of `stim`.

```{r create_pmod_interaction}
emodel_pmod_int <- event_model(onset ~ hrf(stim) + hrf(stim, RT_centered), 
                               data = simple_design, 
                               block = ~ run, 
                               sampling_frame = sframe_single_run)
print(emodel_pmod_int)

# Plot the interaction term - results in multiple regressors
plot(emodel_pmod_int, term_name = "stim:RT_centered")
# plotly(emodel_pmod_int, term_name = "stim:RT_centered")
```

## Specifying Different HRFs

By default, `hrf()` uses the SPM canonical HRF (`HRF_SPMG1`). You can specify a different HRF basis for any term using the `basis` argument within the `hrf()` call in the model formula. 

Here are the main ways to specify the `basis`:

1.  **By Name (String):** Use a recognized name for common HRFs:
    *   `"spmg1"` (default), `"spmg2"`, `"spmg3"`
    *   `"gaussian"`, `"gamma"`
    *   `"bspline"` or `"bs"` (use `nbasis` to set number of splines)
    *   `"tent"` (use `nbasis` to set number of tent functions)

2.  **Using Pre-defined `HRF` Objects:** Pass an exported `HRF` object directly. Examples include `HRF_SPMG1`, `HRF_GAUSSIAN`, `HRF_SPMG2`, `HRF_SPMG3`, `HRF_BSPLINE`, `HRF_DAGUERRE`, `HRF_DAGUERRE_BASIS`.

3.  **Using Custom Functions `f(t)`:** Provide your own function definition that accepts time `t` as the first argument.

4.  **Using `gen_hrf()` Results:** Create a customized HRF using `gen_hrf()`, `hrf_blocked()`, `hrf_lagged()`, etc., and pass the resulting `HRF` object.

**Discovering Available Options:** You can easily explore all available HRF types using the `list_available_hrfs()` function:

```{r list_hrfs}
# List basic information about available HRFs
list_available_hrfs()

# Get detailed descriptions
list_available_hrfs(details = TRUE)
```

For more details and examples of different HRF types, refer to the documentation for `hrf()` and the "Hemodynamic Response Functions" vignette (`vignette("a_01_hemodynamic_response", package = "fmrireg")`).

Here are a few examples within the `event_model` context:

```{r specify_hrf}
# Example 1: Using basis name string "gaussian"
# Model stim effects with SPMG1 and RT effects with Gaussian HRF
emodel_diff_hrf <- event_model(onset ~ hrf(stim, basis="spmg1") + hrf(RT_centered, basis="gaussian"), 
                               data = simple_design, 
                               block = ~ run, 
                               sampling_frame = sframe_single_run)
print(emodel_diff_hrf)
plot(emodel_diff_hrf, term_name = "RT_centered") # Plot the Gaussian RT regressor

# Example 2: Using a pre-defined HRF object (SPMG3)
# Model stim effects with SPM+derivs (HRF_SPMG3)
emodel_spmg3 <- event_model(onset ~ hrf(stim, basis=HRF_SPMG3), 
                            data = simple_design, 
                            block = ~ run, 
                            sampling_frame = sframe_single_run)
print(emodel_spmg3) # Note nbasis=3 for the hrf
plot(emodel_spmg3, term_name = "stim") # Plotting shows the 3 basis functions for one condition
# plotly(emodel_spmg3) # Better for exploring many conditions

# Example 3: Using a custom function (simple linear ramp)
linear_ramp_hrf <- function(t) { ifelse(t > 0 & t < 10, t/10, 0) }
emodel_custom_hrf <- event_model(onset ~ hrf(stim, basis=linear_ramp_hrf), 
                                 data = simple_design, 
                                 block = ~ run, 
                                 sampling_frame = sframe_single_run)
print(emodel_custom_hrf)
plot(emodel_custom_hrf, term_name = "stim")

# Example 4: Using gen_hrf() to create a lagged Gaussian
lagged_gauss <- gen_hrf(hrf_gaussian, lag = 2, name = "Lagged Gaussian")
emodel_gen_hrf <- event_model(onset ~ hrf(stim, basis = lagged_gauss), 
                              data = simple_design, 
                              block = ~ run, 
                              sampling_frame = sframe_single_run)
print(emodel_gen_hrf)
plot(emodel_gen_hrf, term_name = "stim")
```

## Trialwise Models for Beta-Series Analysis

To model each trial individually (e.g., for beta-series correlation), use `trialwise()` in the formula. This creates a separate regressor for every single event specified by the `onset` variable.

```{r create_trialwise}
emodel_trialwise <- event_model(onset ~ trialwise(), 
                                data = simple_design, 
                                block = ~ run, 
                                sampling_frame = sframe_single_run)
print(emodel_trialwise)

# Plotting trialwise models can be very dense!
# It generates one condition per trial.
# plot(emodel_trialwise)
# plotly(emodel_trialwise) # Use plotly to explore
```

## Accessing Model Components

Once an `event_model` is created, you can extract its components:

```{r access_components}
# List the terms in the model
terms_list <- terms(emodel_pmod_int)
print(names(terms_list))

# Get all condition names (column names in the design matrix)
conds <- conditions(emodel_pmod_int)
cat("\nFirst 10 conditions:", head(conds, 10), "...\n")

# Extract the full design matrix
dmat_events <- design_matrix(emodel_pmod_int)
cat("\nEvent design matrix dimensions:", dim(dmat_events), "\n")
# head(dmat_events[, 1:6])

# Contrast weights and F-contrasts can also be extracted (see contrasts vignette)
# contrast_weights(emodel_pmod_int)
# Fcontrasts(emodel_pmod_int)
```

## Visualizing the Design Matrix and Correlations

Two functions help visualize the structure of the event design matrix:

*   `design_map`: Shows the matrix values as a heatmap.
*   `correlation_map`: Shows the correlation between regressors.

```{r matrix_viz}
# Heatmap of the design matrix for the 2-factor model
design_map(emodel_two_factor, rotate_x_text = TRUE) +
  labs(title = "Design Matrix Heatmap (Two-Factor Model)")

# Correlation map for the same model
correlation_map(emodel_two_factor, rotate_x_text = TRUE) +
  labs(title = "Regressor Correlation Map (Two-Factor Model)")

```

This `event_model` represents the task-related part of the fMRI model. It's typically combined with a `baseline_model` (containing drift terms, intercepts, nuisance regressors) within functions like `fmri_model` or `fmri_lm` to perform a full GLM analysis.
