---
title: Contrasts and Hypothesis Tests
author: Bradley R. Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Contrasts and Hypothesis Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(fmrireg)
library(dplyr)
library(knitr)
library(ggplot2)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=7, fig.height=6, message=FALSE, warning=TRUE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Introduction to Contrasts

Statistical contrasts turn questions about differences between experimental
conditions into linear hypotheses about fitted coefficients. `fmrireg` provides
constructors for defining those hypotheses, inspecting their weights, and
fitting them with an fMRI linear model.

This vignette shows how to turn a scientific question into weights, inspect
those weights before fitting, and extract the resulting estimates and test
statistics. If you have not yet fitted a model, start with
`vignette("a_09_linear_model", package = "fmrireg")`.

## Example: A 2x2 Factorial Design

To illustrate the contrast functionalities, let's use a simple two-by-two factorial design. We have two factors:

1.  **category**: with levels "face" and "scene"
2.  **attention**: with levels "attend" and "ignore"

We'll assume each unique condition is repeated twice within a single run.

First, we construct the event table representing this design:

```{r design_setup}
design <- expand.grid(category = c("face", "scene"), 
                      attention = c("attend", "ignore"), 
                      replication = c(1, 2))
design$onset <- seq(1, 100, length.out = nrow(design)) # Assign arbitrary onsets
design$block <- rep(1, nrow(design))                  # Single block (run)

# Ensure factors are factors
design$category <- factor(design$category)
design$attention <- factor(design$attention)

kable(design, caption = "2x2 Experimental Design Table")

# Define a sampling frame and create the event model
sframe <- sampling_frame(blocklens = 120, TR = 2)
emodel <- event_model(onset ~ hrf(category, attention), 
                      block = ~block, 
                      data = design, 
                      sampling_frame = sframe)

# Extract the event term for contrast calculation
# In this simple model, there's only one event term
event_term <- terms(emodel)[[1]] 

cell_order <- unique(design[c("category", "attention")])
cell_labels <- paste(cell_order$category, cell_order$attention, sep = " / ")
label_weight_matrix <- function(x, labels = cell_labels) {
  out <- as.matrix(x)
  stopifnot(nrow(out) == length(labels))
  rownames(out) <- labels
  out
}

kable(cell_order,
      row.names = FALSE,
      caption = "Cells within the 'category:attention' event term")
```

This `event_term` object encapsulates the structure of our experimental conditions and will be used to compute contrast weights.

## Basic Contrasts: `pair_contrast`

The most common type of contrast compares the average activation of one set of conditions against another. The `pair_contrast` function provides a convenient way to define such sum-to-zero contrasts.

### Defining Pair Contrasts

`pair_contrast` takes two formulas, `A` and `B`, defining the conditions to compare, and a mandatory `name`.

Let's define contrasts for the main effects of *category* (face vs. scene) and *attention* (attend vs. ignore):

```{r define_pair_contrasts}
# Main effect of category: face > scene
con_face_vs_scene <- pair_contrast(~ category == "face", 
                                 ~ category == "scene", 
                                 name = "face_vs_scene")

# Main effect of attention: attend > ignore
con_attend_vs_ignore <- pair_contrast(~ attention == "attend", 
                                    ~ attention == "ignore", 
                                    name = "attend_vs_ignore")
```

### Computing Contrast Weights

Contrast specifications are abstract until applied to a specific model term structure. The `contrast_weights` function computes the numerical weights based on the levels within the term.

```{r compute_pair_weights}
wts_face_vs_scene <- contrast_weights(con_face_vs_scene, event_term)
wts_attend_vs_ignore <- contrast_weights(con_attend_vs_ignore, event_term)

cat("Weights for 'face_vs_scene':\n")
kable(label_weight_matrix(wts_face_vs_scene$weights),
      col.names = wts_face_vs_scene$name)

cat("\nWeights for 'attend_vs_ignore':\n")
kable(label_weight_matrix(wts_attend_vs_ignore$weights),
      col.names = wts_attend_vs_ignore$name)
```

Notice how `pair_contrast` automatically averages over the levels *not* mentioned in the formulas. For `face_vs_scene`, it averages over 'attend' and 'ignore' within each category level before contrasting. The resulting weights sum to zero (0.5 * 2 + (-0.5) * 2 = 0).

## Unit Contrasts: Comparing to Baseline

Sometimes, we want to test whether activation for a condition (or set of conditions) is significantly different from the implicit baseline (often represented by the intercept in the model). `unit_contrast` is used for this purpose. It creates contrasts that sum to 1.

```{r unit_contrasts}
con_face_vs_baseline <- unit_contrast(
  ~ category, name = "face_gt_baseline",
  where = ~ category == "face"
)
con_attend_vs_baseline <- unit_contrast(
  ~ attention, name = "attend_gt_baseline",
  where = ~ attention == "attend"
)

wts_face_vs_baseline <- contrast_weights(con_face_vs_baseline, event_term)
wts_attend_vs_baseline <- contrast_weights(con_attend_vs_baseline, event_term)

cat("Weights for 'face_gt_baseline':\n")
kable(label_weight_matrix(wts_face_vs_baseline$weights),
      col.names = wts_face_vs_baseline$name)

cat("\nWeights for 'attend_gt_baseline':\n")
kable(label_weight_matrix(wts_attend_vs_baseline$weights),
      col.names = wts_attend_vs_baseline$name)
```
These weights average the specified conditions and compare them against zero (the implicit baseline).

```{r check-unit-weights, include=FALSE}
stopifnot(
  identical(as.numeric(wts_face_vs_baseline$weights), c(0.5, 0, 0.5, 0)),
  identical(as.numeric(wts_attend_vs_baseline$weights), c(0.5, 0.5, 0, 0))
)
```

## General Contrasts with `contrast()`

The `contrast()` function allows for more complex contrasts defined by a single formula expression. This is useful for interactions or specific linear combinations of conditions.

Let's define the interaction contrast: (face:attend - face:ignore) - (scene:attend - scene:ignore). This tests if the effect of attention differs between categories.

```{r general_contrast}
# Interaction contrast
con_interaction <- contrast(
  ~ (`face:attend` - `face:ignore`) - (`scene:attend` - `scene:ignore`),
  name = "category_X_attention"
)

wts_interaction <- contrast_weights(con_interaction, event_term)

cat("Weights for 'category_X_attention':\n")
kable(label_weight_matrix(wts_interaction$weights),
      col.names = wts_interaction$name)
```

Note: In the formula for `contrast()`, condition names are formed by joining factor levels with colons (e.g., `face:attend`). When condition names contain special characters like colons, they must be enclosed in backticks. The `contrast_weights` function evaluates this formula in an environment where each condition name corresponds to a column vector (1 for that condition, 0 otherwise).

## Contrasts for Main Effects and Interactions

While `pair_contrast` and `contrast` are flexible, `fmrireg` provides convenience functions for standard ANOVA-like contrasts.

*   `oneway_contrast`: Generates contrasts for the main effect of a single factor (sum-to-zero coding).
*   `interaction_contrast`: Generates contrasts for interaction effects between two or more factors.

These often result in multiple contrast columns (F-contrasts) testing the overall effect.

```{r anova_contrasts}
# Main effect of category (will produce 1 contrast vector)
con_main_category <- oneway_contrast(~ category, name = "Main_Category")
wts_main_category <- contrast_weights(con_main_category, event_term)

# Interaction effect (will produce 1 contrast vector for a 2x2 design)
con_int_cat_att <- interaction_contrast(~ category * attention, name = "Interaction_CatAtt")
wts_int_cat_att <- contrast_weights(con_int_cat_att, event_term)

cat("Weights for 'Main_Category' (oneway_contrast):\n")
kable(label_weight_matrix(wts_main_category$weights))

cat("\nWeights for 'Interaction_CatAtt' (interaction_contrast):\n")
kable(label_weight_matrix(wts_int_cat_att$weights))
```
Note that the weights generated by `interaction_contrast(~ category * attention)` match those we manually specified using `contrast()`. For factors with more levels, these functions would generate multiple orthogonal contrast columns.

## Polynomial Contrasts for Ordered Factors

If a factor represents ordered levels (e.g., different task difficulty levels, time points), `poly_contrast` can test for trends (linear, quadratic, etc.).

Let's add an 'intensity' factor to our design:

```{r poly_design}
design_poly <- expand.grid(category = c("face", "scene"), 
                           intensity = c(1, 2, 3), # Ordered factor
                           replication = c(1))
design_poly$onset <- seq(1, 60, length.out = nrow(design_poly))
design_poly$block <- rep(1, nrow(design_poly)) 
design_poly$intensity <- factor(design_poly$intensity, ordered = TRUE)

kable(design_poly, caption = "Design with Ordered 'intensity' Factor")

emodel_poly <- event_model(onset ~ hrf(category, intensity), 
                           block = ~block, 
                           data = design_poly, 
                           sampling_frame = sframe)
event_term_poly <- terms(emodel_poly)[[1]]
```

Now, define a polynomial contrast to test for linear and quadratic trends of intensity:

```{r poly_contrast}
con_poly_intensity <- poly_contrast(~ intensity, degree = 2, name = "Intensity_Trend")
wts_poly_intensity <- contrast_weights(con_poly_intensity, event_term_poly)
poly_cell_order <- unique(design_poly[c("category", "intensity")])
poly_labels <- paste(
  poly_cell_order$category,
  paste("intensity", poly_cell_order$intensity),
  sep = " / "
)

cat("Weights for 'Intensity_Trend' (poly_contrast, degree=2):\n")
kable(label_weight_matrix(wts_poly_intensity$weights, poly_labels))
```
The output has two columns: `Intensity_Trend_1` for the linear trend and
`Intensity_Trend_2` for the quadratic trend.

## Helper Functions for Common Comparisons

Two helpers simplify common multi-level comparisons:

*   `pair_contrast`: Creates a pairwise comparison between two levels of a factor.
*   `one_against_all_contrast`: Compares each level against the average of all other levels.

```{r helper_contrasts}
# Pairwise contrast for category using pair_contrast
con_pairwise_cat <- pair_contrast(~ category == "face", 
                                 ~ category == "scene", 
                                 name = "cat_face_scene")

# Compare each attention level vs the other
con_one_all_att <- one_against_all_contrast(levels(design$attention), facname = "attention")

# Since con_one_all_att is already a contrast_set, we need to extract its elements
# and combine them properly with the single contrast
all_helper_contrasts <- list(con_pairwise_cat)
all_helper_contrasts <- append(all_helper_contrasts, con_one_all_att)
con_set_helpers <- do.call(contrast_set, all_helper_contrasts)

# Calculate weights (demonstrating contrast_set)
wts_helpers <- contrast_weights(con_set_helpers, event_term)

# Display weights for one_against_all
cat("Weights for 'con_attend_vs_other':\n")
kable(label_weight_matrix(wts_helpers$con_attend_vs_other$weights))
cat("\nWeights for 'con_ignore_vs_other':\n")
kable(label_weight_matrix(wts_helpers$con_ignore_vs_other$weights))
```

## Grouping Contrasts: `contrast_set`

You can group multiple contrast specifications using `contrast_set`. When `contrast_weights` is called on a `contrast_set`, it returns a named list of computed contrast weight objects.

```{r contrast_set_example}
# Combine several previously defined contrasts
all_contrasts <- contrast_set(
  con_face_vs_scene,
  con_attend_vs_ignore,
  con_interaction,
  con_face_vs_baseline
)

print(all_contrasts)

# Compute weights for the entire set
all_weights <- contrast_weights(all_contrasts, event_term)

# Access weights for a specific contrast within the set
cat("\nAccessing weights for 'face_vs_scene' from the set:\n")
kable(label_weight_matrix(all_weights$face_vs_scene$weights))
```

## Applying Contrasts in `fmri_lm`

Contrasts are typically specified within the `hrf()` function in the `fmri_lm` formula. You can provide a single contrast specification or a `contrast_set`.

```{r simulate-contrast-data, include=FALSE}
signal_design <- as.matrix(design_matrix(emodel))
true_cell_effects <- cbind(
  voxel1 = c(1.2, 0.4, 0.8, 0.2),
  voxel2 = c(1.0, 0.5, 0.7, 0.3),
  voxel3 = c(0.8, 0.3, 0.6, 0.2)
)
rownames(true_cell_effects) <- colnames(signal_design)
set.seed(99)
ysim <- signal_design %*% true_cell_effects +
  matrix(rnorm(120 * 3, sd = 0.15), 120, 3)
dataset_sim <- matrix_dataset(
  ysim, TR = 2, run_length = 120, event_table = design
)
```

```{r apply-contrasts-model}
# Fit model with the contrast set defined earlier
fmri_fit <- fmri_lm(
  onset ~ hrf(category, attention, contrasts = all_contrasts),
  block = ~ block, dataset = dataset_sim
)
```

The simulated cell effects make the face-minus-scene contrast positive in all
three voxels. This gives the example a known direction rather than relying on
an arbitrary noise realization.

## Extracting Contrast Results

After fitting a model with contrasts, you can extract the results using standard accessor functions, specifying `type = "contrasts"` or `type = "F"`:

*   `coef(fmri_fit, type = "contrasts")`: Estimated contrast values.
*   `stats(fmri_fit, type = "contrasts")`: t-statistics for contrasts.
*   `standard_error(fmri_fit, type = "contrasts")`: Standard errors for contrasts.
*   `stats(fmri_fit, type = "F")`: F-statistics (if F-contrasts were defined, e.g., via `oneway_contrast`).

```{r extract-results}
contrast_estimates <- coef(fmri_fit, type = "contrasts")
contrast_tstats <- stats(fmri_fit, type = "contrasts")
contrast_se <- standard_error(fmri_fit, type = "contrasts")
contrast_p <- p_values(fmri_fit, type = "contrasts")

contrast_summary_numeric <- data.frame(
  contrast = names(contrast_estimates),
  estimate = as.numeric(contrast_estimates[1, ]),
  std_error = as.numeric(contrast_se[1, ]),
  statistic = as.numeric(contrast_tstats[1, ]),
  p_value = as.numeric(contrast_p[1, ]),
  df = as.numeric(fmri_fit$result$df$inference)[1]
)
contrast_summary <- transform(
  contrast_summary_numeric,
  p_value = format.pval(p_value, digits = 3, eps = 1e-4)
)
kable(contrast_summary, digits = 3,
      row.names = FALSE,
      caption = "Contrast results for voxel 1")
```

```{r check-contrast-results, include=FALSE}
stopifnot(
  all(is.finite(as.matrix(contrast_estimates))),
  all(is.finite(as.matrix(contrast_se))),
  all(is.finite(as.matrix(contrast_tstats))),
  all(is.finite(as.matrix(contrast_p))),
  all(contrast_estimates$face_vs_scene > 0),
  !any(contrast_summary$p_value == "0")
)
```

The estimate is the effect size in response-data units. The standard error is
its uncertainty, and the t statistic is their ratio; the p-value additionally
depends on the reported inference degrees of freedom. These p-values describe
planned tests at one voxel. A whole-brain analysis must also declare how it
controls multiplicity across voxels and across any unplanned contrasts.

## Execute an omnibus F-contrast

A multi-column F-contrast asks whether any of several restrictions is nonzero.
Here a three-level factor gives a two-degree-of-freedom omnibus test. This is
different from choosing one signed pairwise difference.

```{r execute-f-contrast}
three_level_events <- data.frame(
  onset = seq(5, 80, length.out = 12),
  condition = factor(rep(c("faces", "objects", "scenes"), 4)),
  run = 1L
)
three_level_frame <- sampling_frame(100, TR = 2)
three_level_event <- event_model(
  onset ~ hrf(condition), block = ~ run,
  data = three_level_events, sampling_frame = three_level_frame
)
three_level_X <- as.matrix(design_matrix(three_level_event))
set.seed(404)
three_level_Y <- three_level_X %*% matrix(c(1.0, 0.4, -0.2), ncol = 1) +
  matrix(rnorm(100, sd = 0.15), ncol = 1)
three_level_dataset <- matrix_dataset(
  three_level_Y, TR = 2, run_length = 100,
  event_table = three_level_events
)

condition_omnibus <- oneway_contrast(~ condition, name = "condition_F")
omnibus_fit <- fmri_lm(
  onset ~ hrf(condition, contrasts = contrast_set(condition_omnibus)),
  block = ~ run, dataset = three_level_dataset
)
omnibus_weights <- contrast_weights(
  condition_omnibus, terms(three_level_event)[[1]]
)$weights
omnibus_F <- as.numeric(stats(omnibus_fit, type = "F")$condition_F)
omnibus_df1 <- ncol(omnibus_weights)
omnibus_df2 <- as.numeric(omnibus_fit$result$df$inference)[1]
omnibus_p <- pf(omnibus_F, omnibus_df1, omnibus_df2, lower.tail = FALSE)

omnibus_summary <- data.frame(
  F = omnibus_F,
  numerator_df = omnibus_df1,
  denominator_df = omnibus_df2,
  p_value = format.pval(omnibus_p, digits = 3, eps = 1e-4)
)
kable(
  omnibus_summary,
  digits = 3,
  row.names = FALSE,
  caption = "Omnibus test of equality across three condition effects"
)
```

```{r check-f-contrast, include=FALSE}
stopifnot(
  omnibus_df1 == 2L,
  is.finite(omnibus_F), omnibus_F > 0,
  is.finite(omnibus_p), omnibus_p < 0.05,
  omnibus_summary$p_value != "0"
)
```

## Visualizing Contrast Weights

Before fitting, inspect the cell weights using labels that match the scientific
conditions. This catches sign errors without exposing internal design-column
names.

```{r plot-contrasts-example, fig.alt="Heatmap of signed contrast weights for the four category-by-attention experimental cells across four planned hypotheses."}
weight_matrix <- do.call(cbind, lapply(
  all_weights,
  function(weight) as.numeric(weight$weights[, 1])
))
cell_order <- unique(design[c("category", "attention")])
rownames(weight_matrix) <- paste(cell_order$category, cell_order$attention,
                                 sep = " / ")
colnames(weight_matrix) <- names(all_weights)
weight_long <- as.data.frame(as.table(weight_matrix), stringsAsFactors = FALSE)
names(weight_long) <- c("condition", "contrast", "weight")

ggplot(weight_long, aes(contrast, condition, fill = weight)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f", weight)), size = 3.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B") +
  labs(x = NULL, y = NULL, fill = "Weight",
       title = "Planned contrast weights by experimental cell") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```
The full-design helper `plot_contrasts()` remains useful when you also need to
audit how these weights align with nuisance and baseline columns.

## Working with Contrast Weights

The computed contrast weights can be used directly in statistical analyses or exported to other analysis packages. The weights matrix shows exactly how each experimental condition contributes to the contrast:

```{r view_weights}
interaction_weights <- data.frame(
  condition = rownames(weight_matrix),
  weight = as.numeric(wts_interaction$weights[, 1])
)
kable(interaction_weights,
      row.names = FALSE,
      caption = "Weights for the category-by-attention interaction")
```

## Conclusion

This vignette showed how to define pairwise comparisons (`pair_contrast`),
baseline tests (`unit_contrast`), formula-based contrasts (`contrast`), polynomial
trends (`poly_contrast`), and multi-degree-of-freedom effects
(`oneway_contrast`, `interaction_contrast`). Inspect the resulting weight matrix,
fit the declared hypothesis with `fmri_lm`, and report the effect, uncertainty,
degrees of freedom, and multiplicity procedure together.

## Next

- `vignette("multisubject_fanout", package = "fmrireg")` — Reuse the model across subjects
- `vignette("group_analysis", package = "fmrireg")` — Group analysis
