---
title: "Contrasts in fmrireg"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contrasts in fmrireg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(fmrireg)
library(dplyr)
library(knitr)
library(ggplot2)
options(width=100)
knitr::opts_chunk$set(fig.width=7, fig.height=6, message=FALSE, warning=FALSE)
```

## Introduction to Contrasts

Statistical contrasts are a fundamental component of *f*MRI analyses, allowing researchers to test specific hypotheses about differences in brain activity between experimental conditions. The `fmrireg` package provides a flexible and powerful system for defining, computing, and applying contrasts to linear models fitted to fMRI data.

This vignette explores the various ways to specify contrasts in `fmrireg`, from simple pairwise comparisons to more complex interactions and polynomial trends.

## Example: A 2x2 Factorial Design

To illustrate the contrast functionalities, let's use a simple two-by-two factorial design. We have two factors:

1.  **category**: with levels "face" and "scene"
2.  **attention**: with levels "attend" and "ignore"

We'll assume each unique condition is repeated twice within a single run.

First, we construct the event table representing this design:

```{r design_setup}
design <- expand.grid(category = c("face", "scene"), 
                      attention = c("attend", "ignore"), 
                      replication = c(1, 2))
design$onset <- seq(1, 100, length.out = nrow(design)) # Assign arbitrary onsets
design$block <- rep(1, nrow(design))                  # Single block (run)

# Ensure factors are factors
design$category <- factor(design$category)
design$attention <- factor(design$attention)

kable(design, caption = "2x2 Experimental Design Table")

# Define a sampling frame and create the event model
sframe <- sampling_frame(blocklens = 120, TR = 2)
emodel <- event_model(onset ~ hrf(category, attention), 
                      block = ~block, 
                      data = design, 
                      sampling_frame = sframe)

# Extract the event term for contrast calculation
# In this simple model, there's only one event term
event_term <- terms(emodel)[[1]] 

kable(cells(event_term), caption = "Cells within the 'category:attention' event term")
```

This `event_term` object encapsulates the structure of our experimental conditions and will be used to compute contrast weights.

## Basic Contrasts: `pair_contrast`

The most common type of contrast compares the average activation of one set of conditions against another. The `pair_contrast` function provides a convenient way to define such sum-to-zero contrasts.

### Defining Pair Contrasts

`pair_contrast` takes two formulas, `A` and `B`, defining the conditions to compare, and a mandatory `name`.

Let's define contrasts for the main effects of *category* (face vs. scene) and *attention* (attend vs. ignore):

```{r define_pair_contrasts}
# Main effect of category: face > scene
con_face_vs_scene <- pair_contrast(~ category == "face", 
                                 ~ category == "scene", 
                                 name = "face_vs_scene")

# Main effect of attention: attend > ignore
con_attend_vs_ignore <- pair_contrast(~ attention == "attend", 
                                    ~ attention == "ignore", 
                                    name = "attend_vs_ignore")
```

### Computing Contrast Weights

Contrast specifications are abstract until applied to a specific model term structure. The `contrast_weights` function computes the numerical weights based on the levels within the term.

```{r compute_pair_weights}
wts_face_vs_scene <- contrast_weights(con_face_vs_scene, event_term)
wts_attend_vs_ignore <- contrast_weights(con_attend_vs_ignore, event_term)

cat("Weights for 'face_vs_scene':\n")
kable(wts_face_vs_scene$weights, col.names = wts_face_vs_scene$name)

cat("\nWeights for 'attend_vs_ignore':\n")
kable(wts_attend_vs_ignore$weights, col.names = wts_attend_vs_ignore$name)
```

Notice how `pair_contrast` automatically averages over the levels *not* mentioned in the formulas. For `face_vs_scene`, it averages over 'attend' and 'ignore' within each category level before contrasting. The resulting weights sum to zero (0.25 * 2 + (-0.25) * 2 = 0).

## Unit Contrasts: Comparing to Baseline

Sometimes, we want to test whether activation for a condition (or set of conditions) is significantly different from the implicit baseline (often represented by the intercept in the model). `unit_contrast` is used for this purpose. It creates contrasts that sum to 1.

```{r unit_contrasts}
con_face_vs_baseline <- unit_contrast(~ category == "face", name = "face_gt_baseline")
con_attend_vs_baseline <- unit_contrast(~ attention == "attend", name = "attend_gt_baseline")

wts_face_vs_baseline <- contrast_weights(con_face_vs_baseline, event_term)
wts_attend_vs_baseline <- contrast_weights(con_attend_vs_baseline, event_term)

cat("Weights for 'face_gt_baseline':\n")
kable(wts_face_vs_baseline$weights, col.names = wts_face_vs_baseline$name)

cat("\nWeights for 'attend_gt_baseline':\n")
kable(wts_attend_vs_baseline$weights, col.names = wts_attend_vs_baseline$name)
```
These weights average the specified conditions and compare them against zero (the implicit baseline).

## General Contrasts with `contrast()`

The `contrast()` function allows for more complex contrasts defined by a single formula expression. This is useful for interactions or specific linear combinations of conditions.

Let's define the interaction contrast: (face:attend - face:ignore) - (scene:attend - scene:ignore). This tests if the effect of attention differs between categories.

```{r general_contrast}
# Interaction contrast
con_interaction <- contrast(
  ~ (face.attend - face.ignore) - (scene.attend - scene.ignore),
  name = "category_X_attention"
)

wts_interaction <- contrast_weights(con_interaction, event_term)

cat("Weights for 'category_X_attention':\n")
kable(wts_interaction$weights, col.names = wts_interaction$name)
```

Note: In the formula for `contrast()`, condition names are formed by joining factor levels with dots (e.g., `face.attend`). The `contrast_weights` function evaluates this formula in an environment where each condition name corresponds to a column vector (1 for that condition, 0 otherwise).

## Contrasts for Main Effects and Interactions

While `pair_contrast` and `contrast` are flexible, `fmrireg` provides convenience functions for standard ANOVA-like contrasts.

*   `oneway_contrast`: Generates contrasts for the main effect of a single factor (sum-to-zero coding).
*   `interaction_contrast`: Generates contrasts for interaction effects between two or more factors.

These often result in multiple contrast columns (F-contrasts) testing the overall effect.

```{r anova_contrasts}
# Main effect of category (will produce 1 contrast vector)
con_main_category <- oneway_contrast(~ category, name = "Main_Category")
wts_main_category <- contrast_weights(con_main_category, event_term)

# Interaction effect (will produce 1 contrast vector for a 2x2 design)
con_int_cat_att <- interaction_contrast(~ category * attention, name = "Interaction_CatAtt")
wts_int_cat_att <- contrast_weights(con_int_cat_att, event_term)

cat("Weights for 'Main_Category' (oneway_contrast):\n")
kable(wts_main_category$weights)

cat("\nWeights for 'Interaction_CatAtt' (interaction_contrast):\n")
kable(wts_int_cat_att$weights)
```
Note that the weights generated by `interaction_contrast(~ category * attention)` match those we manually specified using `contrast()`. For factors with more levels, these functions would generate multiple orthogonal contrast columns.

## Polynomial Contrasts for Ordered Factors

If a factor represents ordered levels (e.g., different task difficulty levels, time points), `poly_contrast` can test for trends (linear, quadratic, etc.).

Let's add an 'intensity' factor to our design:

```{r poly_design}
design_poly <- expand.grid(category = c("face", "scene"), 
                           intensity = c(1, 2, 3), # Ordered factor
                           replication = c(1))
design_poly$onset <- seq(1, 60, length.out = nrow(design_poly))
design_poly$block <- rep(1, nrow(design_poly)) 
design_poly$intensity <- factor(design_poly$intensity, ordered = TRUE)

kable(design_poly, caption = "Design with Ordered 'intensity' Factor")

emodel_poly <- event_model(onset ~ hrf(category, intensity), 
                           block = ~block, 
                           data = design_poly, 
                           sampling_frame = sframe)
event_term_poly <- terms(emodel_poly)[[1]]
```

Now, define a polynomial contrast to test for linear and quadratic trends of intensity:

```{r poly_contrast}
con_poly_intensity <- poly_contrast(~ intensity, degree = 2, name = "Intensity_Trend")
wts_poly_intensity <- contrast_weights(con_poly_intensity, event_term_poly)

cat("Weights for 'Intensity_Trend' (poly_contrast, degree=2):\n")
kable(wts_poly_intensity$weights)
```
The output has two columns: `poly1` for the linear trend and `poly2` for the quadratic trend.

## Helper Functions for Common Comparisons

Two helpers simplify common multi-level comparisons:

*   `pairwise_contrasts`: Creates all unique pairwise comparisons between levels of a factor.
*   `one_against_all_contrast`: Compares each level against the average of all other levels.

```{r helper_contrasts}
# All pairwise contrasts for category (trivial here, but useful for >2 levels)
con_pairwise_cat <- pairwise_contrasts(levels(design$category), 
                                     facname = "category", 
                                     name_prefix = "cat")
# (This would create 'cat_face_scene')

# Compare each attention level vs the other
con_one_all_att <- one_against_all_contrast(levels(design$attention), facname = "attention")

# Combine the *elements* of the two contrast sets into a single list
combined_contrasts <- c(con_pairwise_cat, con_one_all_att)
# Create a new contrast_set from the combined list elements using do.call
con_set_helpers <- do.call(contrast_set, combined_contrasts)

# Calculate weights (demonstrating contrast_set)
wts_helpers <- contrast_weights(con_set_helpers, event_term)

# Display weights for one_against_all
cat("Weights for 'con_attend_vs_other':\n")
kable(wts_helpers$con_attend_vs_other$weights)
cat("\nWeights for 'con_ignore_vs_other':\n")
kable(wts_helpers$con_ignore_vs_other$weights)
```

## Grouping Contrasts: `contrast_set`

You can group multiple contrast specifications using `contrast_set`. When `contrast_weights` is called on a `contrast_set`, it returns a named list of computed contrast weight objects.

```{r contrast_set_example}
# Combine several previously defined contrasts
all_contrasts <- contrast_set(
  con_face_vs_scene,
  con_attend_vs_ignore,
  con_interaction,
  con_face_vs_baseline
)

print(all_contrasts)

# Compute weights for the entire set
all_weights <- contrast_weights(all_contrasts, event_term)

# Access weights for a specific contrast within the set
cat("\nAccessing weights for 'face_vs_scene' from the set:\n")
kable(all_weights$face_vs_scene$weights)
```

## Applying Contrasts in `fmri_lm`

Contrasts are typically specified within the `hrf()` function in the `fmri_lm` formula. You can provide a single contrast specification or a `contrast_set`.

```{r apply_contrasts_model, eval=FALSE}
# Simulate some simple data for demonstration
ysim <- matrix(rnorm(120 * 3), 120, 3) # 3 voxels
dataset_sim <- matrix_dataset(ysim, TR = 2, run_length = 120, event_table = design)

# Fit model with the contrast set defined earlier
fmri_fit <- fmri_lm(
  formula = onset ~ hrf(category, attention, contrasts = all_contrasts),
  block = ~ block,
  dataset = dataset_sim,
  strategy = "chunkwise" # Use chunkwise for matrix_dataset
)

# Print summary of the fitted model (shows contrasts)
print(fmri_fit)
```

*(Note: The above fitting code is set to `eval=FALSE` to avoid lengthy computation in the vignette build, but demonstrates the principle.)*

## Extracting Contrast Results

After fitting a model with contrasts, you can extract the results using standard accessor functions, specifying `type = "contrasts"` or `type = "F"`:

*   `coef(fmri_fit, type = "contrasts")`: Estimated contrast values.
*   `stats(fmri_fit, type = "contrasts")`: t-statistics for contrasts.
*   `standard_error(fmri_fit, type = "contrasts")`: Standard errors for contrasts.
*   `stats(fmri_fit, type = "F")`: F-statistics (if F-contrasts were defined, e.g., via `oneway_contrast`).

```{r extract_results, eval=FALSE}
# Extract estimated contrast values
contrast_estimates <- coef(fmri_fit, type = "contrasts")
kable(contrast_estimates, caption = "Estimated Contrast Values")

# Extract t-statistics
contrast_tstats <- stats(fmri_fit, type = "contrasts")
kable(contrast_tstats, caption = "Contrast t-statistics")

# Extract standard errors
contrast_se <- standard_error(fmri_fit, type = "contrasts")
kable(contrast_se, caption = "Contrast Standard Errors")
```

## Visualizing Contrast Weights

The `plot_contrasts` function provides a heatmap visualization of the contrast weights applied across all regressors in the design matrix (including baseline terms if present).

```{r plot_contrasts_example}
# We need to add contrasts *directly* to the event model for plotting
emodel_with_cons <- event_model(
  onset ~ hrf(category, attention, contrasts = all_contrasts),
  block = ~ block, 
  data = design, 
  sampling_frame = sframe
)

# Plot the contrasts (using default baseline model)
plot_contrasts(emodel_with_cons, rotate_x_text = TRUE, coord_fixed = FALSE)
```
This plot helps verify that contrasts are specified correctly relative to the full model design matrix.

## Exporting Contrasts to AFNI GLT Format

For users interacting with AFNI, the `to_glt` function can convert computed contrast weights into AFNI's General Linear Test (GLT) string format.

```{r afni_glt}
glt_interaction <- to_glt(wts_interaction)
cat("AFNI GLT string for interaction contrast:\n")
print(glt_interaction$glt_str)
```

## Conclusion

The `fmrireg` package offers a comprehensive system for defining and applying statistical contrasts in fMRI analysis. From simple pairwise comparisons (`pair_contrast`) and baseline tests (`unit_contrast`) to complex formula-based definitions (`contrast`), trend analysis (`poly_contrast`), and ANOVA-style effects (`oneway_contrast`, `interaction_contrast`), researchers have fine-grained control over hypothesis testing. The integration with `fmri_lm` and visualization tools like `plot_contrasts` facilitates robust and interpretable fMRI modeling.
