---
title: "Single Trial Beta Estimation (Draft)"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single Trial Beta Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Estimating single-trial event amplitudes for use in "beta series" analyses and MVPA analyses.

It is common in fMRI analysis want to obtain estimates of BOLD amplitude for each *trial* (rather than condition) in an experiment. Estimation of single-trial beta coefficients can be done in `fmrireg` with the `estimate_betas` function.

First we load in an experimental design.

```{r, message=FALSE, warning=FALSE}

library(dplyr)
library(fmrireg)
facedes <- read.table(system.file("extdata", "face_design.txt", package = "fmrireg"), header=TRUE) %>%
  filter(run == 1)
## we had a constant value to the design used later to model the mean stimulus response over all trials.
facedes$constant <- factor(rep(1, nrow(facedes)))

```

Now we generate an fMRI data using random values.

```{r, message=FALSE, warning=FALSE}

D <- 5
scans <- lapply(1:length(unique(facedes$run)), function(i) {
    arr <- array(rnorm(D*D*D*300), c(D,D,D, 300))
    bspace <- neuroim2::NeuroSpace(dim=c(D,D,D,300))
    neuroim2::NeuroVec(arr, bspace)
})
  
mask <- neuroim2::LogicalNeuroVol(array(rnorm(D*D*D), c(D,D,D)) > 0, neuroim2::NeuroSpace(dim=c(D,D,D)))
  
```

Now we create a instance of type `fmri_mem_dataset` which encapsulates infromation about the data, temporal layout, and experimental design of an "in memory" fmri dataset.

```{r, message=FALSE,warning=FALSE}

dset <- fmri_mem_dataset(scans=scans, 
                 mask=mask, 
                 TR=1.5, 
                 event_table=facedes)

```

Now we're ready to run the beta estimation analysis. There are two components to the model: a fixed effects formula and a random effects formula. The fixed effects part models the condition-specific responses, and the random effects part models the trial-specific responses. For the random effect part, we want to model every onset separately, so we use the `trialwise` hrf modeling function. For the fixed part, we use a standard `hrf` modeling function.

The method we use is "pls" for **partial least squares** (via the `pls` package), which effectively acts as a regularized least squares estimation procedure. Simulations show that `pls` produces trialwise beta estimates for closely space onsets (e.g. overlapping hrfs) that are similar to the "least squares sum" approach (citation needed). By modeling the mean effect with `fixed = onset ~ hrf(constant)` the "random effects" `onset ~ trialwise()` capture the deviations from that mean, with regularization achieved by using partial least squares estimation with a small number of components (here, n=3). Other fixed effects can also be modeled, and could include events of no interest that should be covaried out when estimating the beta coefficients of interest.

When the onsets are *not* closely space together, ordinary least squares using the "ols" method is appropriate. For method "ols", however, a constant fixed effect should not be used, since it is a linear combination of the random effects, so we omit the "fixed" term.

```{r, message=FALSE, warning=FALSE}
  

res <- estimate_betas(dset, fixed = onset ~ hrf(constant), ran = onset ~ trialwise(), block = ~ run, 
                       method="pls", ncomp=3)

names(res)

res_ols <- estimate_betas(dset, ran = onset ~ trialwise(), block = ~ run, 
                       method="ols", ncomp=3)

  
```

We can plot the fixed effects model, which here is a single regressor. 

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}

plot(res$fixed_model)

```

We can also plot the random effects model, where there is one regressor per trial.


```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}

plot(res$ran_model)

```





