---
title: "Single-Trial Beta Estimation for fMRI Analysis"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single Trial Beta Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dplyr)
library(fmrireg)
```

## Introduction: Why Estimate Single-Trial Betas?

Standard fMRI analysis often estimates the average BOLD response amplitude for each experimental condition. However, sometimes it's valuable to estimate the response amplitude for *each individual trial* or event. This approach yields a "beta series" - a sequence of beta coefficients, one for each trial.

Single-trial beta estimates are useful for several advanced analysis techniques:

*   **Beta Series Correlation:** Analyzing functional connectivity by correlating the trial-by-trial beta series between different brain regions.
*   **Multivoxel Pattern Analysis (MVPA):** Using the pattern of single-trial betas across voxels as input for machine learning classifiers to decode experimental conditions or cognitive states.
*   **Relating Neural Activity to Behavior:** Examining trial-by-trial correlations between brain activation (betas) and behavioral measures (e.g., reaction time, accuracy).

`fmrireg` provides the `estimate_betas()` function to compute these single-trial coefficients using various estimation methods.

## Setting up the Data and Design

First, let's load an example experimental design table and create a dummy fMRI dataset.

```{r load_design}
# Load design, filter for one run for simplicity
facedes <- read.table(system.file("extdata", "face_design.txt", package = "fmrireg"), header = TRUE) %>% 
  dplyr::filter(run == 1)

# Add a constant factor - useful for modeling the mean response
facedes$constant <- factor(rep(1, nrow(facedes)))

head(facedes)
```

Now, we generate some random fMRI data for demonstration purposes.

```{r generate_data}
D <- 5 # Dimension for a small simulated brain
# Create a list containing one NeuroVec for the single run
scans <- list(local({
  arr <- array(rnorm(D*D*D*300), c(D, D, D, 300))
  bspace <- neuroim2::NeuroSpace(dim = c(D, D, D, 300))
  neuroim2::NeuroVec(arr, bspace)
}))

# Create a mask volume
mask <- neuroim2::LogicalNeuroVol(array(rnorm(D*D*D), c(D,D,D)) > 0, 
                                  neuroim2::NeuroSpace(dim = c(D, D, D)))
```

Finally, we create an `fmri_mem_dataset` instance, which holds the data, timing information (TR), and the experimental design table.

```{r create_dataset}
dset <- fmri_mem_dataset(scans = scans, 
                         mask = mask, 
                         TR = 1.5, 
                         event_table = facedes)
print(dset)
```

## Estimating Betas with `estimate_betas`

The `estimate_betas()` function requires specifying the model structure and the estimation method.

**Model Structure:**

The model typically has two main parts, specified via formulas:

1.  `ran = ...`: This defines the **random effects** component, which models the trial-specific responses. To estimate a separate beta for *every* trial, we use the special `trialwise()` term. For example, `ran = onset ~ trialwise()` creates one regressor for each row in the `event_table` based on the `onset` column.
2.  `fixed = ...` (Optional): This defines the **fixed effects** component, modeling effects assumed to be constant across trials. This is often used to model the *average* response across all trials or specific conditions, allowing the `trialwise()` betas to represent deviations from this mean. It can also include nuisance regressors (e.g., events of no interest) that should be regressed out *before* estimating the trialwise betas of interest.
3.  `block = ...`: Specifies the run structure, essential for correctly handling timing and baseline modeling across runs.

**Estimation Methods (`method` argument):**

Estimating single-trial betas can be challenging, especially with rapid event-related designs where the HRF responses to consecutive trials overlap significantly, leading to high collinearity between trial regressors. `estimate_betas` offers several methods to address this:

*   `"ols"` (Ordinary Least Squares): Standard GLM estimation. **Warning:** Highly sensitive to collinearity. Often yields unstable/unreliable beta estimates in typical fMRI designs. Generally *not* recommended for beta series analysis unless trials are very widely spaced. If used, typically *omit* the `fixed` term if it models an average response, as it would be perfectly collinear with the sum of `trialwise` regressors.
*   `"lss"` / `"lss_cpp"` (Least Squares - Separate): Implements the approach described by Mumford et al. (2012). It fits a separate GLM for each trial, where the trial of interest is modeled explicitly and all *other* trials are collapsed into a single nuisance regressor. This aims to reduce collinearity between the trial of interest and other trials. `lss_cpp` uses a faster C++ implementation.
*   `"pls"` (Partial Least Squares): Uses PLS regression (via the `pls` package). Acts as a regularized regression method, often performing well with collinear regressors. By modeling the mean effect in `fixed` and trial deviations in `ran`, PLS can effectively estimate the trialwise components.
*   `"fracridge"` (Fractional Ridge Regression): Implements ridge regression but with a fractional penalty, offering another form of regularization useful for collinear designs. Requires the `fracridge` package.
*   `"mixed"` / `"mixed_cpp"` (Linear Mixed Models): Uses LMMs (via `rrBLUP` or an internal C++ solver) to model trial effects as random deviations from fixed effects. Can be robust but potentially computationally intensive.
*   `"r1"` / `"r1_glms"`: Implements Rank-1 GLM methods that can jointly estimate the HRF shape along with betas (Pedregosa et al., 2015). `r1_glms` is adapted for single-trial estimation.
*   `"lowrank_hrf"`: A newer method that first estimates voxelwise HRFs using `find_best_hrf` and then uses these estimated HRFs in an LSS-like procedure. Can be beneficial if HRF shape varies significantly across the brain.

**Choosing a Method:** `"lss"` (or `"lss_cpp"`) is a common and often effective choice. `"pls"` and `"fracridge"` offer good regularization for highly collinear designs. `"mixed"` provides a statistically principled approach but can be slower. `"ols"` should generally be avoided.

**Example using PLS:**

Here, we use PLS. We model the average response across all trials using `fixed = onset ~ hrf(constant)` and the trial-specific deviations using `ran = onset ~ trialwise()`.

```{r estimate_pls}
# Using ::: for internal function visibility during development/testing
# In package use, call fmrireg::estimate_betas
res_pls <- fmrireg:::estimate_betas(dset, 
                                  fixed = onset ~ hrf(constant), 
                                  ran = onset ~ trialwise(), 
                                  block = ~ run, 
                                  method = "pls", 
                                  ncomp = 3) # ncomp controls PLS regularization

print(names(res_pls))
```

**Example using OLS (for illustration):**

If using OLS, we typically omit the `fixed` term modeling the average response to avoid collinearity.

```{r estimate_ols}
# Note: OLS is generally not recommended due to collinearity
res_ols <- fmrireg:::estimate_betas(dset, 
                                  ran = onset ~ trialwise(), # No fixed term
                                  block = ~ run, 
                                  method = "ols") 
```

## Interpreting the Output

The `estimate_betas` function returns a list object of class `fmri_betas` containing:

*   `betas_fixed`: A `NeuroVec` (if input was `fmri_dataset`) or matrix containing the estimated coefficients for the `fixed` effects model.
*   `betas_ran`: A `NeuroVec` or matrix containing the estimated single-trial coefficients for the `ran` effects model (the beta series).
*   `design_fixed`: The design matrix for the fixed effects.
*   `design_ran`: The design matrix for the random (trialwise) effects.
*   `design_base`: The design matrix for the baseline model (intercepts, drift, etc.).
*   `fixed_model`: The `event_model` object for the fixed effects.
*   `ran_model`: The `event_model` object for the random effects.
*   `basemod`: The baseline model used.

Let's look at the dimensions of the random betas matrix (voxels x trials):

```{r show_betas_ran}
# For matrix_dataset input, betas_ran is a matrix
# For fmri_dataset input, it would be a NeuroVec
if (is.matrix(res_pls$betas_ran)) {
  print(dim(res_pls$betas_ran))
} else {
  print(neuroim2::space(res_pls$betas_ran))
}
```

## Visualizing the Models

We can plot the design matrices for the fixed and random effects components.

The fixed effects model here contains just one regressor representing the average response (`constant`).

```{r plot_fixed, fig.width=8, fig.height=3}
if (!is.null(res_pls$fixed_model)) {
  plot(res_pls$fixed_model)
}
```

The random effects model contains one regressor for *each trial*.

```{r plot_random, fig.width=8, fig.height=5}
# Plotting all trial regressors can be busy!
# Consider using plotly for interaction if needed, or plotting subsets.
# plot(res_pls$ran_model)
# For demonstration, let's plot the design map instead:
design_map(res_pls$ran_model, rescale_cols = FALSE) + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle=90, hjust=1, size=6))
```

The resulting `betas_ran` contains the estimated amplitudes for each of these trial-specific regressors, ready for use in further analyses like beta-series correlation or MVPA.





