---
title: 03 Simulating fMRI Data
author: Bradley R. Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: >
  %\VignetteIndexEntry{03 Simulating fMRI Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: ochre
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load-packages}
library(fmrireg)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stats)
set.seed(123)
```

## Introduction to fMRI Data Simulation

When developing or validating an fMRI analysis pipeline, you need data where the ground truth is known -- real fMRI data doesn't come with known effect sizes or noise parameters. Simulation fills this gap. The `fmrireg` package offers several functions to simulate fMRI data with varying levels of complexity:

1. **`simulate_bold_signal`**: Simulates clean BOLD responses for multiple experimental conditions
2. **`simulate_noise_vector`**: Generates realistic fMRI noise with temporal autocorrelation, drift, and physiological components
3. **`simulate_simple_dataset`**: Combines signal and noise for a complete dataset based on SNR
4. **`simulate_fmri_matrix`**: Creates multiple time series with shared event timing but column-specific variation in parameters

This vignette demonstrates how to use these functions to create realistic fMRI simulations for various purposes.

## Simulating Clean BOLD Signals

Let's start with `simulate_bold_signal`, which generates a clean hemodynamic response signal for multiple experimental conditions.

```{r simulate-bold-signals}
# Simulate 3 conditions with different amplitudes
sim <- simulate_bold_signal(ncond = 3, amps = c(1.0, 1.5, 2.0), TR = 2)

# Extract the data
time <- sim$mat[,1]  # First column contains time
signals <- sim$mat[,-1]  # Other columns contain condition signals
```

We reshape the data into long format so ggplot2 can map each condition to a separate color.

```{r reshape-bold-data}
df <- data.frame(
  Time = time,
  Cond1 = signals[,1],
  Cond2 = signals[,2],
  Cond3 = signals[,3]
)

df_long <- tidyr::pivot_longer(df, cols = c(Cond1, Cond2, Cond3),
                        names_to = "Condition",
                        values_to = "Response")
```

```{r plot-bold-signals}
ggplot(df_long, aes(x = Time, y = Response, color = Condition)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Simulated BOLD Signals for Three Conditions",
       x = "Time (seconds)",
       y = "BOLD Response",
       color = "Condition") +
  scale_color_brewer(palette = "Set1")
```

Adding dashed vertical lines at each event onset makes the relationship between stimulus timing and the hemodynamic response easier to see.

```{r plot-bold-with-onsets}
ggplot(df_long, aes(x = Time, y = Response, color = Condition)) +
  geom_line() +
  geom_vline(xintercept = sim$onset, linetype = "dashed", alpha = 0.3) +
  theme_minimal() +
  labs(title = "Simulated BOLD Signals with Event Onsets",
       x = "Time (seconds)",
       y = "BOLD Response",
       color = "Condition") +
  scale_color_brewer(palette = "Set1")
```

The function returns a list containing:
- **`onset`**: Event onset times
- **`condition`**: Condition labels for each event
- **`mat`**: Matrix with time points and BOLD responses for each condition

You can control:
- Number of conditions (`ncond`)
- Number of repetitions per condition (`nreps`)
- HRF shape (`hrf`)
- Amplitudes for each condition (`amps`)
- Inter-stimulus interval range (`isi`)
- Amplitude variability (`ampsd`)

## Simulating Realistic fMRI Noise

The `simulate_noise_vector` function generates realistic fMRI noise by combining various noise sources common in real fMRI data. To understand its components, let's simulate and visualize them separately and combined.

```{r setup-noise-params}
n_timepoints <- 200
TR <- 2
time <- seq(0, (n_timepoints - 1) * TR, by = TR)
set.seed(123)
```

First, pure white noise -- random fluctuations with no temporal structure.

```{r simulate-white-noise}
noise_white <- simulate_noise_vector(n_timepoints, TR = TR,
                                  ar = 0, ma = 0,
                                  drift_amplitude = 0, physio = FALSE, sd = 1)
```

Next, ARMA noise introduces temporal autocorrelation, which makes the signal "smoother" than white noise.

```{r simulate-arma-noise}
noise_arma <- simulate_noise_vector(n_timepoints, TR = TR,
                                 ar = c(0.6), ma = c(0.3),
                                 drift_amplitude = 0, physio = FALSE, sd = 1)
```

Scanner drift is a slow oscillation that shifts the baseline over the course of a run.

```{r simulate-drift}
drift_freq <- 1/128
drift_amplitude <- 2
noise_drift <- drift_amplitude * sin(2 * pi * drift_freq * time)
```

Physiological noise adds quasi-periodic fluctuations at respiratory and cardiac-like frequencies.

```{r simulate-physio-noise}
noise_cardiac <- 1.5 * sin(2 * pi * 0.3 * time)
noise_respiratory <- 1.0 * sin(2 * pi * 0.8 * time)
noise_physio <- noise_cardiac + noise_respiratory
```

Finally, we combine the ARMA, drift, and physiological components to mimic what a real scanner would produce.

```{r combine-noise}
noise_combined <- noise_arma + noise_drift + noise_physio
```

```{r reshape-noise-data}
noise_df <- data.frame(
  Time = time,
  White_Noise = noise_white,
  ARMA_Noise = noise_arma,
  Drift_Component = noise_drift,
  Physiological_Noise = noise_physio,
  Combined_Noise = noise_combined
)

noise_long <- tidyr::pivot_longer(noise_df,
                                  cols = -Time,
                                  names_to = "NoiseType",
                                  values_to = "Signal")

noise_long$NoiseType <- factor(noise_long$NoiseType,
                              levels = c("White_Noise", "ARMA_Noise",
                                        "Drift_Component", "Physiological_Noise",
                                        "Combined_Noise"),
                              labels = c("White Noise", "ARMA Noise",
                                        "Drift Component", "Physiological Noise",
                                        "Combined Noise"))
```

```{r plot-noise-timecourses, fig.height=9}
ggplot(noise_long, aes(x = Time, y = Signal, color = NoiseType)) +
  geom_line() +
  facet_wrap(~NoiseType, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Simulated fMRI Noise Components: Time Courses",
       x = "Time (seconds)",
       y = "Signal Amplitude")
```

### Power Spectrum Analysis

Power spectra reveal the frequency content of each noise type. White noise has a flat spectrum, while ARMA noise concentrates power at lower frequencies.

```{r calculate-spectrum-helper, include=FALSE}
calculate_spectrum <- function(signal, TR) {
  # Detrend
  signal <- signal - mean(signal)

  # Calculate FFT
  n <- length(signal)
  fft_result <- fft(signal)
  # Get power - divide by n to normalize
  power <- Mod(fft_result)^2 / n

  # Get positive frequencies only (up to Nyquist)
  nyquist_index <- floor(n/2) + 1
  freq <- seq(0, 1/(2*TR), length.out = nyquist_index)
  power_pos <- power[1:nyquist_index]

  # For one-sided spectrum, multiply by 2 (except DC and Nyquist)
  if (n > 1) {
    power_pos[2:(length(power_pos)-ifelse(n %% 2 == 0, 1, 0))] <-
      2 * power_pos[2:(length(power_pos)-ifelse(n %% 2 == 0, 1, 0))]
  }

  data.frame(Frequency = freq, Power = power_pos)
}
```

```{r compute-spectra}
spec_white <- calculate_spectrum(noise_white, TR)
spec_arma <- calculate_spectrum(noise_arma, TR)
spec_drift <- calculate_spectrum(noise_drift, TR)
spec_physio <- calculate_spectrum(noise_physio, TR)
spec_combined <- calculate_spectrum(noise_combined, TR)
```

```{r label-spectra}
spec_white$NoiseType <- "White Noise"
spec_arma$NoiseType <- "ARMA Noise"
spec_drift$NoiseType <- "Drift Component"
spec_physio$NoiseType <- "Physiological Noise"
spec_combined$NoiseType <- "Combined Noise"

spec_df <- rbind(spec_white, spec_arma, spec_drift, spec_physio, spec_combined)

spec_df$NoiseType <- factor(spec_df$NoiseType,
                           levels = c("White Noise", "ARMA Noise",
                                     "Drift Component", "Physiological Noise",
                                     "Combined Noise"))
```

We also compute a high-resolution drift spectrum from a longer signal to better resolve the low-frequency peak.

```{r compute-hires-drift}
n_long <- 1024
time_long <- seq(0, (n_long - 1) * TR, by = TR)
drift_long <- drift_amplitude * sin(2 * pi * drift_freq * time_long)
spec_drift_long <- calculate_spectrum(drift_long, TR)
spec_drift_long$NoiseType <- "Drift Component (High Resolution)"
```

Each panel below shows the power spectrum of a different noise component. The dashed green line in the drift panel shows the high-resolution version.

```{r plot-spectra-faceted, fig.height=9}
ggplot() +
  geom_line(data = spec_df, aes(x = Frequency, y = Power, color = NoiseType)) +
  geom_line(data = subset(spec_drift_long, Frequency <= 0.05),
           aes(x = Frequency, y = Power), color = "darkgreen", linetype = "dashed") +
  theme_minimal() +
  facet_wrap(~NoiseType, ncol = 1, scales = "free_y") +
  theme(legend.position = "none") +
  labs(title = "Power Spectra of Different Noise Components",
       x = "Frequency (Hz)",
       y = "Power") +
  scale_color_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(0, 0.25))
```

Plotting all spectra on a single log-scaled axis makes it easy to compare their relative magnitudes.

```{r plot-spectra-log}
ggplot(spec_df, aes(x = Frequency, y = Power, color = NoiseType)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Power Spectra of Different Noise Components (Log Scale)",
       x = "Frequency (Hz)",
       y = "Power (log scale)") +
  scale_y_log10() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "top") +
  coord_cartesian(xlim = c(0, 0.25))
```

Zooming in on the very low-frequency region confirms the drift component peaks at the expected frequency (marked by the red dotted line).

```{r plot-drift-zoom}
drift_freq_idx <- which.min(abs(spec_df$Frequency - drift_freq))

ggplot() +
  geom_line(data = subset(spec_df, NoiseType == "Drift Component"),
           aes(x = Frequency, y = Power), color = "darkgreen") +
  geom_line(data = subset(spec_drift_long, Frequency <= 0.05),
           aes(x = Frequency, y = Power), color = "darkgreen", linetype = "dashed") +
  geom_vline(xintercept = drift_freq, linetype = "dotted", color = "red") +
  theme_minimal() +
  labs(title = "Zoomed Power Spectrum of Drift Component",
       subtitle = paste("Expected peak at", drift_freq, "Hz"),
       x = "Frequency (Hz)",
       y = "Power") +
  coord_cartesian(xlim = c(0, 0.05))
```

The simulation shows five distinct types of noise components:

1. **White Noise**: Random fluctuations with equal power across all frequencies (flat power spectrum).

2. **ARMA Noise**: Temporally autocorrelated noise with a characteristic "smoothed" appearance. The power spectrum shows more power at lower frequencies.

3. **Drift Component**: A very slow oscillation typical of scanner drift or physiological trends. The power spectrum shows a dominant peak at a very low frequency.

4. **Physiological Noise**: Regular oscillations at frequencies corresponding to respiration (~0.3 Hz) and cardiac-like activity (~0.8 Hz). In real fMRI data with TR=2s, cardiac frequencies (~1.2 Hz) would be aliased.

5. **Combined Noise**: All components together, creating a complex noise structure typical of real fMRI data. The power spectrum shows features from all contributing components.

These components, when added to task-related signals, create realistic fMRI time series.

## Creating a Complete Dataset with Signal and Noise

The `simulate_simple_dataset` function combines clean signals and noise to create a complete fMRI dataset with a specified signal-to-noise ratio (SNR).

```{r simulate-snr-datasets}
set.seed(42)

data_snr_1.0 <- simulate_simple_dataset(ncond = 3, TR = 2, snr = 1.0)
data_snr_0.5 <- simulate_simple_dataset(ncond = 3, TR = 2, snr = 0.5)
data_snr_0.2 <- simulate_simple_dataset(ncond = 3, TR = 2, snr = 0.2)
```

```{r create-plot-df-helper, include=FALSE}
create_plot_df <- function(data, snr_label) {
  time <- data$clean$mat[,1]

  # Use the condition with the clearest signal pattern
  best_column <- which.max(apply(data$clean$mat[,-1], 2, function(x) max(x) - min(x)))
  clean_signal <- data$clean$mat[, best_column + 1]
  noisy_signal <- data$noisy[, best_column + 1]

  data.frame(
    Time = time,
    Clean = clean_signal,
    Noisy = noisy_signal,
    SNR = snr_label
  )
}
```

```{r prepare-snr-plot-data}
plot_df <- rbind(
  create_plot_df(data_snr_1.0, "SNR = 1.0"),
  create_plot_df(data_snr_0.5, "SNR = 0.5"),
  create_plot_df(data_snr_0.2, "SNR = 0.2")
)

plot_df_long <- tidyr::pivot_longer(plot_df,
                                    cols = -c(Time, SNR),
                                    names_to = "Type",
                                    values_to = "Signal")
```

The overlay plot shows the clean signal (red) against the noisy measurement (blue) at each SNR level. As SNR decreases, the noise increasingly obscures the true signal.

```{r plot-snr-overlay, fig.height=9}
ggplot(plot_df_long, aes(x = Time, y = Signal, color = Type)) +
  geom_line() +
  facet_wrap(~SNR, ncol = 1) +
  theme_minimal() +
  labs(title = "Clean vs. Noisy Signals at Different SNR Levels",
       subtitle = "Single condition shown for clarity",
       x = "Time (seconds)",
       y = "Signal",
       color = "Signal Type") +
  scale_color_manual(values = c("Clean" = "darkred", "Noisy" = "steelblue")) +
  theme(legend.position = "top")
```

```{r plot-faceted-helper, include=FALSE}
plot_faceted <- function(df, snr_value) {
  df_subset <- subset(df, SNR == snr_value)

  noise <- df_subset$Noisy - df_subset$Clean

  decomp_df <- data.frame(
    Time = rep(df_subset$Time, 3),
    Component = factor(rep(c("Signal+Noise", "Signal", "Noise"), each = length(df_subset$Time)),
                      levels = c("Signal+Noise", "Signal", "Noise")),
    Value = c(df_subset$Noisy, df_subset$Clean, noise)
  )

  ggplot(decomp_df, aes(x = Time, y = Value, color = Component)) +
    geom_line() +
    facet_wrap(~Component, ncol = 1, scales = "free_y") +
    theme_minimal() +
    labs(title = paste("Signal Decomposition at", snr_value),
         x = "Time (seconds)",
         y = "Amplitude") +
    scale_color_manual(values = c("Signal+Noise" = "steelblue",
                                 "Signal" = "darkred",
                                 "Noise" = "darkgreen")) +
    theme(legend.position = "none")
}
```

The decomposition plots below separate each SNR level into its signal, noise, and combined components. At SNR = 1.0 the signal is clearly visible; by SNR = 0.2 the noise dominates.

```{r plot-decomposition-snr1, fig.height=7}
plot_faceted(plot_df, "SNR = 1.0")
```

```{r plot-decomposition-snr05, fig.height=7}
plot_faceted(plot_df, "SNR = 0.5")
```

```{r plot-decomposition-snr02, fig.height=7}
plot_faceted(plot_df, "SNR = 0.2")
```

```{r compute-snr-stats}
snr_stats_list <- list()
for (snr_val in unique(plot_df_long$SNR)) {
  for (type_val in unique(plot_df_long$Type)) {
    subset_data <- plot_df_long[plot_df_long$SNR == snr_val & plot_df_long$Type == type_val, ]
    snr_stats_list[[length(snr_stats_list) + 1]] <- data.frame(
      SNR = snr_val,
      Type = type_val,
      Mean = mean(subset_data$Signal),
      SD = sd(subset_data$Signal),
      Range = max(subset_data$Signal) - min(subset_data$Signal)
    )
  }
}
snr_stats <- do.call(rbind, snr_stats_list)
```

```{r display-snr-table}
knitr::kable(snr_stats, caption = "Statistics of clean and noisy signals at different SNR levels")
```

This visualization shows how different SNR levels affect the fMRI time series. The lower the SNR, the more the noise dominates the signal. For each SNR level, we show:

1. **The original clean signal** (red line): The true underlying BOLD response
2. **The noisy signal** (blue line): What would actually be measured by the scanner
3. **Signal decomposition**: Visualization of how the signal, noise, and combined signal relate at each SNR level

With SNR = 1.0, the signal pattern remains clearly visible despite the noise. At SNR = 0.5, some features of the signal are obscured, while at SNR = 0.2, the noise substantially masks the underlying signal, making accurate detection more challenging without proper statistical methods.

The function returns:
- **`clean`**: The simulated signals without noise
- **`noisy`**: The signals with added noise
- **`noise`**: The simulated noise component
- **`onsets`**: Trial onset times
- **`conditions`**: Condition labels for each trial

## Simulating Matrix Time Series with Column-Specific Variation

The most advanced simulation function, `simulate_fmri_matrix`, generates multiple time series (columns) that share the same event timing but have column-specific variations in amplitudes and durations. This is particularly useful for simulating multiple voxels or regions with related but slightly different response profiles.

```{r simulate-fmri-matrix}
sim_matrix <- simulate_fmri_matrix(
  n = 5,                  # 5 voxels/regions
  total_time = 200,       # 200 seconds of scan time
  TR = 2,                 # TR = 2 seconds
  n_events = 10,          # 10 events
  amplitudes = 1,         # Base amplitude = 1
  amplitude_sd = 0.3,     # Amplitude variability
  durations = 2,          # Base duration = 2 seconds
  duration_sd = 0.5,      # Duration variability
  noise_type = "ar1",     # AR(1) noise
  noise_sd = 0.5          # Noise standard deviation
)
```

```{r extract-matrix-data}
ts_data <- sim_matrix$time_series
matrix_data <- ts_data$datamat

time_points <- seq(0, by = 2, length.out = nrow(matrix_data))
plot_data <- data.frame(Time = time_points)

for(i in 1:ncol(matrix_data)) {
  plot_data[[paste0("Voxel", i)]] <- matrix_data[, i]
}

plot_data_long <- tidyr::pivot_longer(
  plot_data,
  cols = starts_with("Voxel"),
  names_to = "Voxel",
  values_to = "Signal"
)
```

All five simulated voxels share the same event timing, but their response amplitudes and durations vary independently.

```{r plot-matrix-timeseries}
ggplot(plot_data_long, aes(x = Time, y = Signal, color = Voxel)) +
  geom_line(alpha = 0.8) +
  theme_minimal() +
  labs(title = "Multiple Simulated Time Series with Column-Specific Variation",
       x = "Time (seconds)",
       y = "Signal",
       color = "Voxel") +
  scale_color_brewer(palette = "Set2")
```

```{r reshape-amp-dur-data}
amp_df <- as.data.frame(sim_matrix$ampmat)
colnames(amp_df) <- paste0("Voxel", 1:ncol(amp_df))
amp_df$Event <- 1:nrow(amp_df)

dur_df <- as.data.frame(sim_matrix$durmat)
colnames(dur_df) <- paste0("Voxel", 1:ncol(dur_df))
dur_df$Event <- 1:nrow(dur_df)

amp_long <- tidyr::pivot_longer(
  amp_df,
  cols = starts_with("Voxel"),
  names_to = "Voxel",
  values_to = "Amplitude"
)

dur_long <- tidyr::pivot_longer(
  dur_df,
  cols = starts_with("Voxel"),
  names_to = "Voxel",
  values_to = "Duration"
)
```

The amplitude plot shows how each voxel's response magnitude varies from event to event around the base amplitude of 1.

```{r plot-amplitude-variation}
ggplot(amp_long, aes(x = Event, y = Amplitude, color = Voxel, group = Voxel)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Amplitude Variation Across Events and Voxels",
       x = "Event Number",
       y = "Amplitude",
       color = "Voxel") +
  scale_color_brewer(palette = "Set2")
```

Duration variation follows the same pattern -- each voxel draws its event durations independently around the base of 2 seconds.

```{r plot-duration-variation}
ggplot(dur_long, aes(x = Event, y = Duration, color = Voxel, group = Voxel)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Duration Variation Across Events and Voxels",
       x = "Event Number",
       y = "Duration (seconds)",
       color = "Voxel") +
  scale_color_brewer(palette = "Set2")
```

This function is particularly powerful for simulating multiple related time series with:
- **Shared event timing** but individual variation in:
  - **Amplitude** (per event, per column)
  - **Duration** (per event, per column)
- **Independent noise** generation for each column
- **Complex output** including:
  - Time series matrix
  - Amplitude and duration matrices
  - HRF and noise parameter information

## Summary and Comparison

The four simulation functions in `fmrireg` serve different purposes and offer increasing levels of complexity:

1. **`simulate_bold_signal`**: Generate clean BOLD signals for multiple conditions
2. **`simulate_noise_vector`**: Create realistic fMRI noise with temporal structure
3. **`simulate_simple_dataset`**: Combine signal and noise with a specific SNR
4. **`simulate_fmri_matrix`**: Create multiple time series with trial-by-trial, column-specific parameter variation

Choose the appropriate function based on your simulation needs:
- For basic signal generation: use `simulate_bold_signal`
- For realistic noise: use `simulate_noise_vector`
- For a complete dataset with controlled SNR: use `simulate_simple_dataset`
- For simulating multiple voxels/regions with shared timing but response variation: use `simulate_fmri_matrix`

These functions provide a powerful toolkit for method development, validation, or teaching fMRI analysis concepts through realistic simulations.

## Next

- 04 fMRI Linear Model (GLM)
