---
title: fMRI Linear Model (GLM)
author: Bradley R. Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{fMRI Linear Model (GLM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=TRUE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
load_vignette_package <- function(package) {
  withCallingHandlers(
    suppressPackageStartupMessages(
      library(package, character.only = TRUE)
    ),
    warning = function(w) {
      if (grepl("^package '.+' was built under R version", conditionMessage(w))) {
        message("Qualified host note: ", conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    }
  )
}
invisible(lapply(
  c("fmrireg", "fmrihrf", "ggplot2", "dplyr", "tidyr", "knitr"),
  load_vignette_package
))

set.seed(42)  # For reproducible examples

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Introduction to fMRI Linear Models

Statistical analysis of fMRI data typically involves fitting a linear model to each voxel's time series. This approach, often called the General Linear Model (GLM), estimates how much each experimental condition contributes to the observed signal. The `fmrireg` package provides a flexible framework for:

1. Modeling the hemodynamic response to experimental stimuli
2. Accounting for baseline trends and noise
3. Estimating condition-specific effects
4. Computing contrasts between conditions
5. Testing statistical hypotheses about brain activity

Thread usage in internal routines can be adjusted globally. Set `options(fmrireg.num_threads = <n>)` or the environment variable `FMRIREG_NUM_THREADS` before loading the package to control how many threads RcppParallel uses.

This vignette demonstrates how to conduct a complete linear model analysis using the `fmrireg` package, from data simulation to statistical inference.

## Simulating a Dataset for Analysis

First, let's create a controlled fMRI teaching dataset with known parameters. We'll simulate a simple experiment with two conditions that have different amplitudes.

```{r create-event-table, message=FALSE, fig.width=7, fig.height=5, fig.alt="Event onsets for condition 1 and condition 2 across the 400-second simulated run."}
# Create an experimental design with two conditions
# Condition 1: 10 events with amplitude 1.0
# Condition 2: 10 events with amplitude 2.0

# Define parameters
TR <- 2                  # Repetition time (2 seconds)
run_length <- 200        # 200 timepoints per run = 400 seconds
nruns <- 1               # Number of runs

# Create an event table
run_id <- rep(1, 20)
condition <- factor(rep(c("condition1", "condition2"), each = 10))
onset_times <- sort(runif(20, min = 10, max = 380))  # Random onsets between 10s and 380s

event_table <- data.frame(
  run = run_id,
  onset = onset_times,
  condition = condition
)

# Display the experiment design
kable(head(event_table), caption = "First few rows of the experimental design")

# Create a sampling frame
sframe <- sampling_frame(blocklens = run_length, TR = TR)

# Plot event onsets without drawing a dense row of zero-valued samples.
ggplot(event_table, aes(x = onset, y = 1, color = condition)) +
  geom_segment(aes(xend = onset, y = 0, yend = 1), linewidth = 1) +
  geom_point(size = 3) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top",
        text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16)) +
  labs(title = "Experimental Design with Event Onsets",
       x = "Time (seconds)",
       y = NULL) +
  scale_y_continuous(breaks = 1, labels = "onset", limits = c(0, 1.1)) +
  scale_color_brewer(palette = "Set1")
```

Now that we have our experimental design, let's simulate the fMRI time series. We'll create signals for each condition with different amplitudes, add noise, and combine them into a dataset.

```{r simulate-signals, message=FALSE, fig.width=7, fig.height=8, fig.alt="Faceted time series showing both condition responses, their combined signal, structured noise, and the observed signal."}
# Simulate the true BOLD signals for each condition
# First, convert our events to global indices
global_onsets <- global_onsets(sframe, event_table$onset, blockids(sframe)[event_table$run])

# Create regressors for each condition
condition1_indices <- which(event_table$condition == "condition1")
condition2_indices <- which(event_table$condition == "condition2")

reg1 <- regressor(global_onsets[condition1_indices], hrf = fmrihrf::HRF_SPMG1, amplitude = 1.0)
reg2 <- regressor(global_onsets[condition2_indices], hrf = fmrihrf::HRF_SPMG1, amplitude = 2.0)

# Sample time points
time_points <- samples(sframe, global = TRUE)

# Evaluate regressors at each time point
signal1 <- evaluate(reg1, time_points)
signal2 <- evaluate(reg2, time_points)

# Combine signals (this is the "true" signal without noise)
true_signal <- signal1 + signal2

# Create noise with temporal autocorrelation and drift
noise <- simulate_noise_vector(
  n = length(time_points),
  TR = TR,
  ar = c(0.7),         # Stronger AR(1) coefficient to accentuate autocorrelation
  ma = numeric(0),     # Pure AR structure keeps the example simple
  drift_freq = 1/128,  # Slow drift
  drift_amplitude = 1, # Moderate drift amplitude
  physio = TRUE,       # Include physiological noise
  sd = 0.5             # Noise standard deviation
)

# Create the observed signal by adding noise
observed_signal <- true_signal + noise

# Create a data frame for visualization
signal_df <- data.frame(
  time = time_points,
  true_signal = true_signal,
  noise = noise,
  observed_signal = observed_signal,
  condition1 = signal1,
  condition2 = signal2
)

# Create a matrix dataset for the model fitting. A shared noise realization
# keeps the example focused on the known change in signal amplitude.
voxel_scales <- c(1.0, 0.8, 0.6)
simulated_data <- vapply(
  voxel_scales,
  function(scale) scale * true_signal + noise,
  numeric(length(true_signal))
)
dataset <- fmridataset::matrix_dataset(
  datamat = simulated_data,
  TR = TR,
  run_length = run_length,
  event_table = event_table
)

# Visualize the signals
signal_long <- signal_df %>%
  select(time, condition1, condition2, true_signal, noise, observed_signal) %>%
  pivot_longer(cols = -time, names_to = "component", values_to = "signal")

# Set the factor levels for better plotting order
signal_long$component <- factor(signal_long$component, 
                               levels = c("condition1", "condition2", "true_signal", "noise", "observed_signal"))

# Plot signals
ggplot(signal_long, aes(x = time, y = signal, color = component)) +
  geom_line() +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14)) +
  labs(title = "Simulated fMRI Time Series Components",
       x = "Time (seconds)",
       y = "Signal Amplitude")
```

Our simulated dataset now contains:

1. **Condition-specific signals** with known amplitudes (1.0 and 2.0)
2. **Structured teaching noise** with temporal autocorrelation, drift, and physiological components
3. **Three teaching voxels** with known signal scales and a shared noise realization
4. **A complete event table** with condition labels and onset times

## Fitting a Linear Model

Now we can fit a linear model to our simulated data using the `fmri_lm` function. We need to specify:

1. The formula describing the experimental effects
2. The block structure of the data
3. The dataset

```{r fit-linear-model, message=FALSE, fig.width=7, fig.height=6}
# Fit the central model with the AR(1) structure used by the generator.
model <- fmri_lm(
  formula = onset ~ hrf(condition),  # Model experimental effects
  block = ~ run,                     # Block structure
  dataset = dataset,                 # Our simulated dataset
  control = fmri_lm_control(noise = noise_spec("ar1", iter_gls = 2L)),
  compute = compute_spec(voxel_chunks = 1L)
)

# Print a summary of the model
model
```

The onset plot above shows the schedule; the matrix actually fitted by the GLM
also includes the sampled HRF regressors and baseline columns. Inspect that
matrix before interpreting coefficients, especially when designs are larger or
contain nuisance regressors.

```{r inspect-design-matrix, fig.width=8, fig.height=4.5, fig.alt="Column-scaled fitted design matrix across the 200 samples, including both condition regressors and baseline terms."}
fitted_design <- as.matrix(design_matrix(model$model))
design_denom <- pmax(apply(abs(fitted_design), 2, max), .Machine$double.eps)
design_scaled <- sweep(fitted_design, 2, design_denom, "/")
design_labels <- colnames(design_scaled)
design_labels <- sub("^condition_condition\\.", "condition: ", design_labels)
design_labels <- gsub("_", " ", design_labels, fixed = TRUE)
colnames(design_scaled) <- design_labels

design_plot_data <- as.data.frame(design_scaled, check.names = FALSE) %>%
  mutate(sample = row_number()) %>%
  pivot_longer(-sample, names_to = "regressor", values_to = "scaled_value") %>%
  mutate(regressor = factor(regressor, levels = rev(design_labels)))

ggplot(design_plot_data, aes(sample, regressor, fill = scaled_value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    limits = c(-1, 1), name = "Column-scaled\nvalue"
  ) +
  labs(title = "Fitted GLM design matrix", x = "Sample", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

### Accounting for Temporal Autocorrelation

The simulated noise contains AR(1) structure, so the central model above uses
AR(1) prewhitening with `noise_spec("ar1")`. By default, the built-in shared
path averages the voxel residuals at each time point, estimates one AR
coefficient vector from that mean residual series for each run, whitens the
data and design matrix, and refits the GLM. A spatially coherent residual
component can therefore make this shared estimate larger than the median of
separately estimated voxel coefficients.

To estimate a separate AR model for every voxel, request the supported
runwise-meta path explicitly:

```r
voxelwise_control <- fmri_lm_control(
  estimation = estimation_spec("runwise_meta"),
  noise = noise_spec("ar1", voxelwise = TRUE)
)
```

Built-in voxelwise estimation currently supports AR-only models with run
pooling, one GLS iteration, no censoring, and no volume weighting or
soft-subspace projection. Robust fitting is supported, but robust AR
re-estimation is not. Unsupported combinations fail at the fitting boundary
instead of silently using the shared estimator.

Prewhitening changes both the coefficient covariance and the effective
information in the time series. It is therefore important to inspect the
result rather than assume that correction must make standard errors smaller.

```{r fit-ar1-model, message=FALSE}
model_ols <- fmri_lm(
  formula = onset ~ hrf(condition),
  block   = ~ run,
  dataset = dataset,
  compute = compute_spec(voxel_chunks = 1L)
)
model_ar1 <- model

se_ols <- standard_error(model_ols)
se_ar1 <- standard_error(model_ar1)
se_comparison <- data.frame(
  model = c("OLS", "AR(1) prewhitened"),
  median_event_se = c(median(as.matrix(se_ols)), median(as.matrix(se_ar1)))
)
kable(se_comparison, digits = 4, caption = "Median standard error across event coefficients")

# Inspect the estimated AR coefficient (shared across voxels in this example)
model_ar1$ar_coef[[1]]
```

```{r validate-ar1-result, include=FALSE}
stopifnot(
  all(is.finite(se_comparison$median_event_se)),
  se_comparison$median_event_se[2] > se_comparison$median_event_se[1],
  any(abs(model_ar1$ar_coef[[1]]) > 0.05)
)
```

Here the AR(1) fit estimates non-zero serial dependence and produces larger
standard errors than plain OLS. The practical lesson is not that prewhitening
always increases uncertainty, but that OLS can be overconfident when residuals
are serially correlated.

The `noise_spec()` constructor also accepts `struct = "arp"` for higher-order autoregressive
models. This setting models AR coefficients only (no moving-average terms).

### Handling Outliers with Row-Wise Robust Fitting

Real fMRI runs sometimes contain entire time points corrupted by motion or scanner artifacts. The `fmri_lm`
function can mitigate their impact by enabling row-wise robust weighting: an Iteratively Reweighted Least
Squares loop down-weights frames with large residuals. Select the weighting
function and its iteration limit once with `robust_spec()`.

```{r fit-robust-model, message=FALSE}
model_robust <- fmri_lm(
  formula = onset ~ hrf(condition),
  block   = ~ run,
  dataset = dataset,
  control = fmri_lm_control(
    robust = robust_spec("huber", max_iter = 2L)
  ),
  compute = compute_spec(voxel_chunks = 1L)
)

se_robust <- standard_error(model_robust)
robust_comparison <- data.frame(
  model = c("OLS", "Huber"),
  median_event_se = c(median(as.matrix(se_ols)), median(as.matrix(se_robust)))
)
kable(robust_comparison, digits = 4, caption = "Median event-coefficient standard error")
```

```{r validate-robust-result, include=FALSE}
stopifnot(
  all(is.finite(robust_comparison$median_event_se)),
  robust_comparison$median_event_se[2] < robust_comparison$median_event_se[1]
)
```

Robust fitting guards against frames with multivariate residual excursions; it
does not specifically target a spike confined to one voxel. In this realization
Huber fitting modestly decreases the reported standard errors. That direction
is data-dependent, not a general guarantee. Robust p-values use a robust
residual scale and should be interpreted as approximate.



## Extracting Model Results

Let's extract and visualize the results from our linear model.

### 1. Coefficient Estimates

```{r extract-coefficients, message=FALSE, fig.width=7, fig.height=5}
# Extract coefficient estimates
beta_estimates <- coef(model)
kable(beta_estimates, caption = "Coefficient estimates for each condition and voxel")
```



```{r clean-condition-label, include=FALSE}
# Helper to clean condition column labels for readability
clean_condition_label <- function(x) {
  x <- gsub("^conditioncondition_condition\\.", "", x)
  x <- gsub("^condition_condition\\.", "", x)
  gsub("\\.", " ", x)
}
```

```{r plot-coefficients, message=FALSE, fig.width=7, fig.height=7, fig.alt="Condition 1 and condition 2 coefficient estimates for each of the three simulated voxels."}
# Reshape for plotting (works with both approaches)
beta_long <- as.data.frame(t(beta_estimates)) %>%
  mutate(voxel_index = row_number()) %>%
  pivot_longer(cols = -voxel_index, names_to = "condition", values_to = "estimate") %>%
  mutate(
    condition = clean_condition_label(condition),
    voxel = paste0("voxel", voxel_index)
  )

# Plot coefficient estimates
ggplot(beta_long, aes(x = condition, y = estimate, fill = condition)) +
  geom_bar(stat = "identity") +
  facet_wrap(~voxel, ncol = 3, scales = "free_y") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top",
        text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Estimated Coefficients by Condition and Voxel",
       x = "Condition",
       y = "Coefficient Estimate")
```

The generator assigns Condition2 twice the underlying amplitude of Condition1
and scales the three voxels by 1.0, 0.8, and 0.6. Finite noisy estimates need
not preserve either ratio exactly. The plot should therefore be read together
with the uncertainty table below; the scientifically relevant checks are the
direction and uncertainty of the planned contrast, not visual agreement with
an exact 2:1 ratio.

### 2. T-Statistics and P-Values

```{r extract-statistics, message=FALSE, fig.width=7, fig.height=5}
estimate_stats <- tidy(model, type = "estimates") %>%
  filter(grepl("condition", term)) %>%
  mutate(condition = clean_condition_label(term),
         voxel = paste0("voxel", voxel)) %>%
  select(voxel, condition, estimate, std_error, statistic, p_value)

estimate_stats_display <- estimate_stats %>%
  mutate(p_value = format.pval(p_value, digits = 4, eps = 1e-4))

kable(estimate_stats_display, digits = 4, row.names = FALSE,
      caption = "Coefficient estimates with associated statistics by voxel and condition")
```

```{r validate-estimate-statistics, include=FALSE}
stopifnot(
  nrow(estimate_stats) == 6L,
  all(is.finite(as.matrix(estimate_stats[c("estimate", "std_error", "statistic", "p_value")]))),
  any(estimate_stats$p_value < 0.05),
  any(estimate_stats$p_value >= 0.05)
)
```

The t-statistics quantify the reliability of the estimated effects. Higher
absolute t-values indicate more precise estimates relative to their standard
errors. Here `r sum(estimate_stats$p_value < 0.05)` of the six condition-by-voxel
tests are below 0.05. The weakest simulated signal is correspondingly the least
certain; this is why the table reports both effect size and uncertainty rather
than reducing the result to a binary label.

### 3. Contrasts Between Conditions

A key advantage of the GLM approach is the ability to directly compare conditions using contrasts.

```{r define-and-fit-contrast, message=FALSE, fig.width=7, fig.height=5}
# Define a contrast specification for comparing condition2 vs condition1
con_spec <- pair_contrast(~ condition == "condition2", ~ condition == "condition1", name = "cond2_minus_cond1")

# Define a contrast model using the specified contrast in hrf()
contrast_model <- fmri_lm(
  formula = onset ~ hrf(condition, contrasts = con_spec),
  block = ~ run,
  dataset = dataset,
  control = fmri_lm_control(noise = noise_spec("ar1", iter_gls = 2L)),
  compute = compute_spec(voxel_chunks = 1L)
)

# Extract contrast results using tidy helper
contrast_results <- tidy(contrast_model, type = "contrasts") %>%
  filter(term == "cond2_minus_cond1") %>%
  mutate(
    voxel = paste0("voxel", voxel),
    significant = p_value < 0.05
  )
```

```{r validate-contrast-statistics, include=FALSE}
stopifnot(
  nrow(contrast_results) == 3L,
  all(is.finite(contrast_results$estimate)),
  all(contrast_results$estimate > 0),
  all(contrast_results$significant)
)
```

```{r plot-contrast-results, message=FALSE, fig.width=7, fig.height=6, fig.alt="Positive condition 2 minus condition 1 contrast estimates for all three simulated voxels."}
# Display contrast results
contrast_results_display <- contrast_results %>%
  mutate(p_value = format.pval(p_value, digits = 4, eps = 1e-4))
kable(contrast_results_display,
      caption = "Contrast results: condition2 - condition1",
      digits = 4, row.names = FALSE)

# Visualize the contrast
ggplot(contrast_results, aes(x = as.factor(voxel), y = estimate, fill = significant)) +
  geom_bar(stat = "identity") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 15)) +
  labs(title = "Condition2 - Condition1 Contrast",
       subtitle = "Positive values indicate stronger activation for Condition2",
       x = "Voxel",
       y = "Contrast Estimate") +
  scale_fill_manual(values = c("FALSE" = "gray", "TRUE" = "red"))
```

The contrast results show that Condition2 consistently elicits significantly stronger activation than Condition1 across all voxels, which matches our simulation parameters (where Condition2 had twice the amplitude of Condition1).

## Fitted HRF Curves

Another useful visualization is the fitted hemodynamic response for each condition. This shows the estimated BOLD response over time.

```{r extract-fitted-hrf, message=FALSE, fig.width=7, fig.height=8, fig.alt="Fitted hemodynamic response curves for both conditions in each simulated voxel."}
hrf_long <- dplyr::bind_rows(lapply(
  seq_len(ncol(get_data_matrix(dataset))),
  function(voxel_index) tidy_fitted_hrf(
    model,
    sample_at = seq(0, 20, by = 0.5),
    term = "condition",
    term_match = "contains",
    voxel = voxel_index
  )
))

# Plot the fitted HRF curves for each condition and voxel
ggplot(hrf_long, aes(x = time, y = estimate, color = condition)) +
  geom_line() +
  facet_grid(voxel ~ term) +
  theme_minimal(base_size = 14) +
  theme(text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14)) +
  labs(title = "Fitted Hemodynamic Response Functions",
       x = "Time (seconds)",
       y = "BOLD Response",
       color = "Condition")
```

```{r validate-fitted-hrf, include=FALSE}
stopifnot(
  nrow(hrf_long) > 0L,
  all(is.finite(hrf_long$estimate)),
  length(unique(hrf_long$voxel)) == 3L
)
```

The fitted HRF curves show the temporal profile of the BOLD response for each condition. We can observe:

1. The peak response around 5-6 seconds post-stimulus
2. The stronger response for Condition2 compared to Condition1
3. The decreasing response amplitude across voxels

## Estimating the HRF Shape from Data

The curves above inherit the canonical HRF chosen for the GLM. When the shape
itself is the target, `estimate_hrf()` instead fits smooth condition-level
curves without assuming a canonical response. The example below uses white
noise because its reported intervals currently assume independent,
homoscedastic time-point errors; they are not autocorrelation-robust.

To make success visible, we simulate two known responses: Condition A uses the
canonical SPM shape, while Condition B is 1.5 seconds later and slightly taller.
Two ROIs have different amplitudes, and a linear drift is removed as a nuisance
term. The simulation mechanics are hidden so the teaching path stays focused on
the estimator.

```{r simulate-empirical-hrf, include=FALSE}
set.seed(731)
hrf_n <- 300L
hrf_onsets <- 8 + c(0, cumsum(sample(c(6, 7, 9, 10), 31L, replace = TRUE)))
hrf_events <- data.frame(
  onset = hrf_onsets,
  condition = factor(sample(rep(c("A", "B"), 16L))),
  run = 1L
)
hrf_sframe <- sampling_frame(blocklens = hrf_n, TR = 1)
hrf_acquisition_time <- samples(hrf_sframe, global = TRUE)
hrf_truth_a <- function(time) as.numeric(fmrihrf::HRF_SPMG1(time))
hrf_truth_b <- function(time) {
  ifelse(
    time >= 1.5,
    1.25 * as.numeric(fmrihrf::HRF_SPMG1(time - 1.5)),
    0
  )
}
hrf_signal_a <- evaluate(
  regressor(
    hrf_events$onset[hrf_events$condition == "A"],
    hrf = hrf_truth_a,
    span = 24
  ),
  hrf_acquisition_time
)
hrf_signal_b <- evaluate(
  regressor(
    hrf_events$onset[hrf_events$condition == "B"],
    hrf = hrf_truth_b,
    span = 24
  ),
  hrf_acquisition_time
)
hrf_amplitudes <- rbind(A = c(1, 0.65), B = c(0.8, 1.1))
hrf_drift <- outer(seq(-0.15, 0.15, length.out = hrf_n), rep(1, 2L))
hrf_response <- hrf_signal_a %o% hrf_amplitudes["A", ] +
  hrf_signal_b %o% hrf_amplitudes["B", ] + hrf_drift +
  matrix(rnorm(hrf_n * 2L, sd = 0.025), nrow = hrf_n)
colnames(hrf_response) <- c("visual_ROI", "motor_ROI")
hrf_dataset <- matrix_dataset(
  hrf_response,
  TR = 1,
  run_length = hrf_n,
  event_table = hrf_events
)
hrf_baseline <- baseline_model(
  "poly", degree = 1, sframe = hrf_dataset$sampling_frame
)
```

The public call is short. GCV chooses one smoothing strength shared across the
ROIs, preventing high-variance columns from dominating the choice.

```{r estimate-empirical-hrf}
empirical_hrf <- estimate_hrf(
  onset ~ hrf(condition),
  block = ~run,
  dataset = hrf_dataset,
  basemod = hrf_baseline,
  rsam = seq(0, 24, by = 0.5),
  k = 8,
  lambda = "gcv"
)
empirical_hrf
```

```{r prepare-empirical-hrf-results, include=FALSE}
hrf_recovery <- tidy(empirical_hrf) %>%
  mutate(
    condition = sub("^.*[.]", "", condition),
    amplitude = ifelse(
      condition == "A",
      hrf_amplitudes["A", match(voxel, colnames(hrf_response))],
      hrf_amplitudes["B", match(voxel, colnames(hrf_response))]
    ),
    truth = ifelse(
      condition == "A",
      hrf_truth_a(time) * amplitude,
      hrf_truth_b(time) * amplitude
    )
  )

hrf_recovery_metrics <- hrf_recovery %>%
  group_by(condition, voxel) %>%
  summarise(
    correlation = cor(estimate, truth),
    normalized_rmse = sqrt(mean((estimate - truth)^2)) / max(abs(truth)),
    .groups = "drop"
  )

hrf_peak_times <- hrf_recovery %>%
  group_by(condition, voxel) %>%
  slice_max(estimate, n = 1L, with_ties = FALSE) %>%
  ungroup() %>%
  select(condition, voxel, peak_time = time) %>%
  pivot_wider(names_from = condition, values_from = peak_time)

stopifnot(
  all(is.finite(empirical_hrf$estimate)),
  all(is.finite(empirical_hrf$std.error)),
  empirical_hrf$lambda %in% empirical_hrf$gcv$lambda,
  all(is.finite(empirical_hrf$gcv$score)),
  all(hrf_recovery_metrics$correlation > 0.97),
  all(hrf_recovery_metrics$normalized_rmse < 0.08),
  all(hrf_peak_times$B - hrf_peak_times$A >= 1.5)
)
```

```{r empirical-hrf-metrics, echo=FALSE}
kable(
  hrf_recovery_metrics,
  digits = 3,
  row.names = FALSE,
  caption = "Recovery against the known synthetic HRFs"
)
```

```{r plot-empirical-hrf, echo=FALSE, fig.width=8, fig.height=6, fig.alt="Known and estimated hemodynamic response curves agree closely in two synthetic ROIs; Condition B peaks later than Condition A, and shaded confidence intervals surround each estimate."}
ggplot(
  hrf_recovery,
  aes(x = time, color = condition, fill = condition)
) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.14, color = NA) +
  geom_line(aes(y = estimate, linetype = "Estimated"), linewidth = 0.9) +
  geom_line(
    aes(y = truth, linetype = "Known truth"),
    color = "#333333",
    linewidth = 0.75
  ) +
  facet_grid(voxel ~ condition) +
  scale_linetype_manual(values = c("Estimated" = "solid", "Known truth" = "dashed")) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top") +
  labs(
    title = "Flexible HRF estimation recovers two response shapes",
    subtitle = "Shading is the model-based 95% confidence interval",
    x = "Seconds after onset",
    y = "BOLD response",
    color = "Condition",
    fill = "Condition",
    linetype = NULL
  )
```

Across the four condition-by-ROI curves, correlation with the known response is
between `r sprintf("%.3f", min(hrf_recovery_metrics$correlation))` and
`r sprintf("%.3f", max(hrf_recovery_metrics$correlation))`; the largest
normalized RMSE is
`r sprintf("%.3f", max(hrf_recovery_metrics$normalized_rmse))`. More
importantly, the estimates recover the later Condition B peak in both ROIs.
These are executable checks for this controlled example, not an accuracy
guarantee for arbitrary event schedules or real fMRI noise.

The complete GCV path remains available for diagnosis rather than returning
only the winning value. Here the minimum is interior to the candidate grid.

```{r plot-empirical-hrf-gcv, echo=FALSE, fig.width=7, fig.height=4.5, fig.alt="Scale-normalized generalized cross-validation score across smoothing strengths, with the selected minimum marked by a vertical line."}
hrf_positive_lambda <- min(empirical_hrf$gcv$lambda[empirical_hrf$gcv$lambda > 0])
hrf_gcv_plot <- empirical_hrf$gcv %>%
  mutate(lambda_axis = ifelse(lambda == 0, hrf_positive_lambda / 10, lambda))
hrf_selected_axis <- if (
  empirical_hrf$lambda == 0
) hrf_positive_lambda / 10 else empirical_hrf$lambda

ggplot(hrf_gcv_plot, aes(x = lambda_axis, y = score)) +
  geom_line(color = "#7F0000", linewidth = 0.8) +
  geom_point(color = "#B2182B", size = 2) +
  geom_vline(xintercept = hrf_selected_axis, linetype = "dashed") +
  scale_x_log10() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Shared smoothing selected by GCV",
    subtitle = "The zero-penalty candidate is shown one step left of the positive grid",
    x = "Smoothing strength (log scale)",
    y = "Scale-normalized GCV score"
  )
```


## Comparing Models with Different HRF Bases

The choice of hemodynamic response function can impact model fit. Let's compare different HRF options.

```{r fit-comparison-models, message=FALSE}
# Fit models with different HRF bases
model_canonical <- fmri_lm(
  formula = onset ~ hrf(condition, basis = "spmg1"),
  block = ~ run,
  dataset = dataset,
  control = fmri_lm_control(noise = noise_spec("ar1", iter_gls = 2L)),
  compute = compute_spec(voxel_chunks = 1L)
)

model_gaussian <- fmri_lm(
  formula = onset ~ hrf(condition, basis = "gaussian"),
  block = ~ run,
  dataset = dataset,
  control = fmri_lm_control(noise = noise_spec("ar1", iter_gls = 2L)),
  compute = compute_spec(voxel_chunks = 1L)
)

model_bspline <- fmri_lm(
  formula = onset ~ hrf(condition, basis = "bspline", nbasis = 5),
  block = ~ run,
  dataset = dataset,
  control = fmri_lm_control(noise = noise_spec("ar1", iter_gls = 2L)),
  compute = compute_spec(voxel_chunks = 1L)
)
```

```{r extract-model-stats-fn, include=FALSE}
# Function to extract model fit statistics
extract_model_stats <- function(model, model_name, dataset) {
  # Get observed data
  observed_data <- get_data_matrix(dataset)

  # Get design matrix (ensure it's a numeric matrix)
  design_mat <- as.matrix(design_matrix(model$model))

  # Get estimated coefficients (include baseline)
  # Get all coefficients including baseline
  betas <- as.matrix(coef(model, include_baseline = TRUE))

  # Calculate fitted values
  fitted_vals <- design_mat %*% t(betas)

  # Calculate residuals
  resids <- observed_data - fitted_vals

  # Calculate sum of squared residuals
  ssr <- colSums(resids^2)

  # Calculate R-squared
  tss <- apply(observed_data, 2, function(y) sum((y - mean(y))^2))
  r_squared <- 1 - ssr/tss

  # Calculate AIC
  n <- nrow(observed_data)
  p <- ncol(design_mat)
  aic <- n * log(ssr/n) + 2 * p

  # Return results
  data.frame(
    model = model_name,
    voxel = 1:ncol(observed_data),
    r_squared = r_squared,
    aic = aic,
    ssr = ssr
  )
}
```

```{r plot-r-squared, message=FALSE, fig.width=7, fig.height=5, fig.alt="R-squared values for canonical, Gaussian, and B-spline HRF models in each simulated voxel."}
# Extract statistics for each model
stats_canonical <- extract_model_stats(model_canonical, "canonical_spm", dataset)
stats_gaussian <- extract_model_stats(model_gaussian, "gaussian", dataset)
stats_bspline <- extract_model_stats(model_bspline, "bspline_n5", dataset)

# Combine results
model_comparison <- rbind(stats_canonical, stats_gaussian, stats_bspline)

# Display model comparison
kable(model_comparison, caption = "Model comparison statistics", digits = 4)

aic_winners <- model_comparison %>%
  group_by(voxel) %>%
  slice_min(aic, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(voxel, selected_model = model, aic)
kable(aic_winners, caption = "Lowest-AIC model in each voxel", digits = 4)

# Reshape for plotting
model_comparison_long <- model_comparison %>%
  pivot_longer(cols = c(r_squared, aic, ssr), names_to = "metric", values_to = "value")

# Plot comparison (separate plots for different metrics)
ggplot(subset(model_comparison_long, metric == "r_squared"), 
       aes(x = model, y = value, fill = model)) +
  geom_bar(stat = "identity") +
  facet_wrap(~voxel, ncol = 3) +
  theme_minimal(base_size = 14) +
  theme(text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "Model Comparison: R-Squared",
       x = "HRF Model",
       y = "R-Squared")
```

```{r validate-model-comparison, include=FALSE}
stopifnot(
  nrow(aic_winners) == 3L,
  all(is.finite(model_comparison$aic)),
  sum(aic_winners$selected_model == "canonical_spm") == 2L,
  sum(aic_winners$selected_model == "bspline_n5") == 1L
)
```

```{r plot-aic, message=FALSE, fig.width=7, fig.height=7, fig.alt="AIC values for canonical, Gaussian, and B-spline HRF models in each simulated voxel; lower values are preferred."}
ggplot(subset(model_comparison_long, metric == "aic"),
       aes(x = model, y = value, fill = model)) +
  geom_bar(stat = "identity") +
  facet_wrap(~voxel, ncol = 3) +
  theme_minimal(base_size = 14) +
  theme(text = element_text(size = 14),
        axis.title = element_text(size = 15),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "Model Comparison: AIC (Lower is Better)",
       x = "HRF Model",
       y = "AIC")
```

The model comparison shows:

1. **R-Squared**: The proportion of variance explained by each model. Higher values indicate better fit.
2. **AIC (Akaike Information Criterion)**: A measure of model quality that balances goodness of fit with model complexity. Lower values indicate better models.

The canonical SPM model has the lowest AIC in two voxels, consistent with the
HRF used to generate the signal. In the weakest-signal voxel, however, the
B-spline model wins despite its larger penalty. This is a useful caution:
finite noisy data need not select the generating model in every voxel. AIC
balances fit and complexity; it is evidence for a model in a particular sample,
not proof that the selected basis is the true physiological response.

## Summary

This vignette demonstrated the complete workflow for fMRI linear model analysis using the `fmrireg` package:

1. **Creating/simulating a dataset** with controlled signal and structured noise
2. **Fitting linear models** with different HRF options
3. **Extracting and visualizing model coefficients** and statistics
4. **Computing and testing contrasts** between conditions
5. **Comparing model performance** using goodness-of-fit metrics
6. **Inspecting fitted and directly estimated hemodynamic response curves**

The `fmri_lm` workflow shown here supports declared temporal covariance,
alternative HRF bases, coefficient extraction, and planned contrasts. Real-data
analyses additionally require prespecified nuisance regressors, preprocessing,
quality control, and a multiplicity strategy appropriate to the scientific
question.

## From first‑level to group analysis (GDS)

You can export subject‑level results with `write_results()` and then load them for group analysis via `group_data()` (which returns a GDS object backed by fmrigds):

```{r gds-example, eval=FALSE}
# HDF5 written by write_results.fmri_lm
h5_paths <- Sys.glob("derivatives/sub-*/task-*_desc-GLMstatmap_bold.h5")
gd <- group_data(h5_paths, format = "h5", subjects = basename(dirname(h5_paths)))
fit <- fmri_meta(gd, formula = ~ 1 + group, method = "pm")

# NIfTI (beta/SE)
gd_nii <- group_data(list(beta = beta_paths, se = se_paths), format = "nifti",
                     subjects = ids, mask = mask_path)
fit_nii <- fmri_meta(gd_nii, formula = ~ 1 + group, method = "fe")

# Inspect the pipeline with fmrigds
pl <- fmrigds::plan(gd)
pl <- fmrigds::reduce(pl, method = "meta:fe", formula = ~ 1 + group)
fmrigds::explain_plan(pl)
invisible(fmrigds::preview(pl, n = 1, assays = "beta"))
res <- fmrigds::compute(pl)
```

Note: `group_data()` now routes through fmrigds (GDS). Legacy constructors (`group_data_from_h5()`/`_nifti()`/`_csv()`) are preserved for backward compatibility but will be deprecated.

## Next

- `vignette("a_05_contrasts", package = "fmrireg")` — Contrasts and Hypothesis Tests
- `vignette("group_analysis", package = "fmrireg")` — Group Analysis
