---
title: "Representing fMRI Data in fmrireg"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Representing fMRI Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=6, fig.height=4)
library(fmrireg)
library(fmridataset) # Dataset functionality refactored from fmrireg
library(neuroim2) # For creating example NeuroVec/NeuroVol
```

## Introduction: Linking Data and Design

Effective fMRI analysis requires associating the measured brain activity (the imaging data) with crucial metadata, including:

*   **Temporal Structure:** When each scan was acquired (TR) and how scans are grouped into runs.
*   **Spatial Structure:** Which brain locations (voxels) are included in the analysis (mask).
*   **Experimental Design:** Timing and properties of experimental events or conditions.

The `fmrireg` package uses several **dataset objects** to encapsulate this information, providing a consistent input format for modeling functions like `event_model`, `baseline_model`, `fmri_lm`, and `estimate_betas`.

This vignette describes the main dataset classes and how to create them.

## The `sampling_frame`

Before diving into datasets, recall the `sampling_frame` object (introduced in the Overview and detailed in other vignettes). It defines the fundamental temporal structure shared by all dataset types:

*   `blocklens`: A vector specifying the number of scans (time points) in each run.
*   `TR`: The repetition time (time between scans) in seconds.

```{r sampling_frame_recap}
sframe_example <- sampling_frame(blocklens = c(150, 160), TR = 2.0)
print(sframe_example)
```

Dataset objects internally create or utilize a `sampling_frame` based on the provided run lengths and TR.

## Overview of Dataset Classes

`fmrireg` offers different dataset classes depending on how your data is stored:

*   **`fmri_mem_dataset`:** For volumetric fMRI data already loaded into R memory (as `NeuroVec` objects).
*   **`fmri_file_dataset`:** For volumetric fMRI data stored in image files (e.g., NIfTI) on disk.
*   **`matrix_dataset`:** For fMRI data represented as a standard R matrix (time points x voxels/components).
*   **`latent_dataset`:** For dimension-reduced data (e.g., PCA/ICA components), typically requiring the `fmristore` package.

All these inherit from a base `fmri_dataset` class.

## In-Memory Volumetric Data (`fmri_mem_dataset`)

Use this when your fMRI runs are loaded as `neuroim2::NeuroVec` objects in your R session.

**Key Arguments:**

*   `scans`: A *list* of `NeuroVec` objects, one for each run.
*   `mask`: A `neuroim2::NeuroVol` or `neuroim2::LogicalNeuroVol` object representing the brain mask.
*   `TR`: Repetition time (seconds).
*   `run_length` (Optional): Vector of run lengths; if omitted, inferred from the dimensions of the `NeuroVec` objects in `scans`.
*   `event_table` (Optional): A `data.frame` containing experimental design information.

```{r mem_dataset_example}
# Create minimal example data (2 runs)
d <- c(5, 5, 5, 20) # Small dimensions for example
mask_vol <- neuroim2::LogicalNeuroVol(array(TRUE, d[1:3]), neuroim2::NeuroSpace(d[1:3]))

scan1 <- neuroim2::NeuroVec(array(rnorm(prod(d)), d), neuroim2::NeuroSpace(d))
scan2 <- neuroim2::NeuroVec(array(rnorm(prod(d)), d), neuroim2::NeuroSpace(d))

# Example event table
events_df <- data.frame(
  onset = c(5, 15, 5, 15), 
  condition = factor(c("A", "B", "A", "B")),
  run = c(1, 1, 2, 2)
)

# Create the dataset object
mem_dset <- fmri_mem_dataset(scans = list(scan1, scan2), 
                             mask = mask_vol, 
                             TR = 2.0, 
                             # run_length automatically inferred as c(20, 20)
                             event_table = events_df)

print(mem_dset)
# Access components
print(mem_dset$sampling_frame)
```

## File-Based Volumetric Data (`fmri_file_dataset`)

This is often the most practical option for typical fMRI analyses where data resides in files.

**Key Arguments:**

*   `scans`: A character vector of file paths to the 4D fMRI image files (e.g., `.nii.gz`), one path per run.
*   `mask`: A character string giving the file path to the 3D mask image file.
*   `TR`: Repetition time (seconds).
*   `run_length`: A numeric vector specifying the number of volumes (time points) in each run file listed in `scans`.
*   `event_table` (Optional): A `data.frame` with experimental design info.
*   `base_path` (Optional): A path to prepend to relative file paths in `scans` and `mask`.
*   `preload` (Optional, Default: `FALSE`): If `TRUE`, load the mask and scan data into memory immediately. If `FALSE` (recommended for large data), data is read only when accessed.
*   `mode` (Optional): Storage mode for `neuroim2` when reading data (e.g., "normal", "mmap").

```{r file_dataset_example}
# --- Create Dummy Files (for illustration only) ---
# In a real analysis, these files would already exist.
tmp_dir <- tempdir()
mask_filename <- "mask.nii.gz"
scan1_filename <- "run1.nii.gz"
scan2_filename <- "run2.nii.gz"

mask_file_full_path <- file.path(tmp_dir, mask_filename)
scan1_file_full_path <- file.path(tmp_dir, scan1_filename)
scan2_file_full_path <- file.path(tmp_dir, scan2_filename)

# Create small dummy mask and scans using neuroim2 functionality
d <- c(5, 5, 5) # Mask dimensions
d_run1 <- c(d, 20) # Run 1 dimensions (time=20)
d_run2 <- c(d, 25) # Run 2 dimensions (time=25)

mask_vol_dummy <- neuroim2::NeuroVol(array(1, d), neuroim2::NeuroSpace(d))
scan1_dummy <- neuroim2::NeuroVec(array(rnorm(prod(d_run1)), d_run1), neuroim2::NeuroSpace(d_run1))
scan2_dummy <- neuroim2::NeuroVec(array(rnorm(prod(d_run2)), d_run2), neuroim2::NeuroSpace(d_run2))

# Ensure dummy files are written using their full paths
neuroim2::write_vol(mask_vol_dummy, mask_file_full_path)
neuroim2::write_vec(scan1_dummy, scan1_file_full_path)
neuroim2::write_vec(scan2_dummy, scan2_file_full_path)
# --- End Dummy File Creation ---

# Create the file-based dataset object
# Pass only filenames to 'scans' and 'mask', and specify the directory in 'base_path'
file_dset <- fmri_dataset(scans = c(scan1_filename, scan2_filename), 
                            mask = mask_filename, 
                            TR = 1.5, 
                            run_length = c(20, 25), # Must match time dim of files
                            event_table = events_df, 
                            base_path = tmp_dir,    # Set base_path to the temp directory
                            preload = FALSE) # Keep data on disk

# This print statement should now work
print(file_dset)

# Clean up dummy files (optional, commented out for vignette)
# file.remove(mask_file_full_path, scan1_file_full_path, scan2_file_full_path)
```

Using `preload=FALSE` is memory-efficient as only the required data segments are read when needed (e.g., during model fitting).

## Matrix Data (`matrix_dataset`)

Use this if your fMRI data is already represented as a 2D matrix where rows are time points and columns are voxels or components (e.g., after surface projection or ROI averaging).

**Key Arguments:**

*   `datamat`: The numeric matrix (time x features).
*   `TR`: Repetition time (seconds).
*   `run_length`: Vector specifying the number of rows (time points) belonging to each run.
*   `event_table` (Optional): A `data.frame` with design info (must have same total number of rows as `datamat`).

```{r matrix_dataset_example}
# Example matrix (100 time points, 50 features/voxels)
# Two runs of 50 time points each
time_points <- 100
features <- 50
run_len <- c(50, 50)
example_matrix <- matrix(rnorm(time_points * features), time_points, features)

# Example event table for matrix data
events_mat_df <- data.frame(
  onset = c(seq(5, 45, by=10), seq(5, 45, by=10)), 
  condition = factor(rep(c("C", "D"), 10)),
  run = rep(1:2, each = 5)
)

mat_dset <- matrix_dataset(datamat = example_matrix, 
                           TR = 2.5, 
                           run_length = run_len,
                           event_table = events_mat_df)

print(mat_dset)
print(mat_dset$sampling_frame)
```

For `matrix_dataset`, the concept of a spatial mask is implicit; all columns provided in `datamat` are included.

## Latent Data (`latent_dataset`)

This class is designed for data that has undergone dimensionality reduction (e.g., PCA, ICA). It wraps a `LatentNeuroVec` object, which stores the basis vectors (latent components over time) and loadings (spatial maps of components). Creating and using `LatentNeuroVec` objects typically requires the `fmristore` package.

**Key Arguments:**

*   `lvec`: A `LatentNeuroVec` object from the `fmristore` package.
*   `TR`: Repetition time (seconds).
*   `run_length`: Vector specifying run lengths (must sum to the time dimension of `lvec`).
*   `event_table` (Optional): Experimental design `data.frame`.

```{r latent_dataset_example, eval=FALSE}
# Conceptual example (requires fmristore package and a LatentNeuroVec)
# Assuming 'my_latent_neuro_vec' is a LatentNeuroVec object representing
# 20 components over 300 time points (2 runs of 150)

# latent_dset <- latent_dataset(lvec = my_latent_neuro_vec, 
#                              TR = 2.0, 
#                              run_length = c(150, 150),
#                              event_table = some_event_df)
# 
# print(latent_dset)
```

This dataset type essentially behaves like a `matrix_dataset` where the matrix columns are the latent component time series.

## Using Dataset Objects

Once created, these dataset objects serve as the primary data input for `fmrireg`'s modeling functions:

*   `event_model(..., sampling_frame = dset$sampling_frame)`
*   `baseline_model(..., sframe = dset$sampling_frame)`
*   `fmri_lm(model, dataset = dset)`
*   `estimate_betas(..., dataset = dset)`

They provide a standardized way to access data (`get_data(dset)`), masks (`get_mask(dset)`), and timing information (`blocklens(dset)`, `blockids(dset)`), regardless of the underlying storage format.

Choosing the appropriate dataset class depends on where your data resides (memory, files) and its format (volumetric, matrix, latent). 