---
title: "Using the fmriDSL Configuration System"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the fmriDSL Configuration System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(fmrireg)
library(yaml)
```

## Introduction

The fmrireg package provides a declarative configuration system (fmriDSL - Domain Specific Language) that allows users to specify complex fMRI analysis designs using a simple YAML format. This approach offers several advantages:

- **Reproducibility**: The analysis configuration is stored in a human-readable text file that can be version-controlled and shared
- **Separation of concerns**: The analysis specification is completely separated from the analysis execution code
- **Validation**: The configuration is validated both at the schema level and against the actual BIDS dataset
- **Standardization**: Analysis parameters are organized into a standardized structure
- **Documentation**: The configuration serves as self-documentation of the analysis choices

This vignette demonstrates how to write, validate, and use a YAML configuration file with the fmriDSL system. The configuration is processed by the `load_fmri_config()` function, which returns an `fmri_config` object that can be used to build and fit models.

## Top-Level YAML Structure

A complete fmriDSL configuration consists of the following top-level sections:

```yaml
dataset:      # [Required] BIDS dataset paths and selection criteria
events:       # [Required] Event column mapping definitions
regressors:   # [Required] Regressor specifications
model:        # [Required] Model definition and mappings

# Optional sections
confounds:    # Confound variable specifications
baseline:     # Baseline/drift model specifications  
hrfs:         # Custom hemodynamic response function definitions
parametric:   # Parametric transformation specifications
contrasts:    # Statistical contrast definitions
```

Let's explore each section in detail.

## Dataset Section

The `dataset` section specifies the BIDS dataset location and selection criteria.

```yaml
dataset:
  path: /path/to/bids/dataset           # [Required] Absolute path to BIDS dataset
  relpath: func                         # [Required] Relative path to functional data (usually "func")
  subjects:                             # [Optional] Subject selection criteria
    include: [sub-01, sub-02, sub-03]   # Specific subjects to include (BIDS format)
    exclude: [sub-04]                   # Subjects to exclude
  tasks: [task-stroop, task-memory]     # [Optional] Tasks to include (if omitted, all available tasks are used)
  runs: [run-01, run-02]                # [Optional] Runs to include (if omitted, all available runs are used)
  scan_params:                          # [Optional] Scan parameters
    TR:                                 # TR (repetition time) in seconds
      default: 2.0                      # Default TR for all scans
      overrides:                        # TR overrides for specific patterns
        - pattern: task-restingstate    # Pattern to match in BIDS filename
          value: 1.5                    # TR value for matching files
    run_length:                         # Run duration in volumes
      default:                          # Default run lengths by task
        task-stroop: 180                # 180 volumes for stroop task
        task-memory: 240                # 240 volumes for memory task
      overrides:                        # Run length overrides for specific patterns
        - pattern: sub-01_task-stroop   # Pattern to match in BIDS filename
          value: 200                    # Run length value for matching files
```

The `path` field should point to the root of your BIDS dataset. The `subjects`, `tasks`, and `runs` sections allow you to select specific portions of your dataset. If these are omitted, all available data will be used.

## Events Section

The `events` section defines how to map columns in the BIDS events.tsv files to variables in your model.

```yaml
events:
  onset: onset           # [Required] Column name for event onsets (in seconds)
  duration: duration     # [Required] Column name for event durations (in seconds)
  block: run             # [Required] Column name for block/run identifier
  
  # Custom variable mappings (optional but typical)
  condition:             # Define a variable named 'condition'
    column: trial_type   # Maps to the 'trial_type' column in events.tsv
    onset: custom_onset  # [Optional] Override the default onset column for this variable 
    duration: custom_dur # [Optional] Override the default duration column for this variable
  
  response_time:         # Define a variable named 'response_time'
    column: rt           # Maps to the 'rt' column in events.tsv
    
  accuracy:              # Define a variable named 'accuracy'
    column: correct      # Maps to the 'correct' column in events.tsv
```

The `onset`, `duration`, and `block` fields are required and specify the column names in your events.tsv files that contain this information. Additional variables can be defined by mapping them to columns in the events.tsv files.

## Regressors Section

The `regressors` section defines how to create regressors from your event data. This is where you specify the design of your GLM.

```yaml
regressors:
  # Simple categorical regressor using canonical HRF
  faces:                      # Regressor name
    type: hrf                 # Type: Convolve events with HRF
    variables: [condition]    # Variable to use (defined in events section)
    subset: "condition == 'face'"  # Filter condition (optional)
    hrf: canonical            # HRF to use (defined in hrfs section or built-in)
    lag: 0                    # [Optional] Additional lag to apply to the HRF (in seconds)
    transform: [zscore]       # [Optional] Transformations to apply to the regressor values
    
  # Parametric modulation regressor
  rt_mod:
    type: hrf_parametric      # Type: Parametric modulator
    variables: [response_time]  # Parametric variable to use
    subset: "accuracy == 1"   # Only correct trials (optional)
    hrf: canonical            # HRF to use
    basis:                    # Basis expansion for parametric effect
      type: polynomial        # Polynomial expansion (or 'Poly', case-insensitive)
      parameters:
        degree: 2             # Quadratic expansion (degree 2)
        
  # Trialwise regressor (no convolution)
  motion_outliers:
    type: trialwise           # Type: No convolution, direct inclusion
    variables: [framewise_displacement]  # Variable to use
    transform: [zscore]       # Optional transformation
```

Each regressor is given a name (e.g., `faces`, `rt_mod`, `motion_outliers`) and requires a `type` field that determines how it's constructed:

- `hrf`: Standard convolution of events with an HRF
- `hrf_parametric`: Parametric modulation with HRF convolution
- `trialwise`: Direct inclusion without convolution (e.g., for motion parameters)

The `variables` field specifies which variable(s) from the `events` section to use, and `subset` allows filtering specific events.

## HRFs Section

The `hrfs` section allows you to define custom hemodynamic response functions.

```yaml
hrfs:
  canonical:                # HRF name
    type: HRF_SPMG1         # Built-in canonical HRF
    lag: 0                  # [Optional] Additional lag to apply (in seconds)
    width: 32               # [Optional] Width of the HRF in seconds
    summate: true           # [Optional] Whether to sum the response over each trial duration
    normalize: true         # [Optional] Whether to normalize the HRF to have a peak height of 1
    
  double_gamma:             # Custom HRF name
    type: HRF_SPMG2         # Double gamma function (canonical + derivative)
    
  custom_gaussian:          # Custom HRF name
    type: HRF_GAUSSIAN      # Gaussian HRF
    parameters:             # Parameters for this HRF
      mean: 5               # Mean at 5 seconds
      sd: 1.5               # Standard deviation of 1.5 seconds
      
  custom_basis:             # Custom HRF name
    type: HRF_BSPLINE       # B-spline basis set
    parameters:             # Parameters for this basis
      knots: [0, 4, 8, 12, 16, 20]  # Knot locations in seconds
      degree: 3             # Cubic B-splines
      
  user_defined:             # For custom HRF functions defined by the user
    type: custom            # Indicates a user-defined HRF
    definition: "my_hrf_function"  # Name of the R function to use
```

The package provides several built-in HRF types:

- `HRF_SPMG1`: Standard canonical HRF
- `HRF_SPMG2`: Canonical HRF with temporal derivative
- `HRF_SPMG3`: Canonical HRF with temporal and dispersion derivatives
- `HRF_GAUSSIAN`: Gaussian HRF
- `HRF_GAMMA`: Gamma HRF
- `HRF_BSPLINE`: B-spline basis set
- `HRF_DAGUERRE`: Daguerre basis
- `HRF_DAGUERRE_BASIS`: Daguerre basis set

## Parametric Section

The `parametric` section defines transformations for parametric variables.

```yaml
parametric:
  response_time:            # Variable name (from events section)
    transform: [center, log]  # Transformations to apply (in order)
    
  framewise_displacement:   # Variable name
    transform: [zscore]     # Z-score transformation
    
  accuracy:                 # Variable name
    basis:                  # Basis expansion
      type: polynomial      # Polynomial basis (or 'Poly', case-insensitive)
      parameters:
        degree: 2           # Quadratic (degree 2)
```

Supported transformations include:

- `center`: Center variable to mean 0
- `scale`: Scale variable to standard deviation 1
- `zscore`: Center and scale (equivalent to `[center, scale]`)
- `log`: Natural logarithm
- `exp`: Exponential
- `within_subject`: Center within each subject

Basis expansions include:

- `polynomial` or `Poly`: Polynomial expansion
- `bspline` or `BSpline`: B-spline expansion
- `ns`: Natural spline expansion
- `Standardized`: Standardized version of the variable
- `Ident`: Identity transformation (no change)

## Baseline Section

The `baseline` section configures the baseline/drift model.

```yaml
baseline:
  basis: bspline            # Basis function for drift modeling
  degree: 3                 # Degree of basis expansion
  intercept: runwise        # Intercept handling: runwise, global, or none
  confounds:                # Confound variables (optional)
    include: ["trans_.*", "rot_.*"]  # Include translation and rotation parameters
    exclude: [".*_derivative.*"]     # Exclude derivatives
```

The `basis` can be one of:

- `constant`: No drift modeling, just intercept
- `poly`: Polynomial drift
- `bspline`: B-spline drift (recommended)
- `ns`: Natural spline drift

The `intercept` setting determines how intercepts are handled:

- `runwise`: Separate intercept for each run (recommended)
- `global`: Single intercept across all runs
- `none`: No intercept

## Confounds Section

The `confounds` section specifies which confound variables to include from the BIDS confounds.tsv files.

```yaml
confounds:
  include: ["trans_x", "trans_y", "trans_z", "rot_x", "rot_y", "rot_z"]  # Motion parameters
  exclude: []  # Patterns to exclude (optional)
```

The `include` field specifies which variables to include. Regex patterns are supported. If omitted, all available confounds are included.

## Contrasts Section

The `contrasts` section defines statistical contrasts for your model.

```yaml
contrasts:
  # Simple contrast using a formula
  faces_vs_houses:            # Contrast name
    type: formula             # Formula-based contrast
    expression: "faces - houses"  # Linear combination of regressor names
    where: "condition != 'other'"  # [Optional] Filter expression for data
    include: "sub-01.*"       # [Optional] Include pattern for subjects/runs
    exclude: ".*_broken"      # [Optional] Exclude pattern for subjects/runs
    
  # One condition versus all others
  faces_vs_all:
    type: one_against_all     # One condition vs all others
    factors: [condition]      # Factor to use
    levels: [faces]           # Level to contrast against others
    
  # Contrast for a specific factor
  main_effect_task:
    type: oneway              # One-way ANOVA style contrast
    factors: [condition]      # Factor to use
    
  # Interaction contrast
  task_by_difficulty:
    type: interaction         # Interaction contrast
    factors: [condition, difficulty]  # Factors that interact
    
  # Pairwise contrasts
  pairwise_conditions:
    type: pair                # Pairwise contrasts between levels
    factors: [condition]      # Factor to use
    
  # Individual conditions vs baseline
  unit_contrasts:
    type: unit                # Each condition vs baseline
    factors: [condition]      # Factor to use
    
  # Polynomial contrasts for ordinal factors
  difficulty_trend:
    type: polynomial          # Polynomial trend analysis
    factors: [difficulty]     # Ordinal factor
    degree: 2                 # Degree of polynomial (quadratic)
```

Supported contrast types include:

- `formula`: Manual formula with regressor names
- `one_against_all`: One level vs all others
- `pair`: Pairwise comparisons
- `unit`: Unit contrasts (each condition vs baseline)
- `oneway`: Main effect of factor
- `interaction`: Interaction between factors
- `polynomial`: Polynomial contrasts for ordered factors

The optional fields `where`, `include`, and `exclude` can be used to filter contrasts by data condition or subject/run patterns.

## Model Section

The `model` section ties everything together, defining which regressors and contrasts to include in your specific model.

```yaml
model:
  name: full_model           # Model name
  factors: [condition, difficulty]  # Variables to treat as factors
  parametric: [response_time]  # Variables to treat as parametric
  
  # Model-specific event mappings
  events:
    onset: onset             # Onset column
    duration: duration       # Duration column
    block: run               # Block/run column
    condition: trial_type    # Map model variable 'condition' to column 'trial_type'
    difficulty: difficulty   # Map model variable 'difficulty' to column 'difficulty'
    response_time: rt        # Map model variable 'response_time' to column 'rt'
  
  # Model-specific regressor selection
  regressors:
    faces:                   # Include the 'faces' regressor defined earlier
    houses:                  # Include the 'houses' regressor
    rt_mod:                  # Include the 'rt_mod' parametric regressor
    
  # Model-specific contrast selection
  contrasts:
    faces_vs_houses:         # Include the 'faces_vs_houses' contrast
    main_effect_task:        # Include the 'main_effect_task' contrast
```

The `factors` and `parametric` fields declare how variables should be treated in the model. The `events` section maps model variables to event file columns. The `regressors` and `contrasts` sections select which regressors and contrasts to include in this particular model.

## Loading and Validation

Once you've created your YAML configuration file, you can load it with the `load_fmri_config()` function:

```{r, eval=FALSE}
config <- load_fmri_config("my_analysis_config.yaml")
```

The function performs two levels of validation:

1. **Schema validation**: Checks that the YAML structure matches the expected schema, all required fields are present, and values have the correct types
2. **Context-dependent validation**: Checks that the configuration is valid for your specific BIDS dataset (subjects exist, tasks exist, event columns exist, etc.)

If validation fails, you'll receive detailed error messages explaining what needs to be fixed. For example, if a required field is missing, you might see "Required field 'X' is missing at path 'Y'", or if an event column doesn't exist in your data, you might see "Column 'Z' not found in event file for subject 'sub-01', task 'task-X'".

## Example

Below is a complete example of a configuration file and how to load it:

```yaml
# Complete example configuration (my_analysis_config.yaml)
dataset:
  path: /path/to/bids/dataset
  relpath: func
  subjects:
    include: [sub-01, sub-02, sub-03]
  tasks: [task-face]

events:
  onset: onset
  duration: duration
  block: run
  condition:
    column: trial_type
  response_time:
    column: rt

hrfs:
  canonical:
    type: HRF_SPMG1

regressors:
  faces:
    type: hrf
    variables: [condition]
    subset: "condition == 'face'"
    hrf: canonical
  houses:
    type: hrf
    variables: [condition]
    subset: "condition == 'house'"
    hrf: canonical
  rt_mod:
    type: hrf_parametric
    variables: [response_time]
    hrf: canonical
    basis:
      type: polynomial
      parameters:
        degree: 1

baseline:
  basis: bspline
  degree: 3
  intercept: runwise

confounds:
  include: ["trans_x", "trans_y", "trans_z", "rot_x", "rot_y", "rot_z"]

contrasts:
  faces_vs_houses:
    type: formula
    expression: "faces - houses"
  main_effect_condition:
    type: oneway
    factors: [condition]

model:
  name: face_vs_house
  factors: [condition]
  parametric: [response_time]
  events:
    onset: onset
    duration: duration
    block: run
    condition: trial_type
    response_time: rt
  regressors:
    faces:
    houses:
    rt_mod:
  contrasts:
    faces_vs_houses:
    main_effect_condition:
```

Loading and inspecting the configuration:

```{r, eval=FALSE}
# Load the configuration
config <- load_fmri_config("my_analysis_config.yaml")

# Inspect the configuration
print(config)
```

The `print` method for `fmri_config` objects provides a convenient summary of the loaded configuration:

```
fMRI Analysis Configuration

=== Dataset ===
BIDS Project Path: /path/to/bids/dataset
Relative Path (func): func
Selected Subjects: 3 total
  - sub-01
  - sub-02
  - sub-03
Selected Tasks:
  - task-face

=== Events ===
--- Top-Level Event Column Mappings ---
  Onset Column: 'onset'
  Duration Column: 'duration'
  Block Column: 'run'
--- Variable Mappings (Top Level) ---
  condition -> column: 'trial_type'
  response_time -> column: 'rt'

=== Model: face_vs_house ===
--- Model Variable Roles ---
Factors:
  - condition
Parametric:
  - response_time
--- Model-Specific Event Variable -> Column Maps ---
  Onset Column (model): 'onset'
  Duration Column (model): 'duration'
  Block Column (model): 'run'
  condition -> column: 'trial_type'
  response_time -> column: 'rt'
--- HRF Specifications (Used in Model) ---
  canonical: type=HRF_SPMG1
--- Regressors (Used in Model) ---
  faces: type=hrf, vars=[condition]
  houses: type=hrf, vars=[condition]
  rt_mod: type=hrf_parametric, vars=[response_time], basis=polynomial

=== Build Status ===
Status: ✓ Successfully Built
```

## Using the Configuration

Once you have loaded and validated a configuration, you can use it to build and fit models. The `fmri_config` object contains all the necessary information about your dataset, variables, and model specifications.

Here's a brief overview of how you might use the configuration in your analysis workflow:

```{r, eval=FALSE}
# Load the configuration
config <- load_fmri_config("my_analysis_config.yaml")

# Build a subject-level model
subject_id <- "sub-01"
subject_model <- build_subject_model(config, subject_id)

# Fit the model
fit <- fit_model(subject_model)

# Generate contrasts
contrasts <- generate_contrasts(fit, config)

# Save results
save_results(fit, contrasts, output_dir = "results")
```

**Note:** The function names above (`build_subject_model`, `fit_model`, `generate_contrasts`, and `save_results`) are illustrative examples of the typical workflow steps that would use the `config` object, not necessarily actual function names in the package. Refer to the package documentation for the specific functions available for model building and fitting.

## Conclusion

The fmriDSL configuration system provides a powerful and flexible way to specify complex fMRI analyses. By separating the analysis specification from the analysis code, it promotes reproducibility and facilitates sharing analysis designs. The validation system helps catch errors early, before running computationally expensive analyses.

The YAML format is human-readable and easy to understand, while still capturing all the complexity needed for advanced fMRI modeling. We recommend using this approach for all your fMRI analyses with the fmrireg package. 