---
title: "01 Overview and Workflow"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01 Overview and Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=7, fig.height=5, message=FALSE, warning=FALSE)
library(fmrireg)
library(fmrihrf)
```

## Introduction

If you work with fMRI data, you already know the core analysis is multiple regression applied voxel-by-voxel — but specifying the design matrix, convolving events with hemodynamic response functions, and managing multi-run structures can be tedious and error-prone.

The `fmrireg` package gives you a concise, formula-based interface for these tasks, inspired by `lm()` and `glm()`. You describe your experimental design and HRF choices in a formula, and fmrireg handles convolution, baseline modeling, contrast computation, and model fitting.

This vignette provides a high-level overview of the key steps and components involved in using `fmrireg`.

## Key Components and Workflow

A typical fMRI analysis workflow with `fmrireg` involves these main stages:

1.  **Data Representation:** Defining how your fMRI data and experimental design are organized.
2.  **Model Building:** Specifying the regression model using formulas, including:
    *   Defining Hemodynamic Response Functions (HRFs).
    *   Creating event-related regressors (`event_model`).
    *   Modeling baseline components like drift and nuisance factors (`baseline_model`).
    *   Defining contrasts of interest (`contrast`).
3.  **Model Estimation:** Fitting the specified model to the data (e.g., using `fmri_lm`).
4.  **Advanced Topics:** Exploring single-trial estimation, simulation, or translation to other packages.

### 1. Data Representation

Before modeling, you need to represent your fMRI data and its structure. `fmrireg` uses objects like:

*   `fmri_dataset` / `fmri_mem_dataset`: Encapsulates the 4D fMRI data (potentially file-backed or in-memory), mask information, TR, run structure, and the experimental design table.
*   `sampling_frame`: Defines the temporal structure – the number of scans per run (`blocklens`) and the repetition time (`TR`).

```{r sampling_frame_example}
# Example: 5 runs, 200 scans/run, TR=2s
sframe <- sampling_frame(blocklens = rep(200, 5), TR = 2)
sframe
```

### 2. Model Building

#### a) Hemodynamic Response Function (HRF)

The HRF models the BOLD signal response to a brief neural event. `fmrireg` provides various pre-defined HRFs (e.g., `HRF_SPMG1`, `HRF_GAUSSIAN`) and functions to create or modify them (`gen_hrf`, `hrf_blocked`, `hrf_lagged`).

```{r hrf_example, fig.alt = "Line plot of the SPM Canonical HRF (BOLD response over time in seconds)"}
# Plot the SPM canonical HRF
time <- seq(0, 24, by = 0.2)
plot(time, HRF_SPMG1(time), type = 'l', xlab = "Time (s)", ylab = "BOLD Response",
     main = "SPM Canonical HRF (HRF_SPMG1)")
```

*   **Related Topic:** Hemodynamic Response Functions

#### b) Event Model (`event_model`)

This defines the task-related part of your GLM. You specify how experimental events (onsets, conditions, durations) map to regressors using a formula syntax. The `hrf()` or `afni_hrf()` functions within the formula link event variables to specific HRF shapes.

```{r event_design_table}
# A simple design table with two stimulus types
design_table <- data.frame(
  onset = c(10, 40, 70, 100, 130, 160, 190, 220),
  stim_type = factor(rep(c("face", "scene"), 4)),
  run_id = rep(1, 8)
)
```

```{r event_model_example}
# Model the effect of stim_type, convolved with the SPM canonical HRF
event_mod <- event_model(onset ~ hrf(stim_type),
                         data = design_table,
                         block = ~ run_id,
                         sampling_frame = sframe)
event_mod
```

*   **Related Topic:** Event Models

#### c) Baseline Model (`baseline_model`)

This models non-task-related components like scanner drift, run intercepts, and other nuisance factors (e.g., motion parameters).

```{r baseline_model_example}
# Example: Cubic B-spline drift model + run intercepts
# (Requires 'sframe' defined earlier)
base_mod <- baseline_model(basis = "bs", degree = 3, sframe = sframe)
print(base_mod)
```

*   **Related Topic:** Baseline Models

#### d) Contrasts (`pair_contrast`, `unit_contrast`, etc.)

Contrasts define specific comparisons between condition estimates.

```{r contrast_example}
# Example: Contrast comparing 'face' vs 'scene' levels of 'stim_type'
face_vs_scene <- pair_contrast(~ stim_type == "face", ~ stim_type == "scene", name = "FvS")
print(face_vs_scene)
```

*   **See Vignette:** `vignette("a_05_contrasts", package = "fmrireg")`

### 3. Model Estimation (`fmri_model`, `fmri_lm`)

You combine the event and baseline models into a full `fmri_model` and then fit it to your `fmri_dataset` using estimation functions like `fmri_lm` (for standard GLM) or `estimate_betas` (for single-trial betas).

```{r fmri_model_example, eval=FALSE}
# Combine event and baseline models
full_model <- fmri_model(event_mod, base_mod)

# Fit the GLM
# (Requires 'my_dataset' of class fmri_dataset)
fit <- fmri_lm(full_model, dataset = my_dataset)

# Robust fit down-weighting outlier time points
fit_robust <- fmri_lm(full_model, dataset = my_dataset, robust = TRUE)

# Extract contrast results (if contrasts were defined)
# con_results <- coef(fit)  # Get coefficients
```

Setting `robust = TRUE` invokes row-wise weighting to lessen the influence of
motion spikes or other frame-level artifacts.

*   **See Vignettes:**
    *   `vignette("a_09_linear_model", package = "fmrireg")`
    *   Linear Models and Trialwise Analysis

### 4. Advanced Topics

`fmrireg` also supports:

*   **Simulation:** Creating artificial fMRI data (`simulate_fmri_matrix`).
    *   `vignette("a_08_simulation", package = "fmrireg")`
*   **AFNI Translation:** Generating scripts for AFNI's `3dDeconvolve` (`afni_lm`).
*   **Low-level Regressors:** Manually creating and manipulating regressors (`regressor`).

#### Acceleration: Sketched / Nyström Analysis

For large problems, you can accelerate voxelwise GLM estimation with:
- SRHT/IHS temporal sketching to reduce timepoints while preserving LS geometry.
- Nyström/landmark spatial extension to solve on a subset of voxels and extend with parcel‑aware kernels.

See `vignette("sketched-ar", package = "fmrireg")` for guidance, tradeoffs, and practical defaults.

#### Group-Level Meta-Analysis

Combine subject‑level maps or ROI summaries using fixed/random‑effects meta‑regression:
- Inputs: HDF5/NIfTI (beta+SE or t+df) or ROI CSV.
- Random‑effects: Paule–Mandel/DerSimonian–Laird/REML alias; optional robust Huber.
- t‑only combine: Stouffer/Fisher/Lancaster (supports equal/IVW/custom weights where applicable).
- Exact contrasts: at fit‑time (contrasts=...) or post‑hoc via `return_cov = "tri"` + `contrast()`.
- Multiple comparisons: BH FDR or spatial FDR (block/parcel aware).

See `vignette("group_analysis", package = "fmrireg")` for end‑to‑end examples.

## Getting Started

This overview provides a map of the `fmrireg` package. For detailed explanations and examples, please refer to the specific vignettes linked above.

## Next Steps

- `vignette("a_10_dataset", package = "fmrireg")` — Dataset management
- `vignette("a_08_simulation", package = "fmrireg")` — Simulating fMRI data
- `vignette("group_analysis", package = "fmrireg")` — Group analysis
- `vignette("sketched-ar", package = "fmrireg")` — Sketched/Nyström acceleration
