---
title: Group Analysis
author: Bradley R. Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Group Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=7, fig.height=5, message=FALSE, warning=TRUE)
load_vignette_package <- function(package) {
  withCallingHandlers(
    suppressPackageStartupMessages(library(package, character.only = TRUE)),
    warning = function(w) {
      if (grepl("^package '.+' was built under R version", conditionMessage(w))) {
        message("Qualified host note: ", conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    }
  )
}
invisible(lapply(c("fmrireg", "ggplot2"), load_vignette_package))
set.seed(123)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Overview

This vignette walks through a compact, end-to-end example of group analysis with fmrireg. We first construct a small ROI-level dataset to illustrate the basic meta-regression interface, and then move to a voxelwise example using tiny synthetic NIfTI files. Along the way we show how to compare groups with fixed- and random-effects meta-analysis, how to obtain exact contrasts either at fit-time or post-hoc by saving covariance, and how to perform group inference from t-maps alone using Stouffer, Fisher, or Lancaster combinations.

## Create a small ROI dataset

We simulate 10 subjects across two groups (A/B) for a single ROI. Group B has
an effect that is 1 unit larger than group A. All subjects have the same SE for
clarity.

```{r create-roi-data}
n_per_group <- 5
subjects <- sprintf("s%02d", 1:(2 * n_per_group))
group <- factor(rep(c("A", "B"), each = n_per_group))

beta <- ifelse(group == "A", 0.5, 1.5)
se <- rep(0.25, length(beta))

roi_df <- data.frame(
  subject = subjects,
  roi = "ExampleROI",
  beta = beta,
  se = se,
  group = group,
  stringsAsFactors = FALSE
)

gd <- fmrireg::group_data(
  roi_df,
  format = "csv",
  effect_cols = c(beta = "beta", se = "se"),
  subject_col = "subject",
  roi_col = "roi",
  covariate_cols = c("group")
)

gd
```

## Fit group meta-analysis

We first fit an intercept-only model, then a model including a group term.

```{r fit-intercept-and-group}
fit_fe <- fmri_meta(gd, formula = ~ 1, method = "fe", verbose = FALSE)

fit_cov <- fmri_meta(gd, formula = ~ 1 + group, method = "fe", verbose = FALSE)

print(fit_cov)
summary(fit_cov)
```

Note: With equal SE per subject, fixed-effects and random-effects will yield
similar point estimates. Random-effects (`method = "pm"`) will estimate between-
subject heterogeneity (`tau2`) when present.

```{r fit-random-effects}
fit_pm <- fmri_meta(gd, formula = ~ 1 + group, method = "pm", verbose = FALSE)
summary(fit_pm)
```

## Extract coefficients and a contrast

```{r extract-coefficients}
coef_names <- colnames(fit_cov$coefficients)
group_term_roi <- grep("^group", coef_names, value = TRUE)
group_weights_roi <- setNames(as.numeric(coef_names == group_term_roi), coef_names)
con_roi <- contrast(fit_cov, group_weights_roi)

data.frame(
  term = coef_names,
  estimate = as.numeric(fit_cov$coefficients[1, ]),
  standard_error = as.numeric(fit_cov$se[1, ])
)
```

```{r validate-roi-group-effect, include=FALSE}
stopifnot(
  identical(group_term_roi, "groupB"),
  abs(con_roi$estimate - 1) < 1e-10,
  is.finite(con_roi$se),
  con_roi$se > 0
)
```

## A quick visualization

We can visualize the group effects and their 95% CIs for the ROI-level fit.

```{r plot-group-effects, fig.alt="ROI meta-regression coefficients with 95 percent confidence intervals for the intercept and group B effect."}
df_tidy <- data.frame(
  term = colnames(fit_cov$coefficients),
  estimate = as.numeric(fit_cov$coefficients[1, ]),
  std_error = as.numeric(fit_cov$se[1, ])
)
df_tidy$conf.low <- df_tidy$estimate - qnorm(0.975) * df_tidy$std_error
df_tidy$conf.high <- df_tidy$estimate + qnorm(0.975) * df_tidy$std_error
df_tidy

ggplot(df_tidy, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "ROI-level group meta-analysis",
       x = "Term", y = "Estimate +/- 95% CI") +
  theme_minimal()
```

## Notes on voxelwise analysis

For voxelwise analysis, construct `group_data` with format `"nifti"` or `"h5"`:

```{r voxelwise-note, eval=FALSE}
# HDF5 (produced via write_results.fmri_lm)
# gd_h5 <- fmrireg::group_data(h5_paths, format = "h5",
#                     subjects = subject_ids,
#                     covariates = covariates_df,
#                     contrast = "ContrastName")

# NIfTI (provide per-subject paths for beta/SE or t)
# gd_nii <- fmrireg::group_data(list(beta = beta_paths, se = se_paths), format = "nifti",
#                      subjects = subject_ids,
#                      mask = "group_mask.nii.gz")

# fit <- fmri_meta(gd_h5, formula = ~ 1 + group, method = "pm")
# fit <- fmri_meta(gd_nii, formula = ~ 1, method = "fe")
```

For multiple comparisons correction that leverages spatial structure, see
`spatial_fdr()` and `create_3d_blocks()`.

## Minimal NIfTI Example (Reproducible)

This chunk creates tiny synthetic NIfTI volumes on disk (in a temp dir) for a
voxelwise demonstration. Group B has a higher effect in a small cube.

```{r create-synthetic-nifti}
library(neuroim2)

set.seed(42)
tmpdir <- tempdir()
space <- NeuroSpace(c(8, 8, 8), spacing = c(2, 2, 2))

n_per_group <- 3
ids <- sprintf("sub-%02d", 1:(2 * n_per_group))
grp <- factor(rep(c("A", "B"), each = n_per_group))

active <- array(FALSE, dim = c(8, 8, 8))
active[3:5, 3:5, 3:5] <- TRUE

beta_paths <- character(length(ids))
se_paths   <- character(length(ids))

for (i in seq_along(ids)) {
  b <- array(0, dim = c(8, 8, 8))
  b[active] <- if (grp[i] == "A") 0.5 else 1.5
  b <- b + array(rnorm(length(b), sd = 0.05), dim = dim(b))
  v_beta <- NeuroVol(b, space)

  s <- array(0.25, dim = c(8, 8, 8))
  v_se <- NeuroVol(s, space)

  beta_paths[i] <- file.path(tmpdir, sprintf("%s_beta.nii.gz", ids[i]))
  se_paths[i]   <- file.path(tmpdir, sprintf("%s_se.nii.gz", ids[i]))
  write_vol(v_beta, beta_paths[i])
  write_vol(v_se,   se_paths[i])
}

mask_path <- file.path(tmpdir, "mask.nii.gz")
write_vol(NeuroVol(array(1, dim = c(8, 8, 8)), space), mask_path)

gd_nii <- fmrireg::group_data(
  list(beta = beta_paths, se = se_paths),
  format = "nifti",
  subjects   = ids,
  covariates = data.frame(group = grp),
  mask       = mask_path
)

fit_nii <- fmri_meta(gd_nii, formula = ~ 1 + group, method = "fe", verbose = FALSE)
fit_nii

group_col <- grep("group", colnames(fit_nii$coefficients))
group_term_nii <- colnames(fit_nii$coefficients)[group_col]
pvals <- 2 * pnorm(-abs(fit_nii$coefficients[, group_col] / fit_nii$se[, group_col]))
sfr <- spatial_fdr(
  fit_nii,
  p = group_term_nii,
  group = "blocks",
  empirical_null = FALSE
)

active_voxels <- as.vector(active)
discovery_summary <- data.frame(
  method = c("Uncorrected p < 0.05", "Spatial FDR, q < 0.05"),
  discoveries = c(sum(pvals < 0.05), sum(sfr$reject)),
  true_positives = c(sum(pvals < 0.05 & active_voxels), sum(sfr$reject & active_voxels)),
  false_positives = c(sum(pvals < 0.05 & !active_voxels), sum(sfr$reject & !active_voxels))
)
knitr::kable(discovery_summary, caption = "Discoveries against the known active cube")

img_group_est <- coef_image(fit_nii, group_term_nii, statistic = "estimate")
range(as.array(img_group_est), na.rm = TRUE)
```

```{r plot-spatial-fdr-truth, fig.width=9, fig.height=4, fig.alt="Side-by-side slice maps of the estimated group B minus A effect with a numeric color scale and spatial-FDR decisions classified against the known active cube."}
group_estimate_array <- array(
  fit_nii$coefficients[, group_col], dim = c(8, 8, 8)
)
decision_class <- ifelse(
  sfr$reject & active_voxels, 2L,
  ifelse(sfr$reject & !active_voxels, 3L,
         ifelse(active_voxels, 1L, 0L))
)
decision_array <- array(decision_class, dim = c(8, 8, 8))
slice_index <- 4L

effect_limit <- max(abs(group_estimate_array))
old_par <- par(no.readonly = TRUE)
layout(matrix(c(1, 2, 3), nrow = 1), widths = c(1, 1, 0.18))
par(mar = c(3, 3, 3, 1))
image(
  group_estimate_array[, , slice_index],
  zlim = c(-effect_limit, effect_limit),
  col = hcl.colors(64, "Blue-Red 3"), axes = FALSE, asp = 1,
  main = "Estimated B - A effect"
)
contour(active[, , slice_index], levels = 0.5, add = TRUE,
        drawlabels = FALSE, lwd = 2)
image(
  decision_array[, , slice_index],
  breaks = c(-0.5, 0.5, 1.5, 2.5, 3.5),
  col = c("gray90", "#0072B2", "#009E73", "#E69F00"),
  axes = FALSE, asp = 1, main = "Spatial-FDR decision vs truth"
)
legend(
  "bottomright",
  legend = c("correct null", "missed active", "true discovery", "false discovery"),
  fill = c("gray90", "#0072B2", "#009E73", "#E69F00"),
  cex = 0.8, bty = "n", inset = 0.02
)
par(mar = c(3, 0.5, 3, 3))
scale_values <- seq(-effect_limit, effect_limit, length.out = 64)
image(
  x = c(0, 1), y = scale_values,
  z = matrix(rep(scale_values, each = 2), nrow = 2),
  col = hcl.colors(64, "Blue-Red 3"), axes = FALSE,
  xlab = "", ylab = ""
)
axis(4, at = pretty(c(-effect_limit, effect_limit), n = 5), las = 1)
mtext("B - A", side = 4, line = 2.2)
par(old_par)
```

The black outline marks the known active cube. The decision panel distinguishes
true discoveries from both missed active voxels and false discoveries rather
than presenting a thresholded map without its simulation truth.

```{r validate-spatial-fdr-example, include=FALSE}
stopifnot(
  identical(group_term_nii, "groupB"),
  sum(active_voxels) == 27L,
  sum(sfr$reject & active_voxels) == 27L,
  sum(sfr$reject & !active_voxels) == 0L
)
```

Because the subject-level SE maps define the reference variance in this toy
example, we use the theoretical null (`empirical_null = FALSE`). Estimating an
empirical null from this deliberately under-dispersed background would rescale
tiny null fluctuations upward and obscure the intended multiple-testing lesson.

## Exact contrasts and stored covariance

You can request exact contrasts at fit-time or store per-voxel covariance for
exact post-hoc contrasts.

```{r stored-covariance}
fit_nii_pm <- fmri_meta(
  gd_nii, formula = ~ 1 + group, method = "pm",
  return_cov = "tri", verbose = FALSE
)

group_term_pm <- setdiff(colnames(fit_nii_pm$coefficients), "(Intercept)")
group_weights_pm <- setNames(
  as.numeric(colnames(fit_nii_pm$coefficients) == group_term_pm),
  colnames(fit_nii_pm$coefficients)
)
con <- contrast(fit_nii_pm, group_weights_pm)

contrast_summary <- data.frame(
  region = c("Active cube", "Background"),
  mean_estimate = c(mean(con$estimate[active_voxels]), mean(con$estimate[!active_voxels])),
  mean_standard_error = c(mean(con$se[active_voxels]), mean(con$se[!active_voxels]))
)
knitr::kable(contrast_summary, digits = 4, caption = "Exact post-hoc B - A contrast")

fit_nii_con <- fmri_meta(
  gd_nii, formula = ~ 1 + group, method = "pm",
  contrasts = matrix(group_weights_pm, nrow = 1,
                     dimnames = list("B_minus_A", names(group_weights_pm))),
  verbose = FALSE
)
fit_time_estimate <- fit_nii_con$contrasts$estimate[, 1]
fit_time_se <- fit_nii_con$contrasts$se[, 1]
fit_time_summary <- data.frame(
  region = c("Active cube", "Background"),
  mean_estimate = c(mean(fit_time_estimate[active_voxels]), mean(fit_time_estimate[!active_voxels])),
  mean_standard_error = c(mean(fit_time_se[active_voxels]), mean(fit_time_se[!active_voxels]))
)
knitr::kable(fit_time_summary, digits = 4, caption = "Exact fit-time B - A contrast")
```

```{r validate-exact-group-contrasts, include=FALSE}
stopifnot(
  identical(group_term_pm, "groupB"),
  abs(mean(con$estimate[active_voxels]) - 1) < 0.05,
  abs(mean(con$estimate[!active_voxels])) < 0.02,
  all(is.finite(con$se)),
  identical(fit_nii_con$contrasts$names, "B_minus_A"),
  max(abs(fit_time_estimate - con$estimate)) < 1e-10,
  max(abs(fit_time_se - con$se)) < 1e-10
)
```

## Two-sample t-test (Welch and OLS) on NIfTI

As an alternative to meta-analysis, we can run two-sample voxelwise t-tests
directly on the per-subject beta maps, using either Welch's unequal-variance
test or a standard OLS/Student t-test via a simple design matrix.

```{r welch-ttest}
fit_welch <- fmri_ttest(
  gd_nii, formula = ~ 1 + group, engine = "welch", sign = "BminusA"
)
t_welch   <- as.numeric(fit_welch$t["group", ])
df_welch  <- as.numeric(fit_welch$df["group", ])
p_welch   <- 2 * pt(abs(t_welch), df = df_welch, lower.tail = FALSE)

fit_ols <- fmri_ttest(
  gd_nii, formula = ~ 1 + group, engine = "classic", sign = "BminusA"
)
t_ols <- fit_ols$t["group", ]
df_ols <- as.numeric(fit_ols$df["group", ])
p_ols    <- 2 * pt(abs(t_ols), df = df_ols, lower.tail = FALSE)

timg_welch <- NeuroVol(array(NA_real_, dim = c(8, 8, 8)), space)
mask_img   <- if (!is.null(gd_nii$mask_data)) gd_nii$mask_data else neuroim2::read_vol(mask_path)
timg_welch[as.array(mask_img) > 0] <- t_welch
range(as.array(timg_welch), na.rm = TRUE)

welch_called <- p_welch < 0.05
ols_called <- p_ols < 0.05
ttest_discovery_summary <- data.frame(
  method = c("Welch", "OLS / Student"),
  discoveries = c(sum(welch_called), sum(ols_called)),
  true_positives = c(
    sum(welch_called & active_voxels),
    sum(ols_called & active_voxels)
  ),
  false_positives = c(
    sum(welch_called & !active_voxels),
    sum(ols_called & !active_voxels)
  )
)
knitr::kable(
  ttest_discovery_summary,
  caption = "Uncorrected p < 0.05 calls against known voxel truth"
)
```

```{r validate-ttest-direction-and-calls, include=FALSE}
stopifnot(
  mean(t_welch[active_voxels]) > 0,
  mean(t_ols[active_voxels]) > 0,
  all(ttest_discovery_summary$true_positives == sum(active_voxels)),
  all(ttest_discovery_summary$false_positives <
        ttest_discovery_summary$discoveries)
)
```

These are deliberately labelled uncorrected calls. Both methods recover the
active cube, but their extra background calls show why an uncorrected discovery
count is not a whole-brain inferential result. Use the spatial-FDR workflow
above (or another declared multiplicity procedure) for corrected decisions.

## Combining t-statistics only (Stouffer/Fisher/Lancaster)

When only per-subject t-statistics and degrees-of-freedom are available, you can still carry out group inference without betas/SEs by setting `combine =` in `fmri_meta()` (or in `fmri_ttest(..., engine = "meta")`). Stouffer combines signed z-scores and supports equal or explicitly supplied custom weights; inverse-variance weighting is unavailable without SEs. Fisher combines p-values with equal weights, and Lancaster provides a weighted Fisher variant by mapping its supplied weights to combination degrees of freedom.

```{r create-t-maps}
tmat <- t(vapply(seq_along(ids), function(i) {
  as.numeric(as.array(neuroim2::read_vol(beta_paths[i]))) /
    as.numeric(as.array(neuroim2::read_vol(se_paths[i])))
}, numeric(prod(dim(space)))))

t_paths <- character(length(ids))
for (i in seq_along(ids)) {
  img <- NeuroVol(array(NA_real_, dim = c(8, 8, 8)), space)
  img[as.array(neuroim2::read_vol(mask_path)) > 0] <- tmat[i, ]
  pth <- file.path(tmpdir, sprintf("%s_t.nii.gz", ids[i]))
  write_vol(img, pth)
  t_paths[i] <- pth
}

gd_t <- fmrireg::group_data(
  list(t = t_paths, df = 60),
  format = "nifti",
  subjects = ids,
  covariates = data.frame(group = grp),
  mask     = mask_path
)
```

```{r stouffer-combination}
fit_st <- fmri_meta(gd_t, formula = ~ 1, combine = "stouffer",
                    weights = "equal", verbose = FALSE)

w_subj <- rep(1, length(ids))
fit_st_w <- fmri_meta(gd_t, formula = ~ 1, combine = "stouffer",
                      weights = "custom", weights_custom = w_subj,
                      verbose = FALSE)

fit_fi  <- fmri_meta(gd_t, formula = ~ 1, combine = "fisher",
                     weights = "equal", verbose = FALSE)
fit_la  <- fmri_meta(gd_t, formula = ~ 1, combine = "lancaster",
                     weights = "custom", weights_custom = w_subj,
                     verbose = FALSE)
```

## Meta engine via fmri_ttest with weights

The t-test interface supports the meta engine with equal/custom weighting.

```{r meta-engine}
fit_tt_meta <- fmrireg::fmri_ttest(
  gd_nii, formula = ~ 1 + group, engine = "meta", weights = "equal",
  sign = "BminusA"
)
```

```{r meta-engine-weighted}
w_subj <- rep(1, length(ids))
fit_tt_meta_w <- fmrireg::fmri_ttest(
  gd_nii, formula = ~ 1 + group, engine = "meta",
  weights = "custom", weights_custom = w_subj, sign = "BminusA"
)
```

```{r meta-engine-lancaster}
fit_tt_la <- fmrireg::fmri_ttest(
  gd_t, formula = ~ 1, engine = "meta",
  combine = "lancaster", weights = "custom",
  weights_custom = w_subj
)
```

```{r summarize-meta-engine-methods}
meta_engine_summary <- data.frame(
  fit = c("beta/SE, equal", "beta/SE, custom", "t-only Lancaster"),
  method = c(fit_tt_meta$method, fit_tt_meta_w$method, fit_tt_la$method),
  weights = c(fit_tt_meta$weights, fit_tt_meta_w$weights, fit_tt_la$weights),
  maximum_abs_z = c(
    max(abs(fit_tt_meta$z), na.rm = TRUE),
    max(abs(fit_tt_meta_w$z), na.rm = TRUE),
    max(abs(fit_tt_la$z), na.rm = TRUE)
  ),
  minimum_p = c(
    min(fit_tt_meta$p, na.rm = TRUE),
    min(fit_tt_meta_w$p, na.rm = TRUE),
    min(fit_tt_la$p, na.rm = TRUE)
  )
)
meta_engine_display <- transform(
  meta_engine_summary,
  maximum_abs_z = formatC(maximum_abs_z, digits = 3, format = "f"),
  minimum_p = format.pval(minimum_p, digits = 3, eps = 1e-4)
)
knitr::kable(
  meta_engine_display,
  row.names = FALSE,
  caption = "Requested meta engine and its resulting evidence summary"
)
```

```{r validate-meta-engine-methods, include=FALSE}
stopifnot(
  identical(fit_tt_meta$method, "pm"),
  identical(fit_tt_meta_w$weights, "custom"),
  identical(fit_tt_la$method, "combine:lancaster"),
  identical(fit_tt_la$combine, "lancaster"),
  fit_tt_la$n_features == prod(dim(space)),
  all(is.finite(meta_engine_summary$maximum_abs_z)),
  all(is.finite(meta_engine_summary$minimum_p))
)
```

## ROI t-only example (Stouffer and Lancaster)

You can also combine t-statistics at the ROI level from a tabular CSV. Provide
per-subject t and df, then choose a combine method.

```{r roi-t-only}
roi_t_df <- data.frame(
  subject = subjects,
  roi = "ExampleROI",
  t = rnorm(length(subjects), mean = 2.0, sd = 0.5),
  df = 40,
  stringsAsFactors = FALSE
)

gd_roi_t <- fmrireg::group_data(
  roi_t_df,
  format = "csv",
  effect_cols = c(t = "t", df = "df"),
  subject_col = "subject",
  roi_col = "roi"
)

fit_roi_st <- fmri_meta(gd_roi_t, formula = ~ 1, combine = "stouffer",
                        weights = "equal", verbose = FALSE)

w_roi <- rep(1, length(subjects))
fit_roi_la <- fmri_meta(gd_roi_t, formula = ~ 1, combine = "lancaster",
                        weights = "custom", weights_custom = w_roi,
                        verbose = FALSE)

roi_combination_summary <- data.frame(
  combination = c("Stouffer", "Lancaster"),
  method_metadata = c(fit_roi_st$method, fit_roi_la$method),
  weights = c(fit_roi_st$weights, fit_roi_la$weights),
  combined_z = c(fit_roi_st$coefficients[1, 1], fit_roi_la$coefficients[1, 1])
)
roi_combination_p <- 2 * pnorm(
  abs(roi_combination_summary$combined_z), lower.tail = FALSE
)
roi_combination_display <- transform(
  roi_combination_summary,
  combined_z = formatC(combined_z, digits = 3, format = "f"),
  two_sided_p = format.pval(roi_combination_p, digits = 3, eps = 1e-4)
)
knitr::kable(
  roi_combination_display,
  row.names = FALSE,
  caption = "T-only ROI combination results"
)
```

```{r validate-roi-combination-methods, include=FALSE}
stopifnot(
  identical(fit_roi_st$method, "combine:stouffer"),
  identical(fit_roi_la$method, "combine:lancaster"),
  all(is.finite(roi_combination_summary$combined_z))
)
```

## A brief recap

Meta-analysis in fmrireg supports fixed-effects and several random-effects estimators (`method = "fe"|"pm"|"dl"|"reml"`), with optional robust Huber weighting. You can pass subject-level covariates for group comparisons and, when working from t-maps only, set `combine =` to use Stouffer, Fisher, or Lancaster. Exact contrasts are available either at fit-time (via `contrasts=`) or post-hoc by saving packed covariance with `return_cov = "tri"` and then calling `contrast()`. Stouffer accepts equal or custom weights here; Fisher is equal-weighted; Lancaster uses its supplied weights as combination degrees of freedom. The examples above show ROI-based meta-regression, voxelwise fits from NIfTI, and t-only combinations through both `fmri_meta()` and `fmri_ttest(..., engine = "meta")`.
