title: "07. Group Analysis"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{07 Group Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=7, fig.height=5, message=FALSE, warning=FALSE)
```

```{r}
library(fmrireg)
library(ggplot2)
set.seed(123)
```

# Overview

This vignette demonstrates a simple ROI-based group analysis using fmrireg:

- Construct a group dataset from a small tabular data frame
- Fit fixed-effects and random-effects meta-analytic models
- Include a group covariate for a two-group comparison
- Extract and visualize results

The same workflow generalizes to voxelwise analysis by using HDF5/NIfTI inputs
via `group_data(format = "h5"|"nifti")`.

# Create a small ROI dataset

We simulate 10 subjects across two groups (A/B) for a single ROI. Group B has
an effect that is 1 unit larger than group A. All subjects have the same SE for
clarity.

```{r}
n_per_group <- 5
subjects <- sprintf("s%02d", 1:(2 * n_per_group))
group <- factor(rep(c("A", "B"), each = n_per_group))

# True effects: A = 0.5, B = 1.5 (difference = 1.0)
beta <- ifelse(group == "A", 0.5, 1.5)
se <- rep(0.25, length(beta))

roi_df <- data.frame(
  subject = subjects,
  roi = "ExampleROI",
  beta = beta,
  se = se,
  group = group,
  stringsAsFactors = FALSE
)

# Build group dataset (CSV/ROI format)
gd <- group_data(
  roi_df,
  format = "csv",
  effect_cols = c(beta = "beta", se = "se"),
  subject_col = "subject",
  roi_col = "roi",
  covariate_cols = c("group")
)

gd
```

# Fit group meta-analysis

We first fit an intercept-only model, then a model including a group term.

```{r}
# Intercept-only (grand mean across subjects)
fit_fe <- fmri_meta(gd, formula = ~ 1, method = "fe", verbose = FALSE)

# Intercept + group term (difference-coding for group B relative to A)
fit_cov <- fmri_meta(gd, formula = ~ 1 + group, method = "fe", verbose = FALSE)

print(fit_cov)
summary(fit_cov)
```

Note: With equal SE per subject, fixed-effects and random-effects will yield
similar point estimates. Random-effects (`method = "pm"`) will estimate between-
subject heterogeneity (`tau2`) when present.

```{r}
fit_pm <- fmri_meta(gd, formula = ~ 1 + group, method = "pm", verbose = FALSE)
summary(fit_pm)
```

# Extract coefficients and a contrast

```{r}
coef_names <- colnames(fit_cov$coefficients)
coef_names

# Intercept should be near 0.5, the group coefficient near 1.0
coef_est <- as.numeric(fit_cov$coefficients[1, ])
names(coef_est) <- coef_names
coef_est

# Build a simple contrast on the group term (if present)
if (any(grepl("group", coef_names))) {
  # Create a named weight vector that picks out the group coefficient
  w <- rep(0, length(coef_names)); names(w) <- coef_names
  w[grep("group", coef_names)] <- 1
  con <- contrast(fit_cov, w)
  con
}
```

# A quick visualization

We can visualize the group effects and their 95% CIs for the ROI-level fit.

```{r}
df_tidy <- tidy(fit_cov, conf.int = TRUE)
df_tidy

ggplot(df_tidy, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "ROI-level group meta-analysis",
       x = "Term", y = "Estimate ± 95% CI") +
  theme_minimal()
```

# Notes on voxelwise analysis

For voxelwise analysis, construct `group_data` with format `"h5"` or `"nifti"`:

```{r}
# HDF5 (produced via write_results.fmri_lm)
# gd_h5 <- group_data(h5_paths, format = "h5",
#                     subjects = subject_ids,
#                     contrast = "ContrastName",
#                     covariates = data.frame(group = group))

# NIfTI (provide per-subject paths for beta/SE or t)
# gd_nii <- group_data(list(beta = beta_paths, se = se_paths), format = "nifti",
#                      subjects = subject_ids,
#                      mask = "group_mask.nii.gz")

# fit <- fmri_meta(gd_h5, formula = ~ 1 + group, method = "pm")
# fit <- fmri_meta(gd_nii, formula = ~ 1, method = "fe")
```

For multiple comparisons correction that leverages spatial structure, see
`spatial_fdr()` and `create_3d_blocks()`.

## Minimal NIfTI Example (Reproducible)

This chunk creates tiny synthetic NIfTI volumes on disk (in a temp dir) for a
voxelwise demonstration. Group B has a higher effect in a small cube.

```{r}
library(neuroim2)

set.seed(42)
tmpdir <- tempdir()
space <- NeuroSpace(c(8, 8, 8), spacing = c(2, 2, 2))

n_per_group <- 3
ids <- sprintf("sub-%02d", 1:(2 * n_per_group))
grp <- factor(rep(c("A", "B"), each = n_per_group))

# Define a small active cube: x=3:5, y=3:5, z=3:5
active <- array(FALSE, dim = c(8, 8, 8))
active[3:5, 3:5, 3:5] <- TRUE

beta_paths <- character(length(ids))
se_paths   <- character(length(ids))

for (i in seq_along(ids)) {
  b <- array(0, dim = c(8, 8, 8))
  # Baseline effect in active region
  b[active] <- if (grp[i] == "A") 0.5 else 1.5
  # Small random noise per voxel (optional)
  b <- b + array(rnorm(length(b), sd = 0.05), dim = dim(b))
  v_beta <- NeuroVol(b, space)

  # Constant SE per voxel
  s <- array(0.25, dim = c(8, 8, 8))
  v_se <- NeuroVol(s, space)

  beta_paths[i] <- file.path(tmpdir, sprintf("%s_beta.nii.gz", ids[i]))
  se_paths[i]   <- file.path(tmpdir, sprintf("%s_se.nii.gz", ids[i]))
  write_vol(v_beta, beta_paths[i])
  write_vol(v_se,   se_paths[i])
}

# Mask covers all voxels
mask_path <- file.path(tmpdir, "mask.nii.gz")
write_vol(NeuroVol(array(1, dim = c(8, 8, 8)), space), mask_path)

# Build group_data and fit voxelwise meta-analysis
gd_nii <- group_data_from_nifti(
  beta_paths = beta_paths,
  se_paths   = se_paths,
  subjects   = ids,
  covariates = data.frame(group = grp),
  mask       = mask_path
)

fit_nii <- fmri_meta(gd_nii, formula = ~ 1 + group, method = "fe", verbose = FALSE)
fit_nii

# Get p-values for the group term and count discoveries (uncorrected)
group_col <- grep("group", colnames(fit_nii$coefficients))
pvals <- 2 * pnorm(-abs(fit_nii$coefficients[, group_col] / fit_nii$se[, group_col]))
sum(pvals < 0.05)

# Optionally, apply spatial FDR (group term), using simple blocks
sfr <- spatial_fdr(fit_nii, p = colnames(fit_nii$coefficients)[group_col], group = "blocks")
sum(sfr$reject)

# Reconstruct an image for the group effect estimate
img_group_est <- coef_image(fit_nii, colnames(fit_nii$coefficients)[group_col], statistic = "estimate")
range(as.array(img_group_est), na.rm = TRUE)
```

## Two-sample t-test (Welch and OLS) on NIfTI

As an alternative to meta-analysis, we can run two-sample voxelwise t-tests
directly on the per-subject beta maps, using either Welch's unequal-variance
test or a standard OLS/Student t-test via a simple design matrix.

```{r}
# Extract subject x voxel matrix of betas using the mask
dat <- read_nifti_full(gd_nii)
Y <- dat$beta  # S x P (subjects x voxels)

# Group indicator
g_int <- as.integer(grp)           # 1=A, 2=B
X <- cbind("(Intercept)" = 1,      # OLS design for Student t
          group = as.integer(grp == "B"))

# Welch two-sample t-test across features
w <- welch_t_cpp(Y, g_int)
t_welch <- as.numeric(w$t)   # length P
df_welch <- as.numeric(w$df)
p_welch <- 2 * pt(abs(t_welch), df = df_welch, lower.tail = FALSE)

# OLS (Student) two-sample t-test via design matrix
ols <- ols_t_cpp(Y, X)
# Handle missing dimnames on t matrix: use rowname if available, else fall back to 2nd row
rnames_t <- rownames(ols$t)
if (!is.null(rnames_t) && any(rnames_t == "group")) {
  t_ols <- ols$t["group", ]
} else {
  t_ols <- ols$t[2, ]
}
df_ols <- rep(ols$df, ncol(Y))
p_ols <- 2 * pt(abs(t_ols), df = df_ols, lower.tail = FALSE)

# Reconstruct quick image for Welch t (using the same mask/space)
timg_welch <- NeuroVol(array(NA_real_, dim = c(8, 8, 8)), space)
timg_welch[as.array(neuroim2::read_vol(mask_path)) > 0] <- t_welch
range(as.array(timg_welch), na.rm = TRUE)

# Count uncorrected significant voxels at alpha=0.05
sum(p_welch < 0.05)
sum(p_ols   < 0.05)
```
