---
title: Running one model across many subjects
author: Bradley R. Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Running one model across many subjects} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r load, message=FALSE}
library(fmrireg)
```

## The problem

A common study runs the **same model** across many subjects. The design is
constant -- the same conditions, HRF, baseline, contrasts, and fitting options --
but the data changes per subject: different files, different event order,
different run lengths. Looping `fmri_lm()` by hand is tedious and does not lend
itself to parallel or cluster execution.

`fmrireg` separates the part that is **invariant** (an `fmri_template`) from the
part that **varies per subject** (a *binding*: that subject's data). Combining
the two produces a small, serializable `fmri_job` -- a recipe that holds no voxel
data and can be shipped to a worker (a `future`, or a node in an array job) that
reconstructs the dataset, fits, and reduces.

The four pieces:

| Piece | What it is |
|---|---|
| `fmri_template()` | the invariant model: formula, baseline spec, contrasts, fit control, reducer |
| a *binding* | one subject's data: scans, run lengths, TR, events, confounds |
| `fmri_job` | template + one binding -- a serializable recipe (built by `instantiate()`) |
| a *reducer* | what each fitted subject becomes (a tidy table, or files on disk) |

This closes a loop: BIDS in (via `bidser`) → per-subject fit → BIDS-keyed maps
out (`write_results`) → group level (`collect_results` → `fmri_meta`).

## A minimal, runnable example

Define the template once, describe two subjects as in-memory bindings, then fan
the model out. (Real studies use file paths; in-memory `matrix_dataset` bindings
keep this example fast.)

```{r minimal}
# one subject's data as a binding
make_subject <- function(id, seed) {
  set.seed(seed)
  runs <- c(40L, 40L); TR <- 2
  Y <- matrix(rnorm(sum(runs) * 5), sum(runs), 5)        # timepoints x voxels
  ev <- do.call(rbind, lapply(1:2, function(r)
    data.frame(onset = seq(4, 60, by = 8),
               condition = factor(rep(c("A", "B"), length.out = 8)),
               run = r)))
  list(id = id, scans = Y, TR = TR, run_length = runs, events = ev)
}
subjects <- list(make_subject("sub-01", 1), make_subject("sub-02", 2))

# the invariant model, defined once
tmpl <- fmri_template(
  onset ~ hrf(condition), ~ run,
  baseline = baseline_spec(degree = 3),
  reducer  = reduce_betas()        # each subject -> a tidy beta table
)

# bind data -> jobs -> run
jobs <- instantiate(tmpl, subjects)
res  <- run_jobs(jobs)
res

values <- batch_values(res)        # named by job id
head(values[["sub-01"]])
```

`run_jobs()` isolates per-job failures: a subject that errors is recorded in the
result rather than aborting the batch (see `batch_errors()`).

## Reducers: what each subject becomes

A reducer runs on the worker, right after the fit, so only the reduced output
crosses the worker→driver (or worker→disk) boundary -- not a whole fitted model.

| Reducer | Output |
|---|---|
| `reduce_identity()` | the entire `fmri_lm` object (largest) |
| `reduce_betas()` | tidy data frame: `job_id, term, voxel, estimate, se, stat` |
| `reduce_contrasts()` | tidy data frame of fitted contrasts |
| `reduce_write_results()` | writes BIDS-keyed maps to disk, returns the paths |

You can also pass your own `function(fit, job)`; keep it a top-level/package
function so it serializes to a worker.

## Discovering subjects from BIDS

For BIDS-formatted data, `from_bids()` (which uses the `bidser` package) builds
the per-subject bindings -- scans, events, confounds, TR, and run lengths -- so
you do not assemble them by hand.

```{r from-bids, eval=FALSE}
proj <- bidser::bids_project("study/", fmriprep = TRUE)
mani <- from_bids(
  proj, task = "stroop", space = "MNI152NLin2009cAsym",
  confounds = bidser::confound_set("motion6"),
  mask = "study/derivatives/.../space-MNI152NLin2009cAsym_desc-brain_mask.nii.gz"
)
jobs <- instantiate(tmpl, mani)
```

The design formula's variables must match the columns of the BIDS
`events.tsv` (e.g. `trial_type`); a `run` column is added for the block
structure.

## Preflight before you fan out

`preflight()` validates jobs on the driver -- design columns present, TR and run
lengths consistent, confound dimensions correct -- so problems surface in
seconds rather than on a compute node after the queue drains.

```{r preflight}
preflight(jobs)
```

## Running: locally, in parallel, or on a cluster

Sequential is the default. For parallelism, set a `future` plan; `run_jobs()`
dispatches through it -- including `future.batchtools` cluster plans -- with **no
scheduler-specific code** in `fmrireg`.

```{r parallel, eval=FALSE}
library(future)
plan(multisession, workers = 4)        # or future.batchtools::batchtools_slurm
res <- run_jobs(jobs, parallel = TRUE)
```

For an array scheduler, export the jobs to disk and drive them with whatever
system you have:

```{r export, eval=FALSE}
export_jobs(jobs, "study/jobs")
# study/jobs/manifest.rds + a backend-agnostic run_one.R
# then, per array task:
#   SLURM: Rscript study/jobs/run_one.R $SLURM_ARRAY_TASK_ID
#   local: for i in $(seq 1 N); do Rscript study/jobs/run_one.R $i; done
```

`run_one.R` reads the manifest, reconstructs job `i`, fits, and writes its
reduced output. The job is a portable recipe, so the worker need only have
`fmrireg` installed.

## Closing the loop: the group level

With `reduce_write_results()`, each worker writes BIDS-keyed statistical maps.
`collect_results()` gathers them back into a `group_data` object that
`fmri_meta()` meta-analyses. (Writing both `beta` and `se` requires a contrast in
the model, so the group step has per-subject variance.)

```{r group, eval=FALSE}
con <- contrast_set(pair_contrast(~ trial_type == "incongruent",
                                  ~ trial_type == "congruent",
                                  name = "incong_gt_cong"))
tmpl <- fmri_template(
  onset ~ hrf(trial_type, contrasts = con), ~ run,
  baseline = baseline_spec(degree = 3, confounds = bidser::confound_set("motion6")),
  reducer  = reduce_write_results(format = "nifti", stats = c("beta", "se"),
                                  path = "study/glm")
)
run_jobs(instantiate(tmpl, mani))

gd <- collect_results("study/glm", space = "MNI152NLin2009cAsym")
fm <- fmri_meta(gd, ~ 1, method = "fe")
```

## Summary

- Define the model once with `fmri_template()`; vary only the data.
- `instantiate()` produces serializable `fmri_job` recipes; `from_bids()`
  populates them from a BIDS dataset.
- `preflight()` catches problems before fan-out.
- `run_jobs()` runs sequentially, in parallel via `future`, or through a custom
  backend; `export_jobs()` emits a scheduler-agnostic runner for array jobs.
- Reducers keep per-subject output compact; `collect_results()` →
  `fmri_meta()` close the loop at the group level.
