title: "01. Overview and Workflow"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01 Overview and Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=7, fig.height=5, message=FALSE, warning=FALSE)
library(fmrireg)
library(fmrihrf)
```

## Introduction

The `fmrireg` package provides tools for analyzing fMRI regression models in R, inspired by standard R modeling tools like `lm()` and `glm()`. It integrates hemodynamic response functions and design matrix construction for comprehensive fMRI analysis.

While fMRI analysis involves complexities like large datasets, spatio-temporal correlations, and physiological noise, the core of standard univariate analysis is multiple regression applied voxelwise. The `fmrireg` package simplifies the specification and estimation of these regression models.

This vignette provides a high-level overview of the key steps and components involved in using `fmrireg`.

## Key Components and Workflow

A typical fMRI analysis workflow with `fmrireg` involves these main stages:

1.  **Data Representation:** Defining how your fMRI data and experimental design are organized.
2.  **Model Building:** Specifying the regression model using formulas, including:
    *   Defining Hemodynamic Response Functions (HRFs).
    *   Creating event-related regressors (`event_model`).
    *   Modeling baseline components like drift and nuisance factors (`baseline_model`).
    *   Defining contrasts of interest (`contrast`).
3.  **Model Estimation:** Fitting the specified model to the data (e.g., using `fmri_lm`).
4.  **Advanced Topics:** Exploring single-trial estimation, simulation, or translation to other packages.

### 1. Data Representation

Before modeling, you need to represent your fMRI data and its structure. `fmrireg` uses objects like:

*   `fmri_dataset` / `fmri_mem_dataset`: Encapsulates the 4D fMRI data (potentially file-backed or in-memory), mask information, TR, run structure, and the experimental design table.
*   `sampling_frame`: Defines the temporal structure – the number of scans per run (`blocklens`) and the repetition time (`TR`).

```{r sampling_frame_example}
# Example: 5 runs, 200 scans/run, TR=2s
sframe <- sampling_frame(blocklens = rep(200, 5), TR = 2)
sframe
```

### 2. Model Building

#### a) Hemodynamic Response Function (HRF)

The HRF models the BOLD signal response to a brief neural event. `fmrireg` provides various pre-defined HRFs (e.g., `HRF_SPMG1`, `HRF_GAUSSIAN`) and functions to create or modify them (`gen_hrf`, `hrf_blocked`, `hrf_lagged`).

```{r hrf_example}
# Plot the SPM canonical HRF
time <- seq(0, 24, by = 0.2)
plot(time, HRF_SPMG1(time), type = 'l', xlab = "Time (s)", ylab = "BOLD Response",
     main = "SPM Canonical HRF (HRF_SPMG1)")
```

*   **Related Topic:** Hemodynamic Response Functions

#### b) Event Model (`event_model`)

This defines the task-related part of your GLM. You specify how experimental events (onsets, conditions, durations) map to regressors using a formula syntax. The `hrf()` or `afni_hrf()` functions within the formula link event variables to specific HRF shapes.

```{r event_model_example, eval=FALSE}
# Example: Model effect of 'stim_type', using SPMG1 HRF
# (Assuming 'my_design_table' and 'my_sampling_frame' exist)
event_mod <- event_model(onset ~ hrf(stim_type), 
                         data = my_design_table, 
                         block = ~ run_id, 
                         sampling_frame = my_sampling_frame)
```

*   **Related Topic:** Event Models

#### c) Baseline Model (`baseline_model`)

This models non-task-related components like scanner drift, run intercepts, and other nuisance factors (e.g., motion parameters).

```{r baseline_model_example}
# Example: Cubic B-spline drift model + run intercepts
# (Requires 'sframe' defined earlier)
base_mod <- baseline_model(basis = "bs", degree = 3, sframe = sframe)
print(base_mod)
```

*   **Related Topic:** Baseline Models

#### d) Contrasts (`pair_contrast`, `unit_contrast`, etc.)

Contrasts define specific comparisons between condition estimates.

```{r contrast_example}
# Example: Contrast comparing 'face' vs 'scene' levels of 'stim_type'
face_vs_scene <- pair_contrast(~ stim_type == "face", ~ stim_type == "scene", name = "FvS")
print(face_vs_scene)
```

*   **See Vignette:** `vignette("a_05_contrasts", package = "fmrireg")`

### 3. Model Estimation (`fmri_model`, `fmri_lm`)

You combine the event and baseline models into a full `fmri_model` and then fit it to your `fmri_dataset` using estimation functions like `fmri_lm` (for standard GLM) or `estimate_betas` (for single-trial betas).

```{r fmri_model_example, eval=FALSE}
# Combine event and baseline models
full_model <- fmri_model(event_mod, base_mod)

# Fit the GLM
# (Requires 'my_dataset' of class fmri_dataset)
fit <- fmri_lm(full_model, dataset = my_dataset)

# Robust fit down-weighting outlier time points
fit_robust <- fmri_lm(full_model, dataset = my_dataset, robust = TRUE)

# Extract contrast results (if contrasts were defined)
# con_results <- coef(fit)  # Get coefficients
```

Setting `robust = TRUE` invokes row-wise weighting to lessen the influence of
motion spikes or other frame-level artifacts.

*   **See Vignettes:**
    *   `vignette("a_09_linear_model", package = "fmrireg")`
    *   Linear Models and Trialwise Analysis

### 4. Advanced Topics

`fmrireg` also supports:

*   **Simulation:** Creating artificial fMRI data (`simulate_fmri_matrix`).
    *   `vignette("a_08_simulation", package = "fmrireg")`
*   **AFNI Translation:** Generating scripts for AFNI's `3dDeconvolve` (`afni_lm`).
*   **Low-level Regressors:** Manually creating and manipulating regressors (`regressor`).

## Getting Started

This overview provides a map of the `fmrireg` package. For detailed explanations and examples, please refer to the specific vignettes linked above.

## Next

- 02 Data Structures and Sampling Frames
- 03 Simulating fMRI Data
