---
title: "Overview"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Overview of the `fmrireg` package

The analysis of Functional Magnetic Resonance Imaging (fMRI) has a lot of complexities due to the large amount of data, the fact that the data is in the form of multi-dimensional images (usually, 4-dimensional -- x,y,z, and time), the fact that the hemodynamic response is sluggish and delayed (4-6s) with respect to stimulus presentation or task events, the existence of physiological noise and other "nuisance" signals, issues related to multiple comparisons, and the existence of spatially and temporally auto-correlated errors in the signal.

However, if we push all that complexity to the side for a moment, standard "univariate" fMRI analysis involves simple multiple regression analysis applied to every 3d location ("voxel") in the brain. The goal of the `fmrireg` package is to provide a convenient interface, in the spirit of other R regression modeling tools, for analyzing fMRI data. We therefore use a R's formula interface to specify regression models in a way that makes for convenient and readable code.

## The building blocks of an fMRI regression model

The basic, or lowest level, unit of an fMRI regression model is the **hemodynamic reesponse function** (HRF). The HRF is a model of the mapping, or transfer function, between an external stimulus (or neural event) and a BOLD signal. A commonly used HRF is the so-called SPM canonical HRF:


```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
library(fmrireg)
```

```{r}

time <- seq(0,24, by=.2)
plot(time, hrf_spmg1(time), type='l', xlab="time", ylab="BOLD activity")

```

Currently, `fmrireg` provides several builtin HRFs that can be used to model BOLD activity. See the vignette "Hemodynamic Response Functions" for a detailed overview. Choosing the correct HRF depends on the use case and aims of the analysis, but the SPM canaonical function is usually a good default choice.

## Regressors

A "regressor" is simply an independent variable in a multiple regression model. In fMRI analysis, however, regressors are constructed by "convolving" a set of onset times--the vector of times at which certain task events occur--with a chosen HRF. Imagine we have a stimulus that is presented every 16 seconds with 8 repetitions. We can construct a "regressor" by convolving the chosen HRF with a vector onset times (in seconds), whereby each element of the vector corresponds to time at which a stimulus occurs.

Below we use the function `gen_hrf` to create a special function that has the class "HRF". An alternative is to use the pre-made HRF objects, for example, `HRF_SPMG1` if we want to use th SPM canonical function. Once we create a regressor with the `regressor` constructor function, we can evaluate it at any arbitrary time points. In other words, just as the HRFs are functions of time, so too are `regressor`s. The difference is that `regressor`, when evaluated at time *t* sum over all the constituent HRFs, the domain of which are determined by the onset vector.

In the code below, we evaluate the regressor at every time point between 1 and 140.

```{r}

onsets <- seq(0, 16*8, by=16)
hrf <- gen_hrf(hrf_spmg1)
reg <- regressor(onsets, hrf=hrf)

grid <- seq(0,140, by=1)
plot(grid, evaluate(reg, grid), type='l', ylab="BOLD activity", xlab="time", main="ISI = 16")
## same as
## reg <- regressor(onsets, hrf=HRF_SPMG1)

```

To better illustrate how a regressor sums over it's constiturent HRFs, we can place the onsets closer together, so that the HRFs overlap in time:

```{r}

onsets <- seq(0, 16*8, by=6)
hrf <- gen_hrf(hrf_spmg1)
reg <- regressor(onsets, hrf=hrf)

grid <- seq(0,140, by=1)
plot(grid, evaluate(reg, grid), type='l', ylab="BOLD activity", xlab="time", main="ISI = 6")


```

In `fmrireg` it is generally not necessary to manually create regressors as we have done above. Rather, regressors are created "behind the scenes" by the higher level modeling functions, such as `event_model`. However, the facility to create individual regressors, for the purposes of learning fMRI analysis or in conducting special analyses, may still be useful.

## Event Models

An HRF is a function from time to BOLD activity. A regressor is a function of time that sums up BOLD activity over a list of HRFs (one per "onset"). In conventional fMRI analysis, a regressor refers to a single column of a design matrix. For example, if we have two task conditions, "face" and "scene", we would generate one face regressor and one scene regressor. However, in statistical modeling of experiments, we generally think not in terms of single columns of a design matrix, but rather in terms of the "factors" 
















