---
title: Plugin Development
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Plugin Development} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

library(fmrireg)
set.seed(42)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


## Why use the plugin API?

The plugin API is for advanced users who want to extend `fmrireg` without
forking the package. Two extension points are available:

- `register_basis()` adds new basis names that can be used inside `hrf(...)`.
- `register_engine()` adds new fitting engines that still return standard
  `fmri_lm` objects.

This vignette shows how to register both kinds of extensions, how engine
capabilities interact with global `fmri_lm()` options, and how to inspect the
engine metadata that drives dispatch.

## How do you build a minimal dataset?

We will work with a small matrix-backed dataset so every example is fast and
fully reproducible.

```{r make-dataset}
n_time <- 80L
n_voxels <- 24L

events <- data.frame(
  onsets = c(8, 20, 32, 44, 56),
  condition = factor(c("A", "B", "A", "B", "A")),
  run = 1L
)

Y <- matrix(rnorm(n_time * n_voxels), nrow = n_time, ncol = n_voxels)
dset <- fmridataset::matrix_dataset(
  Y,
  TR = 2,
  run_length = n_time,
  event_table = events
)

stopifnot(inherits(dset, "matrix_dataset"))
dim(fmridataset::get_data_matrix(dset))
```

## How do you register a custom basis?

`register_basis()` maps a string name to a constructor. That name can then be
used inside `hrf(...)` just like built-in bases.

```{r register-basis}
basis_name <- "vignette_bspline_basis"

register_basis(
  basis_name,
  function(span = 18, ...) {
    fmrihrf::gen_hrf(fmrihrf::hrf_bspline, N = 5, span = span)
  }
)

model_with_basis <- create_fmri_model(
  formula = onsets ~ hrf(condition, basis = basis_name, span = 16),
  block = ~run,
  dataset = dset
)

stopifnot(inherits(model_with_basis, "fmri_model"))
head(colnames(design_matrix(model_with_basis)))
```

The important rule is that the constructor must return an object that
`fmrihrf` understands as a valid basis.

## How do you register a custom engine?

An engine receives `(model, dataset, args, cfg)` and should return an
`fmri_lm` object. In many cases the easiest path is:

1. extract or transform the response matrix,
2. call one of the helper constructors,
3. return the resulting `fmri_lm` object.

The example below centers each voxel time series before fitting and then
delegates to `fit_glm_on_transformed_series()`.

```{r register-engine}
engine_name <- "vignette_centered_engine"

register_engine(
  name = engine_name,
  capabilities = list(
    robust = FALSE,
    preprocessing = FALSE
  ),
  fit = function(model, dataset, args, cfg) {
    Y_raw <- as.matrix(fmridataset::get_data_matrix(dataset))
    Y_centered <- if (isTRUE(args$center)) {
      sweep(Y_raw, 2, colMeans(Y_raw), FUN = "-")
    } else {
      Y_raw
    }

    fit_glm_on_transformed_series(
      model,
      Y_centered,
      cfg = cfg,
      dataset = dataset,
      engine = engine_name,
      strategy = "engine"
    )
  }
)

fit_plugin <- fmri_lm(
  onsets ~ hrf(condition, basis = basis_name, span = 16),
  block = ~run,
  dataset = dset,
  engine = engine_name,
  engine_args = list(center = TRUE),
  robust = FALSE,
  robust_options = list(max_iter = 10L)
)

stopifnot(inherits(fit_plugin, "fmri_lm"))
stopifnot(identical(attr(fit_plugin, "engine"), engine_name))
stopifnot(identical(attr(fit_plugin, "strategy"), "engine"))

list(
  engine = attr(fit_plugin, "engine"),
  strategy = attr(fit_plugin, "strategy"),
  n_beta_rows = nrow(fit_plugin$result$betas)
)
```

## Which helper should you use?

Use `fit_glm_on_transformed_series()` when your engine has already produced a
time-by-voxel matrix and you want standard OLS-style inference.

Use `fit_glm_with_config()` when your engine still wants the integrated
AR/robust solver applied to a transformed response matrix.

Use `fit_glm_from_suffstats()` when your engine works from streamed
cross-products and never materializes the transformed series directly.

## How do capabilities affect the executed configuration?

The central dispatcher validates unsupported options before engine code runs.
It also keeps two configurations on the fitted object:

- `requested_config`: what the caller supplied to `fmri_lm()`
- `executed_config`: the engine-scoped configuration that the engine actually
  received after unsupported sections were normalized away

```{r config-metadata}
requested_cfg <- attr(fit_plugin, "requested_config")
executed_cfg <- attr(fit_plugin, "executed_config")
value_or_na <- function(x) if (is.null(x) || length(x) == 0) NA else x

data.frame(
  config = c("requested", "executed"),
  robust_type = c(
    as.character(value_or_na(requested_cfg$robust$type)),
    as.character(value_or_na(executed_cfg$robust$type))
  ),
  robust_max_iter = c(
    value_or_na(requested_cfg$robust$max_iter),
    value_or_na(executed_cfg$robust$max_iter)
  ),
  normalized = c(
    FALSE,
    !isTRUE(all.equal(
      value_or_na(requested_cfg$robust$max_iter),
      value_or_na(executed_cfg$robust$max_iter)
    ))
  )
)
```

If a caller enables an unsupported feature such as `robust = TRUE`, the engine
is rejected before `fit()` is called.

```{r capability-error-check, include = FALSE}
unsupported_message <- NULL
tryCatch(
  fmri_lm(
    onsets ~ hrf(condition),
    block = ~run,
    dataset = dset,
    engine = engine_name,
    robust = TRUE
  ),
  error = function(e) {
    unsupported_message <<- conditionMessage(e)
  }
)

stopifnot(grepl("does not support robust fitting", unsupported_message))
```

```{r capability-error}
unsupported_message
```

## How can you inspect engine specs?

Built-in and plugin engines share the same spec object. The exported accessors
`engine_spec()` and `engine_specs()` give you the metadata you need without
touching internal registries.

```{r inspect-engine-specs}
spec_row <- function(name) {
  spec <- engine_spec(name)
  data.frame(
    name = spec$name,
    source = spec$source,
    strategy = spec$strategy,
    aliases = paste(spec$aliases, collapse = ", "),
    robust = spec$capabilities$robust,
    preprocessing = spec$capabilities$preprocessing,
    stringsAsFactors = FALSE
  )
}

rbind(
  spec_row(engine_name),
  spec_row("rrr_gls")
)
```

The full registry is available as a named list of spec objects.

```{r list-engine-specs}
sort(names(engine_specs()))
```

## What should extension authors remember?

- Keep engine inputs small and explicit: `model`, `dataset`, `args`, `cfg`.
- Declare unsupported global features in `capabilities` so the dispatcher can
  reject them before your engine runs.
- Use `requested_config` versus `executed_config` when you need to explain what
  was requested versus what your engine actually used.
- Prefer the helper constructors over hand-building `fmri_lm` objects unless
  your engine truly needs custom result assembly.

```{r cleanup, include = FALSE}
if (exists(basis_name, envir = fmrireg:::.fmrireg_basis_registry, inherits = FALSE)) {
  rm(list = basis_name, envir = fmrireg:::.fmrireg_basis_registry)
}
if (exists(engine_name, envir = fmrireg:::.fmrireg_engine_registry, inherits = FALSE)) {
  rm(list = engine_name, envir = fmrireg:::.fmrireg_engine_registry)
}
```
