---
title: Advanced Plugin Development
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Advanced Plugin Development} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

library(fmrireg)
set.seed(42)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


## Why use the plugin API?

The plugin API is for advanced users who want to extend `fmrireg` without
forking the package. Two extension points are available:

- `register_basis()` adds new basis names that can be used inside `hrf(...)`.
- `register_engine()` adds new fitting engines that still return standard
  `fmri_lm` objects.

This vignette shows how to register both kinds of extensions, how engine
capabilities interact with global `fmri_lm()` options, and how to inspect the
engine-spec metadata that drives dispatch.

## How do you build a minimal dataset?

We will work with a small matrix-backed dataset so every example is fast and
fully reproducible.

```{r make-dataset}
n_time <- 80L
n_voxels <- 24L

events <- data.frame(
  onsets = c(8, 20, 32, 44, 56),
  condition = factor(c("A", "B", "A", "B", "A")),
  run = 1L
)

Y <- matrix(rnorm(n_time * n_voxels), nrow = n_time, ncol = n_voxels)
dset <- fmridataset::matrix_dataset(
  Y,
  TR = 2,
  run_length = n_time,
  event_table = events
)

stopifnot(inherits(dset, "matrix_dataset"))
dim(fmridataset::get_data_matrix(dset))
```

## How do you register a custom basis?

`register_basis()` maps a string name to a constructor. That name can then be
used inside `hrf(...)` just like built-in bases.

```{r register-basis}
basis_name <- "vignette_bspline_basis"

register_basis(
  basis_name,
  function(span = 18, ...) {
    fmrihrf::gen_hrf(fmrihrf::hrf_bspline, N = 5, span = span)
  }
)

model_with_basis <- create_fmri_model(
  formula = onsets ~ hrf(condition, basis = basis_name, span = 16),
  block = ~run,
  dataset = dset
)

stopifnot(inherits(model_with_basis, "fmri_model"))
head(colnames(design_matrix(model_with_basis)))
```

The important rule is that the constructor must return an object that
`fmrihrf` understands as a valid basis.

## How do you register a custom engine?

An engine receives `(model, dataset, args, cfg)` and should return an
`fmri_lm` object. In many cases the easiest path is:

1. extract or transform the response matrix,
2. call one of the helper constructors,
3. return the resulting `fmri_lm` object.

The example below centers each voxel time series before fitting and then
delegates to `fit_glm_on_transformed_series()`.

```{r register-engine}
engine_name <- "vignette_centered_engine"

register_engine(
  name = engine_name,
  capabilities = list(
    robust = FALSE,
    preprocessing = FALSE
  ),
  fit = function(model, dataset, args, cfg) {
    Y_raw <- as.matrix(fmridataset::get_data_matrix(dataset))
    Y_centered <- if (isTRUE(args$center)) {
      sweep(Y_raw, 2, colMeans(Y_raw), FUN = "-")
    } else {
      Y_raw
    }

    fit_glm_on_transformed_series(
      model,
      Y_centered,
      cfg = cfg,
      dataset = dataset,
      engine = engine_name,
      strategy = "engine"
    )
  }
)

fit_plugin <- fmri_lm(
  onsets ~ hrf(condition, basis = basis_name, span = 16),
  block = ~run,
  dataset = dset,
  engine = engine_name,
  engine_args = list(center = TRUE),
  robust = FALSE,
  robust_options = list(max_iter = 10L)
)

stopifnot(inherits(fit_plugin, "fmri_lm"))
stopifnot(identical(attr(fit_plugin, "engine"), engine_name))
stopifnot(identical(attr(fit_plugin, "strategy"), "engine"))

list(
  engine = attr(fit_plugin, "engine"),
  strategy = attr(fit_plugin, "strategy"),
  n_beta_rows = nrow(fit_plugin$result$betas)
)
```

## Which helper should you use?

Use `fit_glm_on_transformed_series()` when your engine has already produced a
time-by-voxel matrix and you want standard OLS-style inference.

Use `fit_glm_with_config()` when your engine still wants the integrated
AR/robust solver applied to a transformed response matrix.

Use `fit_glm_from_suffstats()` when your engine works from streamed
cross-products and never materializes the transformed series directly.

## How do capabilities affect the executed configuration?

The central dispatcher validates unsupported options before engine code runs.
It also keeps two configurations on the fitted object:

- `requested_config`: what the caller supplied to `fmri_lm()`
- `executed_config`: the engine-scoped configuration actually passed into the
  engine after unsupported sections were normalized away. In the current
  implementation, subordinate robust-tuning values such as `max_iter` may be
  reset even when the caller already requested `robust = FALSE`.

```{r config-metadata}
requested_cfg <- attr(fit_plugin, "requested_config")
executed_cfg <- attr(fit_plugin, "executed_config")

cfg_value <- function(cfg, section, field) {
  if (is.null(cfg) || is.null(cfg[[section]]) || is.null(cfg[[section]][[field]])) {
    return(NA)
  }
  cfg[[section]][[field]]
}

requested_max_iter <- cfg_value(requested_cfg, "robust", "max_iter")
executed_max_iter <- cfg_value(executed_cfg, "robust", "max_iter")

data.frame(
  config = c("requested", "executed"),
  robust_type = c(
    as.character(cfg_value(requested_cfg, "robust", "type")),
    as.character(cfg_value(executed_cfg, "robust", "type"))
  ),
  robust_max_iter = c(
    requested_max_iter,
    executed_max_iter
  ),
  normalized = c(
    FALSE,
    !isTRUE(all.equal(requested_max_iter, executed_max_iter))
  )
)
```

If a caller enables an unsupported feature such as `robust = TRUE`, the engine
is rejected before `fit()` is called.

```{r capability-error-check, include = FALSE}
unsupported_message <- NULL
tryCatch(
  fmri_lm(
    onsets ~ hrf(condition),
    block = ~run,
    dataset = dset,
    engine = engine_name,
    robust = TRUE
  ),
  error = function(e) {
    unsupported_message <<- conditionMessage(e)
  }
)

stopifnot(grepl("does not support robust fitting", unsupported_message))
```

```{r capability-error}
unsupported_message
```

## How can you inspect engine specs?

Built-in and plugin engines now share the same spec object. The read-only
accessors `engine_spec()` and `engine_specs()` expose that metadata without
requiring access to internal registries. If you are working against an older
installed build where those helpers are not yet available, you can still fall
back to the registry names for a lightweight diagnostic.

```{r inspect-engine-specs}
has_public_specs <- exists("engine_spec", envir = asNamespace("fmrireg"), inherits = FALSE)

if (has_public_specs) {
  plugin_spec <- get("engine_spec", envir = asNamespace("fmrireg"))(engine_name)
  builtin_spec <- get("engine_spec", envir = asNamespace("fmrireg"))("rrr_gls")

  print(plugin_spec)
  print(builtin_spec)
} else {
  registry <- get(".fmrireg_engine_registry", envir = asNamespace("fmrireg"))
  data.frame(
    name = c(engine_name, "rrr_gls"),
    registered = c(
      exists(engine_name, envir = registry, inherits = FALSE),
      exists("rrr_gls", envir = registry, inherits = FALSE)
    )
  )
}
```

The full registry is also available as a list of specs.

```{r list-engine-specs}
if (exists("engine_specs", envir = asNamespace("fmrireg"), inherits = FALSE)) {
  spec_names <- names(get("engine_specs", envir = asNamespace("fmrireg"))())
} else {
  registry <- get(".fmrireg_engine_registry", envir = asNamespace("fmrireg"))
  spec_names <- sort(ls(registry, all.names = TRUE))
}

spec_names
```

## What should extension authors remember?

- Keep engine inputs small and explicit: `model`, `dataset`, `args`, `cfg`.
- Declare unsupported global features in `capabilities` so the dispatcher can
  reject them before your engine runs.
- Use `requested_config` versus `executed_config` when you need to explain what
  was requested versus what your engine actually used.
- Prefer the helper constructors over hand-building `fmri_lm` objects unless
  your engine truly needs custom result assembly.

```{r cleanup, include = FALSE}
if (exists(basis_name, envir = fmrireg:::.fmrireg_basis_registry, inherits = FALSE)) {
  rm(list = basis_name, envir = fmrireg:::.fmrireg_basis_registry)
}
if (exists(engine_name, envir = fmrireg:::.fmrireg_engine_registry, inherits = FALSE)) {
  rm(list = engine_name, envir = fmrireg:::.fmrireg_engine_registry)
}
```
