---
title: "Translating to AFNI 3dDeconvolve"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translating to AFNI 3dDeconvolve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  error = TRUE # Allow errors to be displayed for AFNI-specific checks
)
library(tibble)
library(fmrireg)
# Check if AFNI is available - needed for AFNI-specific HRFs
# You might replace this with a more robust check, e.g., system("which 3dDeconvolve")
afni_available <- FALSE # Set to TRUE if AFNI is installed and in PATH
```

## Introduction: Generating AFNI Models with `fmrireg`

While `fmrireg` provides tools for fitting fMRI models directly in R (see `fmri_lm`), you might prefer to leverage the robust and efficient estimation methods available in other popular fMRI analysis packages like AFNI.

This vignette demonstrates how to construct a full fMRI regression model within `fmrireg` and then translate it into the syntax and associated files required for AFNI's `3dDeconvolve` program. This allows you to use `fmrireg`'s flexible formula interface and HRF specification tools while utilizing AFNI for the final voxelwise model estimation.

The primary function for this translation is `afni_lm()`.

## A Simple Model Example

Let's start with the simple single-run, single-factor design from the "Building Event Models" vignette.

```{r setup_simple_afni}
TR <- 2
cond <- c("face", "scene", "tool", "object")
NSTIM <- length(cond) * 4

set.seed(123)
simple_design <- data.frame(
  stim = factor(sample(rep(cond, 4))),
  ISI = sample(10:20, NSTIM, replace = TRUE), # Using wider ISI from previous fixes
  run = rep(1, NSTIM),
  trial = factor(1:NSTIM)
)

simple_design$onset <- cumsum(c(0, simple_design$ISI[-NSTIM] + 2))
sframe <- sampling_frame(blocklens = 140, TR = TR)

# Define a contrast
con1 <- pair_contrast(~ stim == "face", ~stim == "scene", name = "face_scene")

# Create the event model
emodel <- event_model(onset ~ hrf(stim, contrasts = con1), 
                     data = simple_design, 
                     block = ~ run, 
                     sampling_frame = sframe)

# Create the baseline model (B-spline drift)
bmodel <- baseline_model(basis = "bs", degree = 5, sframe = sframe)

# Combine into a full fmri_model
fmodel <- fmri_model(emodel, bmodel)

# Define the dataset (using placeholder file names)
dset <- fmri_dataset(scans = "scan01.nii.gz",
                     mask = "mask.nii.gz",
                     TR = TR,
                     run_length = 140,
                     event_table = simple_design,
                     base_path = ".") # Assuming files are in the working directory
```

## Translating the Model with `afni_lm`

The `afni_lm()` function takes the `fmri_model` and `fmri_dataset` objects and generates an AFNI model specification.

```{r create_afni_spec}
alm_spec <- afni_lm(fmodel, dset)
print(alm_spec)
```

This creates an `afni_lm_spec` object which contains:

*   The original `fmri_model` and `fmri_dataset`.
*   The generated `3dDeconvolve` command string (`alm_spec$cmd$cmd`).
*   Parsed command line arguments (`alm_spec$cmd$cmdlines`).
*   Information about stimulus timing files, GLTs, etc.

**Key Translations:**

*   **HRF Convolution:** `fmrireg` performs the HRF convolution specified in `event_model` (e.g., `hrf(stim)` uses `HRF_SPMG1`) and passes the resulting regressors to `3dDeconvolve` via `-stim_file` arguments. Each condition level gets its own stimulus file (e.g., `stim[face]_reg.1D`).
*   **Baseline Model:** Regressors from the `baseline_model` (here, the B-spline basis columns) are included as nuisance regressors via the `-ortvec` argument.
*   **Contrasts:** `fmrireg` contrasts (like `con1` defined with `pair_contrast`) are automatically converted into AFNI-style general linear tests (GLTs) using the `-gltsym` and `-glt_label` arguments. A separate text file is generated for each GLT (e.g., `GLT_face_scene.txt`).

## Handling AFNI's `-polort`

By default, `afni_lm` sets `polort = -1` in the `3dDeconvolve` command. This disables AFNI's built-in polynomial baseline drift modeling. This is the desired behavior when you provide your own drift regressors via `fmrireg`'s `baseline_model` (like our B-spline model `bmodel`), as including both would lead to redundant and collinear regressors.

If you *didn't* include drift terms in your `baseline_model` and wanted to use AFNI's polynomial drift correction, you could specify `polort` in the `afni_lm` call:

```{r afni_polort_example, eval=FALSE}
# Example: No drift in bmodel, use AFNI polort
bmodel_no_drift <- baseline_model(basis = "constant", sframe = sframe) # Only intercept
fmodel_no_drift <- fmri_model(emodel, bmodel_no_drift)

# Specify polort=2 for quadratic drift
alm_spec_polort <- afni_lm(fmodel_no_drift, dset, polort = 2)
print(alm_spec_polort)
```

However, explicitly defining drift terms in `fmrireg`'s `baseline_model` is generally recommended for consistency.

## Passing Custom `3dDeconvolve` Options

You can pass additional options directly to the `3dDeconvolve` command using the `options` argument in `afni_lm`. This accepts a list where element names correspond to `3dDeconvolve` flags (without the leading `-`) and values are the arguments for those flags.

```{r afni_options_example}
# Example: Request statistic and coefficient buckets, disable t-stat output
alm_spec_opts <- afni_lm(fmodel, dset, 
                         options = list(bucket = "stats_afni", 
                                        cbucket = "coefs_afni",
                                        tout = FALSE))
print(alm_spec_opts)
```

## Using AFNI-Native HRF Basis Functions

A powerful feature is the ability to specify basis functions within `event_model` that directly map to AFNI's built-in HRF models (like `BLOCK`, `TENT`, `SPMG1`, etc.). This avoids pre-convolution in R and lets `3dDeconvolve` handle the HRF modeling using its optimized routines.

Use the `afni_hrf()` specifier within the `event_model` formula instead of `hrf()`.

```{r afni_native_hrf, eval=afni_available}
# Example using AFNI's BLOCK function for convolution
# Note: Using afni_hrf requires AFNI to be installed for 3dDeconvolve to run
emodel_afni_block <- event_model(
  onset ~ afni_hrf(stim, basis = "BLOCK(2,1)"), # BLOCK(duration, penalty)
  data = simple_design, 
  block = ~ run, 
  sampling_frame = sframe
)

fmodel_afni <- fmri_model(emodel_afni_block, bmodel)
alm_spec_afni_native <- afni_lm(fmodel_afni, dset)
print(alm_spec_afni_native)

# Check the command: It should now use -stim_times instead of -stim_file
# and specify 'BLOCK(2,1)' as the basis function.
```

```{r afni_native_hrf_msg, eval=!afni_available, echo=FALSE}
cat("Skipping AFNI-native HRF example as AFNI availability check is FALSE.")
```

Refer to the `afni_hrf()` documentation and the `3dDeconvolve` help for details on available AFNI basis functions and their parameters.

## Handling Nuisance Regressors and Censoring

*   **Nuisance Regressors:** If your `baseline_model` includes nuisance regressors loaded from files (e.g., motion parameters), `afni_lm` will include them via the `-ortvec` argument, similar to how drift terms are handled.
*   **Censoring:** You can provide a censoring vector (or list of vectors per run) to the `censor` argument of `afni_lm`. This will generate a `censor.1D` file and add the `-censor censor.1D` option to the `3dDeconvolve` command.

```{r afni_censor_example}
# Example: Censor the first 5 scans
censor_vec <- rep(1, 140) # Start with all scans included (1)
censor_vec[1:5] <- 0       # Censor first 5 scans (0)

alm_spec_censored <- afni_lm(fmodel, dset, censor = censor_vec)
print(alm_spec_censored)
```

## Generating Files and Running the Model

Once you have the `afni_lm_spec` object, you can generate the necessary files and optionally execute the `3dDeconvolve` command using the `run()` method.

```{r afni_run, eval=FALSE}
# This command will:
# 1. Create the output directory 'glm_afni_output'.
# 2. Change into that directory.
# 3. Write all necessary files:
#    - Stimulus files (.1D) for each regressor if using standard hrf()
#    - Stimulus timing files (.1D) if using afni_hrf()
#    - Nuisance regressor files (.1D)
#    - GLT files (.txt)
#    - Censor file (censor.1D) if specified
#    - The 3ddeconvolve.sh script containing the full command
# 4. Execute the 3ddeconvolve.sh script.
# 5. Change back to the original directory.

run(alm_spec_opts, outdir = "glm_afni_output")

# To only generate the script and files without running:
# run(alm_spec_opts, outdir = "glm_afni_output", execute = FALSE)
```

This workflow allows you to define complex models in R using `fmrireg` and then leverage AFNI's `3dDeconvolve` for efficient estimation.



