---
title: "Building fMRI Regressors"
author: "Bradley R. Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Regressors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(fmrireg)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Introduction: What is a Regressor?

In fMRI analysis, a **regressor** (or predictor) represents the expected BOLD signal timecourse associated with a specific experimental condition or event type. It's typically created by convolving a series of event onsets (often represented as delta functions or "sticks") with a hemodynamic response function (HRF).

`fmrireg` provides the `regressor()` function to easily create these objects from event timings and an HRF. While these regressor objects are often constructed automatically behind the scenes by modeling functions like `fmri_lm`, this vignette explores how to create and manipulate them directly, offering finer control over the model components.

## Basic Regressor from Event Onsets

Suppose we have a simple event-related fMRI design with stimuli presented every 12 seconds. We want to model these events using the SPM canonical HRF (`HRF_SPMG1`). The events are brief, so we model them with a duration of 0 seconds (instantaneous).

```{r basic_regressor}
# Define event onsets
onsets <- seq(0, 10 * 12, by = 12)

# Create the regressor object
# Uses HRF_SPMG1 by default if no hrf is specified
# Duration is 0 by default
reg1 <- regressor(onsets = onsets, hrf = HRF_SPMG1)

# Print the regressor object to see its properties (uses new print.Reg method)
print(reg1)

# Access components using helper functions
head(onsets(reg1))
nbasis(reg1)
```

## Evaluating and Plotting a Regressor

A `regressor` object stores the event information but doesn't automatically compute the timecourse. To get the predicted BOLD signal at specific time points (e.g., corresponding to scan acquisition times), we use the `evaluate()` function.

```{r evaluate_plot_basic}
# Define a time grid corresponding to scan times (e.g., TR=2s)
TR <- 2
scan_times <- seq(0, 140, by = TR)

# Evaluate the regressor at scan times
predicted_bold <- evaluate(reg1, scan_times)

# Plot the predicted timecourse using ggplot2
plot_df <- data.frame(Time = scan_times, Response = predicted_bold)
ggplot(plot_df, aes(x = Time, y = Response)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg1), linetype = "dashed", color = "red", alpha = 0.7) +
  labs(title = "Predicted BOLD Response (SPM HRF)",
       x = "Time (seconds)",
       y = "Predicted Response") +
  theme_minimal()

# The autoplot() method provides a quick ggplot object
# library(ggplot2)
# autoplot(reg1, grid = scan_times)
```

The `autoplot()` method (requires `ggplot2`) provides a quick way to visualize the evaluated timecourse.

## Varying Event Durations

Sometimes events have different durations. The `duration` argument in `regressor()` can take a vector matching the length of `onsets`.

```{r varying_duration}
# Example onsets and durations
onsets_var_dur <- seq(0, 5 * 12, length.out = 6)
durations_var <- 1:length(onsets_var_dur) # Durations increase from 1s to 6s

# Create regressor with varying durations
reg_var_dur <- regressor(onsets_var_dur, HRF_SPMG1, duration = durations_var)
print(reg_var_dur)

# Evaluate and plot
scan_times_dur <- seq(0, max(onsets_var_dur) + 30, by = TR)
pred_var_dur <- evaluate(reg_var_dur, scan_times_dur)

plot_df_dur <- data.frame(Time = scan_times_dur, Response = pred_var_dur)
ggplot(plot_df_dur, aes(x = Time, y = Response)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg_var_dur), linetype = "dashed", color = "red", alpha = 0.7) +
  labs(title = "Regressor with Varying Event Durations",
       subtitle = "Duration increases over time",
       x = "Time (seconds)",
       y = "Predicted Response") +
  theme_minimal()
```

### Duration and Summation

By default (`summate=TRUE`), the predicted response accumulates if events overlap or have extended duration. Setting `summate=FALSE` models saturation, where the response width increases with duration, but the peak amplitude doesn't necessarily.

```{r duration_no_summate}
# Create regressor with varying durations, summate=FALSE
reg_var_dur_nosum <- regressor(onsets_var_dur, HRF_SPMG1, 
                               duration = durations_var, summate = FALSE)

# Evaluate and plot
pred_var_dur_nosum <- evaluate(reg_var_dur_nosum, scan_times_dur)

plot_df_dur_nosum <- data.frame(Time = scan_times_dur, Response = pred_var_dur_nosum)
ggplot(plot_df_dur_nosum, aes(x = Time, y = Response)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg_var_dur_nosum), linetype = "dashed", color = "red", alpha = 0.7) +
  labs(title = "Regressor with Varying Durations (summate=FALSE)",
       subtitle = "Models saturation, peak height may not increase with duration",
       x = "Time (seconds)",
       y = "Predicted Response") +
  theme_minimal()
```

## Varying Event Amplitudes (Parametric Modulation)

We can model variations in event intensity or some associated parameter by providing an `amplitude` vector. This creates a *parametric regressor* where the height of the HRF for each event is scaled by the corresponding amplitude value.

```{r parametric_modulation}
# Example onsets and amplitudes (e.g., representing task difficulty)
onsets_amp <- seq(0, 10 * 12, length.out = 11)
amplitudes_raw <- 1:length(onsets_amp)

# It's common practice to center the modulator
amplitudes_scaled <- scale(amplitudes_raw, center = TRUE, scale = FALSE)

# Create the parametric regressor
reg_amp <- regressor(onsets_amp, HRF_SPMG1, amplitude = amplitudes_scaled)
print(reg_amp)

# Evaluate and plot
scan_times_amp <- seq(0, max(onsets_amp) + 30, by = TR)
pred_amp <- evaluate(reg_amp, scan_times_amp)

plot_df_amp <- data.frame(Time = scan_times_amp, Response = pred_amp)
ggplot(plot_df_amp, aes(x = Time, y = Response)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg_amp), linetype = "dashed", color = "red", alpha = 0.7) +
  # Add points showing amplitude (scaled for visibility)
  geom_point(data = data.frame(Time = onsets(reg_amp), Amplitude = amplitudes(reg_amp)),
             aes(x = Time, y = Amplitude * 0.2), color = "blue", size = 2) +
  labs(title = "Parametric Regressor with Varying Amplitude",
       subtitle = "Amplitude (centered) increases over time",
       x = "Time (seconds)",
       y = "Predicted Response") +
  theme_minimal()
```

## Combining Duration and Amplitude Modulation

You can provide both `duration` and `amplitude` vectors to model events that vary in both aspects.

```{r duration_amplitude}
set.seed(123)
onsets_comb <- seq(0, 10 * 12, length.out = 11)
amps_comb <- scale(1:length(onsets_comb), center = TRUE, scale = FALSE)
durs_comb <- sample(1:5, length(onsets_comb), replace = TRUE)

reg_comb <- regressor(onsets_comb, HRF_SPMG1, 
                      amplitude = amps_comb, duration = durs_comb)
print(reg_comb)

# Evaluate and plot
scan_times_comb <- seq(0, max(onsets_comb) + 30, by = TR)
pred_comb <- evaluate(reg_comb, scan_times_comb)

plot_df_comb <- data.frame(Time = scan_times_comb, Response = pred_comb)
ggplot(plot_df_comb, aes(x = Time, y = Response)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg_comb), linetype = "dashed", color = "red", alpha = 0.7) +
  labs(title = "Regressor with Varying Duration and Amplitude",
       x = "Time (seconds)",
       y = "Predicted Response") +
  theme_minimal()
```

## Regressors with HRF Basis Sets

If you use an HRF object with multiple basis functions (e.g., `HRF_SPMG3`, `HRF_BSPLINE`), the `regressor` object will represent multiple timecourses, one for each basis function. `evaluate()` will return a matrix.

```{r basis_set_regressor}
# Use a B-spline basis set
onsets_basis <- seq(0, 10 * 12, length.out = 11)
hrf_basis <- HRF_BSPLINE # Uses N=5 basis functions by default

reg_basis <- regressor(onsets_basis, hrf_basis)
print(reg_basis)
nbasis(reg_basis) # Should be 5

# Evaluate - this returns a matrix
scan_times_basis <- seq(0, max(onsets_basis) + 30, by = TR)
pred_basis_matrix <- evaluate(reg_basis, scan_times_basis)
dim(pred_basis_matrix) # rows = time points, cols = basis functions

# Plot the basis set regressors using autoplot
# autoplot(reg_basis, grid = scan_times_basis)

# Or plot manually with ggplot
colnames(pred_basis_matrix) <- paste0("Basis_", 1:ncol(pred_basis_matrix))
plot_df_basis <- as.data.frame(pred_basis_matrix) %>% 
  mutate(Time = scan_times_basis) %>% 
  pivot_longer(-Time, names_to = "Basis", values_to = "Response")

ggplot(plot_df_basis, aes(x = Time, y = Response, color = Basis)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg_basis), linetype = "dashed", color = "red", alpha = 0.3) +
  labs(title = "Regressor using B-Spline Basis Set (N=5)",
       x = "Time (seconds)",
       y = "Predicted Response",
       color = "Basis Function") +
  theme_minimal()
```

## Shifting Regressors

You can temporally shift all onsets within a regressor using the `shift()` method.

```{r shift_regressor}
# Original regressor
reg_orig <- regressor(onsets = c(10, 30, 50), hrf = HRF_SPMG1)

# Shifted regressor (delay by 5 seconds)
reg_shifted <- shift(reg_orig, shift_amount = 5)

onsets(reg_orig)
onsets(reg_shifted) # Onsets are now 15, 35, 55

# Plot both
scan_times_shift <- seq(0, 80, by = TR)
pred_orig <- evaluate(reg_orig, scan_times_shift)
pred_shifted <- evaluate(reg_shifted, scan_times_shift)

plot_df_shift <- data.frame(
  Time = scan_times_shift,
  Original = pred_orig,
  Shifted_p5s = pred_shifted
) %>% 
  pivot_longer(-Time, names_to = "Regressor", values_to = "Response")

ggplot(plot_df_shift, aes(x = Time, y = Response, color = Regressor)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = onsets(reg_orig), linetype = "dashed", color = "red", alpha = 0.5) +
  geom_vline(xintercept = onsets(reg_shifted), linetype = "dashed", color = "blue", alpha = 0.5) +
  labs(title = "Shifting a Regressor",
       subtitle = "Original (red lines) vs. Shifted +5s (blue lines)",
       x = "Time (seconds)",
       y = "Predicted Response") +
  theme_minimal()
```


